% !TeX root = main.tex

\section{Figures}

Figures are classified according to their role within the manuscript.
Each figure type is intended to support a specific section of the paper
and should be placed where it contributes most directly to the reader's
understanding.

\begin{table}[h!]
\centering
\caption{Figure classification and recommended placement.}
\label{tab:figure_classification}
\begin{tabular}{p{3.0cm} p{6.5cm} p{3.0cm}}
\hline
\textbf{Figure Type} & \textbf{Purpose} & \textbf{Typical Section} \\
\hline

Architecture Figure &
Describes system organization, workflow, or data flow. &
Methodology \\

Benchmark-Dataset Figure & Illustrates benchmark configurations, datasets, particle distributions,
or test environments used during evaluation. & Experimental Setup \\

Performance Figure &
Presents measured results such as timing, throughput, scaling,
efficiency, or resource utilization. &
Results \\

Interpretation Figure & Provides explanation of observed behavior, comparison of trends, or
discussion of performance characteristics. & Discussion \\

\hline
\end{tabular}
\end{table}

The purpose of this classification is to maintain a clear distinction
between describing the framework, defining the evaluation, presenting
measurements, and interpreting results. Figures should generally appear
in the section corresponding to their primary role within the paper.

Examples within this work include architectural overview diagrams in
the Methodology section, benchmark-density visualizations in
Experimental Setup, throughput and timing plots in Results, and
saturation or scaling interpretations in Discussion.

\begin{table}[h!]
\centering
\caption{Current figure inventory and recommended placement within the manuscript.}
\label{tab:figure_inventory}
\begin{tabular}{p{4.0cm} p{3.5cm} p{3.5cm}}
\hline
\textbf{Figure} & \textbf{Type} & \textbf{Section} \\
\hline

OverviewArchitecture &
Architecture Figure &
Methodology \\

Corner Occupancy / Duplicate Collision Figure &
Methodology Figure &
Methodology \\

Cell Flattening Figure (if added) &
Methodology Figure &
Methodology \\

Benchmark Particle Density Figure &
Benchmark/Dataset Figure &
Experimental Setup \\

Throughput Plots &
Performance Figure &
Results \\

Time-per-Particle Plots &
Performance Figure &
Results \\

Scaling Fits &
Performance Figure &
Results \\

Saturation Explanation Figure (if added) &
Interpretation Figure &
Discussion \\

\hline
\end{tabular}
\end{table}

\section{Comparative Analysis}

The present work explicitly accounts for occupancy generation,
occupancy-list traversal, narrow-phase evaluation, and storage
requirements. This level of decomposition is intended to provide a
clear correspondence between the implemented framework and the reported
complexity results.

\section{Complexity}

The computational complexity of gPCD is determined primarily by
broad-phase occupancy generation and narrow-phase collision evaluation.
Because occupancy construction is performed independently for each
particle and each particle contributes a bounded number of occupancy
evaluations, the broad phase exhibits linear scaling with respect to
particle count.


Let $S$ denote the side length of the cubic cell array and let $W$
denote the fixed occupancy-list width assigned to each cell. The cell
array therefore contains $S^3$ cells, each capable of storing up to $W$
particle-corner occupancy entries.

The cell-array storage requirement is

\begin{equation}
M_{\mathrm{cell}} = O(S^3 W),
\end{equation}

where $W$ is the fixed occupancy-list width per cell. Since $W$ is a
constant implementation parameter, the storage requirement reduces to

\begin{equation}
M_{\mathrm{cell}} = O(S^3).
\end{equation}

\input{ComplexityTable}




\subsection{Occupancy Generation}

Broad-phase occupancy generation is performed once per particle. Each
particle contributes a fixed number of corner evaluations, corresponding
to the eight corners of its axis-aligned bounding box. Since the number
of corner evaluations does not depend on the total number of particles,
the occupancy-generation cost per particle is constant.

Let $c_{\mathrm{occ}}$ denote the bounded cost associated with corner
evaluation, cell determination, index computation, and occupancy-list
insertion for a single particle. The total occupancy-generation cost is

\begin{equation}
T_{\mathrm{occ}}(N) = c_{\mathrm{occ}} N .
\end{equation}

Therefore, broad-phase occupancy generation scales linearly with the
number of particles,

\begin{equation}
T_{\mathrm{occ}}(N) = O(N).
\end{equation}

The populated cell array produced by this process constitutes the
potentially colliding set used during narrow-phase evaluation. Thus, no
separate candidate-construction stage is required between occupancy
generation and the narrow phase.

\subsection{Narrow-Phase Evaluation}

The potentially colliding set produced during broad-phase occupancy
generation is evaluated within a compute shader during narrow-phase
collision detection. Each occupied cell is processed independently and
particle interactions are evaluated only among particles contained
within the corresponding occupancy list.

Each occupancy list contains at most $W$ entries, where $W$ is a fixed
implementation parameter. Traversal of an occupancy list therefore
requires $O(W)$ operations per occupied cell. Since $W$ is a fixed
implementation parameter, occupancy-list traversal contributes a
constant cost to each cell evaluation.

Let $K$ denote the total number of interaction evaluations performed
during narrow-phase execution. Since each interaction evaluation
requires bounded work, the total narrow-phase cost is

\begin{equation}
T_{\mathrm{np}}(K)=O(K).
\end{equation}

The value of $K$ depends on particle distribution and occupancy within
the populated cell array. Consequently, narrow-phase complexity is
determined by the number of particle interactions represented within
the potentially colliding set rather than by the total number of
particles alone.


\subsection{Storage Requirements}

The storage requirements of gPCD consist primarily of particle storage
and cell-array storage. Particle storage scales linearly with the 
number of particles,

\begin{equation}
M_{\mathrm{particle}} = O(N).
\end{equation}

Let $N$ denote the number of particles, $S$
denote the side length of the cubic cell array, and $W$ denote the
fixed occupancy-list width assigned to each cell.

The cell array contains

\begin{equation}
S^3
\end{equation}

cells, each capable of storing up to $W$ particle-corner occupancy entries. The total
cell-array capacity is therefore

\begin{equation}
M_{\mathrm{cell}} = S^3W.
\end{equation}

The corresponding storage complexity is

\begin{equation}
M_{\mathrm{cell}} = O(S^3W).
\end{equation}

Since $W$ is a fixed implementation parameter,

\begin{equation}
M_{\mathrm{cell}} = O(S^3).
\end{equation}

The total storage requirement of the framework is the sum of particle
storage and cell-array storage,

\begin{equation}
M_{\mathrm{total}}
=
M_{\mathrm{particle}}
+
M_{\mathrm{cell}}.
\end{equation}

The resulting storage complexity is

\begin{equation}
M_{\mathrm{total}}
=
O(N + S^3W).
\end{equation}

For fixed occupancy-list width,

\begin{equation}
M_{\mathrm{total}}
=
O(N + S^3).
\end{equation}

Although the cell-array storage requirement scales with the total
number of cells, occupancy generation does not require traversal of the
entire cell array. Each particle computes a flattened cell index and
directly accesses the corresponding occupancy list. Consequently,
cell-array storage complexity and occupancy-generation complexity are
independent quantities.


\subsection{Overall Complexity}

The computational complexity of gPCD is determined by the combined
costs of broad-phase occupancy generation and narrow-phase collision
evaluation. From the preceding analysis,

\begin{equation}
T_{\mathrm{occ}}(N) = O(N)
\end{equation}

and

\begin{equation}
T_{\mathrm{np}}(K) = O(K),
\end{equation}

where $N$ is the number of particles and $K$ is the number of particle
interactions evaluated during narrow-phase execution.

The resulting computational complexity of the framework is therefore

\begin{equation}
T_{\mathrm{gPCD}}=O(N + K).
\end{equation}

Similarly, the total storage requirement is the sum of particle storage
and cell-array storage,

\begin{equation}
M_{\mathrm{gPCD}}=O(N + S^3W),
\end{equation}

where $S$ is the side length of the cubic cell array and $W$ is the
fixed occupancy-list width assigned to each cell. For fixed occupancy
width, the storage complexity reduces to

\begin{equation}
M_{\mathrm{gPCD}}=O(N + S^3).
\end{equation}

Consequently, broad-phase occupancy generation scales linearly with
particle count, while narrow-phase cost depends on the number of
particle interactions represented within the potentially colliding set.

The storage requirement of the cell array scales with the total number
of cells, $O(S^3W)$. However, occupancy generation performs direct
cell access through flattened indexing and therefore scales with
particle count rather than cell-array size.