% !TeX root = main.tex
\section{Conclusion and Future Work}

This work presented the graphics-based Parallel Collision Detection
(gPCD) framework, a collision-detection architecture that integrates
broad-phase occupancy generation directly into the graphics pipeline.
Unlike traditional approaches that treat rendering as a consumer of
simulation state, the proposed framework allows occupancy construction
and rendering to proceed concurrently while producing the potentially
colliding set used during narrow-phase evaluation.

A detailed complexity analysis was presented in which occupancy
generation, candidate construction, narrow-phase evaluation, and
storage requirements were examined independently. Experimental results
demonstrated high particle throughput and near-linear scaling across
graphics, compute, and total execution times. Log--log regression and
curvature analyses further showed that the measured scaling behavior is
consistent with the complexity analysis. Additional locality studies
demonstrated that memory organization affects performance through
implementation efficiency while preserving the underlying asymptotic
behavior of the framework.

The results suggest that graphics-pipeline occupancy generation
provides a viable alternative execution architecture for large-scale
particle collision detection. The demonstrated particle counts indicate
that a single CPU--GPU system is capable of supporting particle-system
scales traditionally associated with high-performance computing
environments.

Future work will focus on extending the framework to larger GPU
configurations, multi-GPU execution, and distributed computing
environments. While the present work focuses on collision-detection
performance, the framework has already been applied within
particle-based thermo-fluid simulations, where preliminary studies have
produced promising results and novel physical observations. Future work
will investigate these applications in greater depth and evaluate the
role of graphics-pipeline occupancy generation in large-scale
thermo-fluid and multiphysics simulations.