% !TeX root = main.tex
\section{Experimental Results}

This section presents the measured performance of the gPCD framework
using randomly distributed particle configurations of increasing size.
Unless otherwise noted, benchmark cases maintain an approximately
constant collision fraction while varying the total particle count.
Performance is evaluated using throughput, time-per-particle, and
scaling analyses to characterize the behavior of the framework across
the tested particle range.

\subsection{Memory Locality Benchmark}

This benchmark evaluates the sensitivity of the gPCD framework to
memory locality. Two occupancy configurations were considered. In the
first configuration, cell occupancy data remained sequentially ordered
within memory. In the second configuration, occupancy data were
randomly distributed throughout the cell array.

Figure~\ref{fig:A_PQBS_V_PQBR} compares the resulting execution times as a
function of particle count. Sequential ordering consistently produced
lower execution times throughout the evaluated benchmark range. The
largest benchmark configuration required approximately
\nc{Sequential Largest Time} ms for sequential ordering and
\nc{Random Largest Time} ms for random ordering.

Although both configurations exhibit similar scaling behavior, random
ordering introduces a measurable performance penalty. The results
indicate that memory locality remains an important contributor to
overall performance despite the GPU-resident nature of the framework.
\input{KEEP_PQBRT_TABLE_ALL}

\subsection{Performance Overview}

\Tigo{performance_summary} summarizes the principal
performance metrics obtained during evaluation of the gPCD framework.
The benchmark suite consisted of randomly distributed particle
configurations of increasing size while maintaining an approximately
constant collision fraction.

The framework successfully processed benchmark configurations
containing up to \numprint{\APQBRTTABLEALLbka{}} particles and achieved a
peak throughput of \SUMMARYTHROUGHPUT{} million particles/s. 
Representative execution statistics for the largest 
benchmark case are also reported in Table~\ref{tab:performance_summary}.

\input{A_PQBRT_SUMMARY_TABLE}

Across the evaluated range, performance remained stable and
exhibited the expected throughput-oriented behavior of a
GPU-resident collision-detection framework. More detailed
analysis of throughput trends, time-per-particle behavior,
and scaling characteristics is presented in the following
subsections.

The benchmark measurements used throughout the subsequent
performance analyses are provided in
Table~\ref{tab:A_PQBRT_TABLE_ALL}.


\input{A_PQBR_THROUGHPUT_AND_TPP}

The following subsections examine throughput behavior,
time-per-particle trends, scaling characteristics, and
collision-fraction sensitivity in greater detail.

\subsection{Throughput and Time-per-Particle Behavior}

Throughput and time-per-particle provide complementary views of
framework performance. Throughput measures the aggregate rate at which
particles are processed, whereas time-per-particle normalizes execution
cost by particle count and therefore provides a direct measure of
scaling efficiency.

Figure~\ref{fig:A_PQBR_THROUGHPUT_AND_TPP} presents complementary
throughput and time-per-particle views of framework performance across
the evaluated particle range.

\subsubsection{Utilization-Limited Regime}

At low particle counts, throughput increases rapidly with increasing
particle count as available GPU resources become more fully utilized.
This regime extends to approximately
\numprint{100000} particles.

\subsubsection{Throughput-Transition Regime}

Beyond \numprint{100000} particles, throughput enters a transition
regime in which increasing occupancy and collision workload begin to
influence performance. This regime spans particle counts between
approximately \numprint{100000} and \numprint{2000000}.

\subsubsection{Stable Large-Scale Regime}

At approximately \numprint{2000000} particles, the framework reaches a
stable large-scale operating regime characterized by nearly constant
time-per-particle behavior. This region is used for the subsequent
time-per-particle analysis and represents the asymptotic operating
behavior of the framework.

\subsubsection{Stable Large-Scale Regime}

At approximately \numprint{2000000} particles, the framework reaches a
stable large-scale operating regime. Although throughput continues to
decrease gradually with increasing particle count, the corresponding
time-per-particle curves become nearly constant. This behavior indicates
that the framework has entered an asymptotic operating region in which
the average processing cost per particle remains approximately
constant. Consequently, particle counts exceeding
\numprint{2000000} are used for the time-per-particle analysis reported
in Figure~\ref{fig:A_PQBR_THROUGHPUT_AND_TPP}(c).

\subsection{Scaling Analysis}
\input{A_PQBR_LOGLOG_REGRESS}
Log--log regression was performed to quantify the scaling behavior of
graphics, compute, and total execution times. A power-law model of the
form
\begin{equation}
t = aN^b
\end{equation}
was fitted to the measured data, where $t$ is execution time, $N$ is
particle count, $a$ is a proportionality constant, and $b$ is the
scaling exponent.

Figure~\ref{fig:A_PQBR_LOGLOG_REGRESS} presents the resulting log--log
regressions for graphics, compute, and total execution times. The
measured scaling exponents were found to be
$b_g=\APQBRLOGLOGREGRESSgmsasat{}$,
$b_c=\APQBRLOGLOGREGRESScmsasat{}$, and
$b_t=\APQBRLOGLOGREGRESStotasat{}$.
All three exponents remain close to unity, indicating that execution
time increases approximately linearly with particle count throughout
the evaluated benchmark range.

The measured scaling exponents remain close to unity for graphics,
compute, and total execution times, providing experimental support for
the near-linear scaling behavior predicted by the complexity analysis.

\subsection{Curvature Analysis}
\input{A_PBQR_CURVATURE}
While log-log regression quantifies the overall scaling exponent,
curvature analysis provides a complementary measure of deviation from
ideal linear behavior. Quadratic fits were therefore applied to the
graphics, compute, and total execution-time measurements over the
evaluated benchmark range.

Figure~\ref{fig:A_PBQR_CURVATURE} presents the resulting quadratic fits and
associated curvature bounds. In all cases, the fitted curves remain
close to the measured data, indicating only minor departures from
linear behavior. The graphics and total execution times exhibit
particularly small curvature, while the compute measurements display
slightly larger but still bounded deviations.

The measured curvature bounds remain small relative to the overall
execution-time trends. Combined with the log-log scaling exponents,
these results provide additional evidence that the gPCD framework
maintains near-linear scaling throughout the evaluated benchmark range.

