% !TeX root = main.tex
\section{Introduction}

Particle-based methods provide a natural framework for modeling systems
composed of discrete interacting entities. Such methods are widely used
to study solids, fluids, granular materials, multiphase flows, molecular
systems, and a broad range of engineering and scientific applications 
\cite{Meakin2009,onate2011,allen2017}
Unlike continuum approaches, particle methods directly represent
individual interacting elements, allowing macroscopic behavior to emerge
from local interactions. This capability makes particle simulations
particularly valuable for studying systems in which complex collective
behavior arises from large numbers of discrete constituents.

Beyond their traditional role in engineering analysis, particle
simulations also serve as tools for scientific discovery. The emergence
of previously unobserved physical behavior often requires the direct
simulation of large numbers of interacting particles. While modern
machine-learning methods can effectively learn from existing datasets,
they remain fundamentally dependent upon representative training data.
Consequently, high-fidelity particle simulations continue to play a
critical role in generating new physical insights, producing training
datasets, and investigating phenomena that have not yet been fully
characterized through experiment or analysis.

As simulation fidelity increases, the number of particles required to
accurately represent physical systems may grow into the millions or
billions. Higher particle counts enable finer spatial resolution,
improved representation of complex geometries, and the investigation of
phenomena spanning a broad range of scales, from dilute particulate
systems to dense continuum-like materials. However, these benefits come
at the cost of substantially increased computational requirements,
making efficient particle-processing algorithms a critical component of
large-scale particle simulation.

One of the primary limitations of large-scale particle simulation is the
cost of collision detection. A naïve all-to-all collision search scales
as $(O(N^2))$, making direct evaluation impractical for large particle
counts. Broad-phase collision-detection algorithms reduce this cost
\cite{ericson2004} by restricting collision testing to nearby
particles; however, efficient broad-phase construction remains a
significant challenge for massively parallel implementations.

Modern graphics processing units (GPUs) provide substantial
computational throughput and memory bandwidth, making them attractive
platforms for particle simulation. Despite these advances, generating and
maintaining spatial occupancy structures efficiently at large scales
remains a fundamental bottleneck.

This paper presents GPU-driven high-performance graphics particle collision
detection (gPCD), a collision-detection framework that leverages the
graphics pipeline to generate particle occupancy information and
construct a uniform cell-array representation suitable for massively
parallel collision processing. By organizing occupancy generation around
particle corner locations and exploiting GPU rasterization hardware,
gPCD enables efficient broad-phase construction while maintaining a
simple and scalable data structure. The resulting framework supports
high particle counts on a single CPU--GPU system and provides a
foundation for future large-scale particle simulation applications.



