\documentclass[11pt]{article}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, positioning}
\usepackage{amsmath, amssymb}
\usepackage{graphicx}
\usepackage{geometry}
\geometry{margin=1in}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage{listings}
% Subcaption for subplot
\usepackage{subcaption}
\usepackage{numprint}
\npthousandsep{,}
%
\usepackage{makecell}
% midrule,toprule,...
\usepackage{booktabs}
\usepackage{xfrac}
% need this for references in *.doi
\usepackage{hyperref}
% Numeric citations
\usepackage[numbers,sort&compress]{natbib}
\newcommand{\nc}[1]{!#1!}
\newcommand{\mmrr}{\textit{\textbf{$mmrr$}}}
\newcommand{\arrt}{\textit{\textbf{$arr_T$}}}
\newcommand\Description[2][]{}
\newcommand{\bell}[2]{\textcolor{blue}{BELL:#1:[#2]}}
\newcommand{\figo}[1]{fig.~\ref{fig:#1}}
\newcommand{\Figo}[1]{Fig.~\ref{fig:#1}}
\newcommand{\figob}[2]{fig.~\ref{fig:#1}-(#2)}
\newcommand{\Tigo}[1]{Table~\ref{tab:#1}}
\newcommand{\CN}{\textcolor{red}{[CitationNeeded]}}
\captionsetup{font=footnotesize, labelfont=bf}
\captionsetup[sub]{font=footnotesize, labelfont=bf}


% Keywords command
\providecommand{\keywords}[1]
{
  \small	
  \textbf{\textit{Keywords---}} #1
}
\input{LatexValsDataBase.tex}

\title{A GPU Driven Graphical Particle Collision Detection Method- gPCD}
\author{}
\date{}

\begin{document}
\begin{abstract}

Broad-phase collision detection is a fundamental component of
particle-based simulation systems and is often a dominant contributor
to overall computational cost. Existing GPU-resident approaches
typically perform occupancy generation and candidate construction as
dedicated collision-detection operations, while rendering remains a
consumer of simulation state. This separation limits the ability of the
graphics pipeline to contribute directly to collision detection.

This work presents the graphics-based Parallel Collision Detection
(gPCD) framework, a GPU-resident collision-detection architecture that
integrates broad-phase occupancy generation directly into the graphics
pipeline. The proposed framework allows occupancy construction and
rendering to proceed concurrently while generating the potentially
colliding set used during narrow-phase evaluation. A detailed
complexity analysis is presented in which occupancy generation,
candidate construction, narrow-phase evaluation, and storage
requirements are examined independently.

Experimental evaluation demonstrates high particle throughput and
near-linear scaling across graphics, compute, and total execution
times. Log--log regression produced scaling exponents close to unity,
while curvature analysis indicated only minor departures from linear
behavior over the evaluated benchmark range. Additional locality
studies showed that memory organization affects implementation
efficiency while preserving the underlying asymptotic behavior of the
framework.

The results demonstrate that graphics-pipeline occupancy generation
provides a viable execution architecture for large-scale particle
collision detection and enables particle-system scales approaching those commonly associated
with high-performance computing environments to be achieved on a single
CPU--GPU system.

\end{abstract}

\keywords{
collision detection,
GPU computing,
parallel algorithms,
particle simulation,
graphics pipeline,
real-time simulation
}

\input{Introduction}
\input{RelatedWork}
\input{Methodology}
\input{Complexity}
\input{ExperimentalSetup}
\input{ExperimentalResults}
\input{Discussion}
\input{Conclusions}
\bibliographystyle{ACM-Reference-Format}
\bibliography{ReferencesWithdoi.bib}
\input{Conventions}
\end{document}
