% !TeX root = main.tex
\section{Methodology}

The gPCD framework addresses the architectural gap identified in the
previous section by integrating broad-phase occupancy generation
directly into the graphics pipeline. The resulting framework allows
occupancy construction and rendering to proceed concurrently while
producing the potentially colliding set used during narrow-phase
evaluation.

\subsection{Framework Overview}


Figure~\ref{fig:overview_architecture}(a) illustrates the conventional
organization in which rendering and collision detection operate as
separate processes in many existing
GPU-based collision-detection frameworks. Figure~\ref{fig:overview_architecture}(b) shows the
gPCD framework, where broad-phase occupancy generation is integrated
directly into the graphics pipeline, allowing occupancy construction and
rendering to proceed concurrently.

\input{CellSpans}

\subsection{Duplicate Handling}

Two forms of duplication can arise within the gPCD framework.
Corner duplication occurs when multiple AABB corners from the same
particle map to a common cell location during occupancy generation.
Collision duplication occurs when the same source--target interaction is
encountered through multiple occupied cells during PCS traversal.
Figure~\ref{fig:CellSpans} illustrates both effects.

Corner duplicates are removed during occupancy construction to prevent
multiple insertions of the same particle identifier into a single cell
occupancy list. Collision duplicates are removed during narrow-phase
evaluation using source-owned collision tracking, ensuring that each
source--target interaction is evaluated at most once.


\subsection{Graphics-Pipeline Occupancy Generation}

Broad-phase occupancy generation is implemented within the vertex stage
of the graphics pipeline. Each particle independently evaluates its
occupancy contribution and determines the cell locations required for
broad-phase construction. The resulting occupancy information is stored
within the cell-array structure and subsequently used to construct
candidate collision pairs for narrow-phase evaluation.

Broad-phase occupancy generation is implemented within the vertex stage
of the graphics pipeline. Each particle independently evaluates its
occupancy contribution and determines the cell locations required for
broad-phase construction. The resulting occupancy information is stored
within the cell-array structure and subsequently used to construct
candidate collision pairs for narrow-phase evaluation.

\subsection{Corner-Based Occupancy Construction}

Each particle is represented by an axis-aligned bounding box (AABB).
Occupancy generation is performed by evaluating the eight corners of the
AABB and determining the cell locations intersected by the particle.
The resulting corner locations define the occupancy contribution of the
particle within the cell array.

Let $(x,y,z)$ denote the floating-point particle position and let
$r$ denote the particle radius. The eight corner locations are
constructed as

\begin{equation}
\mathbf{x}_c =\mathbf{x}
+
(\pm r,\pm r,\pm r),
\end{equation}

where $c \in \{0,\ldots,7\}$ identifies one of the eight AABB corners.

Each corner location is subsequently converted to a cell coordinate and
used to determine particle occupancy within the cell array.

Because occupancy is determined from a fixed set of corner locations,
each particle performs a bounded number of occupancy evaluations
independent of the total particle count. Consequently, occupancy
generation requires constant work per particle.

The particle diameter is constrained not to exceed the cell width.
Under this constraint, a particle may intersect at most eight
neighboring cells, allowing occupancy to be determined entirely from
the eight AABB corners. As a result, particle size may be scaled over
the admissible range without altering the occupancy-generation
procedure or increasing the number of occupancy evaluations required
per particle.

\subsection{Cell Addressing and Flattening}

Each corner location is mapped to a cell coordinate within the uniform
cell array. Let $(x,y,z)$ denote a floating-point corner location and
let $(X,Y,Z)$ denote the corresponding unsigned integer cell
coordinates. Cell coordinates are obtained by rounding the corner
location to the nearest integer cell center,

\begin{equation}
X = \mathrm{round}(x),
\qquad
Y = \mathrm{round}(y),
\qquad
Z = \mathrm{round}(z).
\end{equation}

Cell coordinates are represented using zero-based indexing.
However, cells containing a zero-valued coordinate component are
reserved and are not used for occupancy generation. Consequently,
valid occupancy cells satisfy

\begin{equation}
X>0,
\qquad
Y>0,
\qquad
Z>0.
\end{equation}

This convention creates a guard layer surrounding the active domain and
prevents AABB-corner evaluations associated with particles located near
domain boundaries from producing negative cell coordinates during
occupancy construction.

The resulting cell coordinate triplet is converted into a flattened cell
index for storage within the occupancy structure. Let \(W\), \(H\), and
\(D\) denote the width, height, and depth of the cell array,
respectively. The flattened index is

\begin{equation}
I = X + W(Y + HZ).
\end{equation}

The flattened index serves as the primary cell identifier throughout
the gPCD collision-detection pipeline. In addition to directly
addressing the corresponding occupancy list within the cell array, the
flattened representation reduces cell-reference storage from a
three-component coordinate tuple to a single integer identifier. Once
computed, the index is used throughout occupancy construction,
candidate generation, and narrow-phase evaluation without requiring
reconstruction of the original cell coordinates.

Concurrent insertion into shared cell occupancy lists requires an
atomic reservation operation to obtain a unique slot within the target
cell occupancy list. This operation provides thread-safe insertion
during occupancy generation.

\subsection{Cell Array Construction}

During insertion, multiple AABB corners from the same particle may map
to the same cell. These duplicate corner references are suppressed
during broad-phase occupancy generation so that only a single copy of
the particle identifier is stored within a given cell occupancy list.
Consequently, the required occupancy-list capacity is determined by the
maximum number of unique particles that may contribute to a cell rather
than the total number of corner insertions.

For heterogeneous particle systems, the maximum occupancy is governed
by the smallest particle radius present in the simulation. Let

\begin{equation}
r_{\min}=\min_i r_i
\end{equation}

denote the minimum particle radius. Let \(h\) denote the uniform cell side length.
The maximum cell occupancy is then

\begin{equation}
N_{\mathrm{cell}}^{3D}=\left(\left\lfloor\frac{h+2r_{\min}}{2r_{\min}}\right\rfloor\right)^3.
\end{equation}

Consequently, the occupancy-list capacity may be determined during
initialization directly from the smallest allowable particle diameter.
Because all larger particles occupy greater volume, they cannot increase
the maximum number of unique particle identifiers that may be stored
within a cell.


\subsection{Potentially Colliding Set Construction}


The populated cell array constitutes the potentially colliding set
(PCS) used during narrow-phase evaluation. Each occupied cell contains
the identifiers of particles whose corner locations map to that cell.
Consequently, occupancy generation directly constructs the PCS without
requiring a separate candidate-construction stage.

\subsection{Narrow-Phase Evaluation}

Each particle maintains an array of eight corner-cell identifiers,

\begin{equation}
\texttt{cornerCell}[8],
\end{equation}

where each entry stores the flattened cell index associated with one
AABB corner. These indices are generated during occupancy construction
and retained for narrow-phase evaluation.

During narrow-phase execution, each source particle independently
traverses its \texttt{cornerCell} array and accesses the corresponding
occupancy lists. Candidate neighbors are obtained directly from these
lists, eliminating the need for construction of a centralized
collision-pair list. The source particle is then compared against the
particles referenced by each occupancy list to determine contact.

To prevent duplicate collision evaluations, each particle maintains a
collision list containing the identifiers of particles with which a
collision has already been registered during the current evaluation
step. When a collision is detected, the target particle identifier is
stored in the source particle's collision list. Subsequent interaction
evaluations consult this list and discard previously evaluated
source--target pairs.

Because candidate discovery is performed independently for each source
particle, candidate ownership remains local throughout collision
evaluation. This organization simplifies massively parallel execution
and reduces synchronization requirements by avoiding construction of a
globally shared collision-pair list.
