\section{Outline Review Process}

The purpose of the outline review process is to evaluate document
organization before detailed writing and editing begin. The review
focuses on logical flow, paragraph purpose, transitions, and overall
presentation structure. Experience has shown that most revisions arise
from organizational issues rather than grammar or technical content.
Consequently, section flow should be evaluated before detailed
sentence-level editing begins.

\subsection{Review Procedure}

\begin{enumerate}
\item Review one section at a time.

\item Begin each section with a \textbf{Section Flow Overview}
identifying the major topics and their ordering.

\item Identify the questions answered by the section and the paragraph
ranges associated with each question.

\item Review one paragraph at a time.

\item For each paragraph, document:
\begin{itemize}
\item Purpose
\item Contains
\item Flow from Previous
\item Flow to Next
\item Presentation Bullets
\end{itemize}

\item Evaluate transitions before evaluating wording. A paragraph may
be technically correct and well written yet still require relocation,
consolidation, or removal if the surrounding flow is weak.

\item Eliminate unnecessary subsectioning. Subsections should only be
introduced when a genuine topic change occurs.

\item Consolidate repeated concepts whenever possible. Topics
introduced in one location should not be reintroduced unless additional
information is being provided.

\item Use figures and tables as organizational anchors. Discussion
should naturally lead into and out of supporting visual material.

\item After all paragraphs have been reviewed, reassess the Section
Flow Overview and revise the organization as necessary.
\end{enumerate}

\subsection{Section Review Template}

\subsubsection*{Section Name}

\paragraph{Section Flow Overview}

\begin{center}
Topic A\\
$\downarrow$\\
Topic B\\
$\downarrow$\\
Topic C
\end{center}

\paragraph{Questions Answered}

\begin{itemize}
\item Question 1? \\
Paragraphs X--Y

\item Question 2? \\
Paragraphs Y--Z

\item Question 3? \\
Paragraphs Z--N
\end{itemize}

\paragraph{Paragraph N}

\textbf{Purpose}
\begin{itemize}
\item Why does this paragraph exist?
\item What role does it play within the section?
\end{itemize}

\textbf{Contains}
\begin{itemize}
\item Primary technical points.
\item Supporting information.
\item Key conclusions.
\end{itemize}

\textbf{Flow from Previous}

Rating:
\begin{itemize}
\item Good
\item Fair
\item Poor
\item N/A
\end{itemize}

Reason:
\begin{itemize}
\item Explanation of transition quality.
\end{itemize}

\textbf{Flow to Next}

Rating:
\begin{itemize}
\item Good
\item Fair
\item Poor
\item N/A
\end{itemize}

Reason:
\begin{itemize}
\item Explanation of transition quality.
\end{itemize}

\textbf{Presentation Bullets}
\begin{itemize}
\item Slide-style summary point.
\item Slide-style summary point.
\item Slide-style summary point.
\end{itemize}

\subsection{Evaluation Guidelines}

\paragraph{Good Flow}

\begin{itemize}
\item Naturally continues the previous discussion.
\item Introduces information required by the following paragraph.
\item Requires little or no transition explanation.
\end{itemize}

\paragraph{Fair Flow}

\begin{itemize}
\item Topic connection exists but transition is weak.
\item Additional bridge sentences may be required.
\end{itemize}

\paragraph{Poor Flow}

\begin{itemize}
\item Abrupt topic change.
\item Information appears out of order.
\item Topic belongs elsewhere in the section.
\item Paragraph should be moved, merged, or removed.
\end{itemize}

\paragraph{Not Applicable}

\begin{itemize}
\item First paragraph of a section (Flow from Previous).
\item Final paragraph of a section (Flow to Next).
\end{itemize}

\subsection{Common Structural Problems}

\paragraph{Empty Sections}

A subsection containing only one short introductory paragraph often
indicates unnecessary structure. Consider merging the content into the
parent section.

\paragraph{Repeated Information}

If multiple paragraphs communicate the same conclusion, consolidate the
discussion into a single location.

\paragraph{Premature Detail}

Do not introduce implementation details before the concepts required to
understand them have been presented.

\paragraph{Orphaned Topics}

A paragraph that introduces a topic which is not discussed again may
belong elsewhere or may be unnecessary.

\paragraph{Broken Narrative}

Sections should generally answer questions in the following order:

\begin{center}
What?\\
$\downarrow$\\
Why?\\
$\downarrow$\\
How?\\
$\downarrow$\\
Results?\\
$\downarrow$\\
Implications?
\end{center}

When this sequence is violated, the section should be reorganized
before detailed editing begins.