% !TeX root = main.tex
\newpage
\section{Background and Related Work}

Numerous methodologies and execution architectures have been proposed
to improve collision-detection performance, particularly for large-scale
particle simulations. These approaches differ not only in the data
structures used to generate collision candidates, but also in where
collision detection is performed and how it interacts with the rendering
pipeline. This section reviews common collision-detection
methodologies, execution architectures, and rendering strategies, and
identifies the architectural gap addressed by the gPCD framework.

\subsection{Collision-Detection Frameworks}

Collision detection is commonly organized into broad-phase and
narrow-phase stages. The broad phase reduces the number of potential
collision pairs by generating candidate interactions, while the narrow
phase evaluates those candidates to determine whether a collision
exists. Although numerous collision-detection methodologies have been
proposed, most frameworks follow this two-stage organization. This
distinction is important because the primary contribution of gPCD
occurs within the broad phase, while the narrow phase completes the
collision-detection framework.

Numerous broad-phase methodologies have been developed to improve the
efficiency of candidate generation. Common approaches include uniform
grids, spatial hashing, sweep-and-prune methods, and bounding-volume
hierarchies. These techniques differ in how spatial information is
organized and queried, but they share the common objective of reducing
the number of pairwise evaluations passed to the narrow phase. As
particle counts have increased and GPU computing has become more
prevalent, research has expanded beyond broad-phase methodologies to
consider the execution architectures used to implement them.

\subsection{Collision-Detection Architectures}
Collision-detection methodologies are commonly implemented using a variety of execution architectures.
Early systems performed both broad-phase and narrow-phase operations on the CPU, where collision detection 
and rendering existed as largely independent processes. As particle counts increased and graphics hardware became 
increasingly programmable, portions of the collision-detection pipeline were migrated to the GPU to exploit massive 
parallelism. These developments produced a spectrum of architectures ranging from CPU-resident implementations to 
GPU-assisted and fully GPU-resident approaches.

In CPU-resident architectures, collision candidates are generated and evaluated entirely on the 
CPU before simulation state is transferred to the rendering system for visualization. GPU-assisted architectures 
offload selected collision-detection operations to the GPU while maintaining coordination with CPU-managed data 
structures and execution flow. GPU-resident architectures further reduce data movement by performing broad-phase 
and narrow-phase operations directly on the GPU, thereby improving parallel utilization and minimizing host-device communication.

Although these architectures differ in execution location and data movement, collision detection and rendering 
generally remain distinct processes. Collision-detection operations are typically performed using dedicated 
computational stages, while the rendering pipeline functions primarily as a consumer of simulation state. 
This distinction motivates a closer examination of the relationship between rendering and collision detection.


\subsection{Rendering and Collision Detection}

Collision detection and rendering have historically served distinct roles within simulation systems. 
Collision detection is responsible for identifying particle interactions and advancing simulation state, 
whereas rendering is responsible for visualizing the resulting geometry. Consequently, rendering pipelines 
are commonly positioned downstream of simulation and collision-detection processes, operating primarily as consumers of simulation state.

As graphics hardware evolved, both rendering and collision detection increasingly leveraged GPU parallelism. 
However, even within GPU-resident implementations, rendering and collision detection typically remained logically 
separated. Collision-detection operations were executed using dedicated computational stages, while rendering 
continued to transform simulation data into visual output. In this organization, rendering contributes little 
or no information to the collision-detection process itself.

Several approaches have explored tighter integration between rendering and simulation through image-space techniques,
GPU data structures, and graphics-assisted computation. Nevertheless, occupancy generation and candidate construction 
generally remain dedicated collision-detection operations performed independently of rasterization. As a result, 
rendering continues to function primarily as a consumer of simulation state rather than an active participant in 
broad-phase collision detection.

This separation raises an architectural question: can occupancy information be generated directly through the 
graphics pipeline itself, allowing rasterization to contribute to broad-phase collision detection while 
simultaneously supporting rendering operations?


\subsection{Gap in Existing Methods}

Broad-phase occupancy generation serves as the foundation upon which
candidate construction and subsequent narrow-phase evaluation are built.
Since every particle must participate in this process during each
simulation step, the organization of occupancy generation can have a
substantial impact on the overall collision-detection pipeline.

Existing collision-detection research has produced a wide variety of
broad-phase methodologies and execution architectures. These approaches
differ in how collision candidates are generated and where collision
detection is performed; however, occupancy generation and candidate
construction generally remain dedicated collision-detection operations.
Even within GPU-resident frameworks, rendering and collision detection
typically operate as logically separate processes.

As a consequence, the graphics pipeline is commonly treated as a
consumer of simulation state rather than a contributor to collision
detection. Particle positions generated by simulation are transferred to
the rendering pipeline for visualization, while occupancy generation and
candidate construction are performed independently using dedicated
collision-detection operations.

This separation represents an architectural limitation rather than an
algorithmic one. Existing approaches largely focus on improving the
efficiency of broad-phase methodologies or accelerating collision
detection through GPU execution. Comparatively little attention has been
given to the possibility of integrating occupancy generation directly
into the graphics pipeline itself.

The architectural gap addressed by this work is therefore the absence of
a collision-detection framework in which broad-phase occupancy
generation forms an integral part of graphics-pipeline execution. The
gPCD framework addresses this gap by integrating occupancy construction
directly into the graphics pipeline, allowing broad-phase occupancy
generation and particle rendering to proceed concurrently on the GPU.

The following section presents the methodology underlying the gPCD
framework and describes the graphics-pipeline occupancy-generation
process used to construct broad-phase collision candidates.

