% !TeX root = main.tex
\newpage
\section{Boundary-Space Lighting}

\subsection{Purpose}

Boundary-space lighting is the revised direction for \texttt{LightingBall}
surface illumination.  The goal is to light part of an analytic or implicit
surface without adding triangles, visible surface-cell geometry, or dense
boundary rendering particles.

This is the next implementation target, not part of the current cleaned
\texttt{LightingBall} baseline.  The active \texttt{LightingBall\_Py.inc} and
\texttt{LightingBall\_Vk.inc} files intentionally omit boundary-space
configuration until the minimal tracer scene is stable again.

The important separation is:
\begin{itemize}
    \item photons transport color and discover surface contact;
    \item boundary particles represent surface patches contained by spatial
          cells;
    \item the boundary particle stores the persistent color state for that
          patch;
    \item rendering spreads that stored color over nearby surface samples by
          a kernel.
\end{itemize}

This keeps the renderer particle-based while allowing a finite region of a
surface to remain lit after a photon contact.

\subsection{Boundary Particle As Surface Proxy}

A boundary particle is not an infinitesimal lit dot.  In this model it is the
representative for the portion of surface contained in or discovered through
its cell.  A photon may hit the analytic sphere or wall at a point, but the
lighting state is stored on the boundary proxy associated with that local
surface region.

Conceptually:
\begin{verbatim}
cell contains a small surface patch
boundary particle represents that patch
photon hits anywhere in that patch
boundary particle stores the patch color
renderer samples nearby patch colors
\end{verbatim}

The boundary particle keeps its base material identity.  The stored light color
is separate state:
\begin{verbatim}
material_id          = base surface material
boundary_space_light_rgb    = persistent lit color
boundary_space_light_valid  = whether the patch has received light
\end{verbatim}

The base material still answers what the surface is.  The boundary-space light
state answers what color that surface patch is currently emitting or showing.

\subsection{Photon Hit Rule}

The first rule should be intentionally simple:
\begin{verbatim}
photon hits surface proxy
proxy boundary_space_light_rgb = contacted material color
proxy boundary_space_light_valid = true
\end{verbatim}

The color does not clear every compute frame.  It persists until another photon
updates the same proxy.  This is a color-state rule, not an energy-accumulation
rule.

Because the first model sets color instead of adding energy, no atomic add is
required.  If two photons write the same proxy in the same compute dispatch,
the first implementation may accept last-writer-wins behavior.  If later work
requires additive light transport, then atomic accumulation can be introduced
as a separate, deliberate feature.

\subsection{Surface Spread Without Triangles}

The stored color spreads across the visible surface at sample/render time.  A
boundary proxy contributes color to nearby surface points according to an
influence radius and falloff kernel.

For a sphere, the natural distance is angular.  Given a lit proxy at position
\(\mathbf b\), a rendered surface point at \(\mathbf p\), and sphere center
\(\mathbf c\):
\[
    \mathbf n_b =
    \frac{\mathbf b-\mathbf c}{\lVert \mathbf b-\mathbf c\rVert},
    \qquad
    \mathbf n_p =
    \frac{\mathbf p-\mathbf c}{\lVert \mathbf p-\mathbf c\rVert}.
\]

The angular separation is:
\[
    \theta =
    \arccos(\mathrm{clamp}(\mathbf n_b \cdot \mathbf n_p,-1,1)).
\]

A simple linear kernel is:
\[
    w =
    \max\left(0, 1 - \frac{\theta}{\theta_L}\right),
\]
where \(\theta_L\) is the configured lighting patch angle.  A smoother kernel
can square the result:
\[
    w =
    \max\left(0, 1 - \frac{\theta}{\theta_L}\right)^2.
\]

The rendered color at \(\mathbf p\) is then derived from the nearby valid
boundary proxies:
\[
    \mathbf L(\mathbf p) =
    \frac{\sum_i w_i \mathbf C_i}{\max(\sum_i w_i,\epsilon)}.
\]

This gives a continuous lit region without drawing triangles or surface-cell
points.  The only visible geometry remains the existing particle/boundary
rendering path.

\subsection{Wall Spread}

For a rectangle wall, the same idea uses local wall coordinates rather than
spherical angle.  A wall has an origin, two surface axes, and dimensions.  A
point on the wall is projected into local coordinates:
\[
    u = (\mathbf p-\mathbf o)\cdot \mathbf a_u,
    \qquad
    v = (\mathbf p-\mathbf o)\cdot \mathbf a_v.
\]

The distance from a lit boundary proxy to a rendered wall point is measured in
the \((u,v)\) plane.  The same kernel form can be used:
\[
    w =
    \max\left(0, 1 - \frac{r}{R_L}\right)^2.
\]

\subsection{No Per-Frame Clear}

Boundary-space lighting is persistent.  It should not clear all light each
compute frame.  The state only changes when a photon hit writes a new patch
color, or when a later optional decay/filtering rule is enabled.

The first lifecycle is:
\begin{enumerate}
    \item initialize every boundary proxy with no valid light;
    \item photon contact writes the contacted material color into one boundary
          proxy;
    \item rendering samples nearby valid proxies and spreads their colors;
    \item a later photon may overwrite the proxy color.
\end{enumerate}

This is different from the previous boundary-light-buffer idea, where
\texttt{current} was cleared every frame and \texttt{filtered} was blended
over time.

\subsection{Future Configuration}

The first useful configuration items, when this model is reintroduced, are:
\begin{verbatim}
boundary_space_lighting_enabled = true;
boundary_space_patch_angle = 0.20;
boundary_space_patch_radius = 0.50;
boundary_space_patch_falloff = "quadratic";
\end{verbatim}

\texttt{boundary\_space\_patch\_angle} applies to spherical surfaces.
\texttt{boundary\_space\_patch\_radius} applies to planar wall surfaces.
Only the appropriate value is used for a given surface type.

\subsection{Python First Implementation Target}

Python can prove the model without changing the binary particle record.  Runtime
fields can be attached to boundary particles:
\begin{verbatim}
particle.boundary_space_light_valid = True
particle.boundary_space_light_rgb = (r, g, b)
\end{verbatim}

Photon contact handling should:
\begin{enumerate}
    \item identify the boundary proxy or analytic surface proxy;
    \item read the contacted surface material color;
    \item write that color into the proxy's persistent light state;
    \item leave photon debug color behavior separate.
\end{enumerate}

Rendering should:
\begin{enumerate}
    \item keep drawing existing boundary particles or analytic surface markers;
    \item for each rendered surface marker, gather nearby valid boundary-space
          light proxies;
    \item blend the material base color with the sampled boundary-space light;
    \item avoid drawing invisible surface-cell grids.
\end{enumerate}

\subsection{Point-List Rendering}

The Vulkan renderer currently uses:
\begin{verbatim}
inputAssembly.topology = VK_PRIMITIVE_TOPOLOGY_POINT_LIST;
\end{verbatim}

This topology does not interpolate color across a triangulated surface.  Each
boundary particle is submitted as a point primitive.  The spread across a
surface therefore has two separate parts:
\begin{itemize}
    \item surface-space color spread is computed by sampling nearby lit
          boundary proxies;
    \item screen-space patch shape is produced by rendering each point as a
          soft point sprite.
\end{itemize}

The vertex shader should choose the boundary particle's color by applying the
boundary-space kernel to nearby persistent proxy colors.  It should also set
\texttt{gl\_PointSize} large enough for the proxy to represent its cell-local
surface patch.

The fragment shader should use \texttt{gl\_PointCoord} to shape that point
primitive into a soft circular patch instead of a hard square:
\begin{verbatim}
vec2 p = gl_PointCoord * 2.0 - 1.0;
float r2 = dot(p, p);
if (r2 > 1.0) {
    discard;
}
float alpha = pow(max(0.0, 1.0 - r2), patchFalloff);
\end{verbatim}

The final fragment color is the boundary-space lighting color multiplied by
this alpha.  This does not create triangles and does not create a visible
surface grid.  It lets a boundary particle behave as a finite soft surface
patch while preserving the point-list pipeline.

\subsection{Vulkan Implication}

The Vulkan version does not need an additive SSBO for the first implementation.
It needs persistent boundary-space light state associated with boundary
particles or with a compact boundary-proxy table:
\begin{verbatim}
boundary_space_light_valid
boundary_space_light_rgb
surface_id
surface_type
normal or local surface coordinate
\end{verbatim}

Photon contact writes the proxy color.  Boundary rendering reads nearby proxy
colors and applies the spread kernel.  The fragment shader then uses
\texttt{gl\_PointCoord} to render each lit point as a soft circular patch.
Atomics are not required unless a future model changes from last-writer-wins
color state to additive energy accumulation.

\subsection{Decision}

The next lighting implementation should retire visible surface-cell rendering.
The target is boundary-space lighting:
\[
    \textrm{photon hit}
    \rightarrow
    \textrm{persistent boundary proxy color}
    \rightarrow
    \textrm{kernel-spread surface sample color}.
\]

This preserves the useful lesson from surface cells--lighting must belong to
surface space--without introducing triangles or a visible fake surface mesh.
