% !TeX root = GeoDynamics.tex
\section{Schedule-Wise Contact Resolution}

This section documents the current schedule-wise contact combination rule. The
goal is to make the Python path match the GPU execution model as closely as
possible before porting the dynamics to GLSL.

The rule is source-owned:
\begin{itemize}
    \item one source particle computes its own velocity update,
    \item the source particle may read target particles and wall contact state,
    \item the source particle does not write target particle velocity,
    \item pair contacts, wall contacts, and final escape correction are applied
          in a fixed schedule.
\end{itemize}

This replaces the older cluster solve for the GLSL-portable path. The cluster
solver is retained only as a diagnostic comparator.

\subsection{Definitions}

For a source particle \(i\), let its current velocity be
\[
    \mathbf{v}_i .
\]

For each active particle contact \(c\), the contact prediction function returns:
\[
    \mathbf{v}_{i,c}^{\ast}
\]
where \(\mathbf{v}_{i,c}^{\ast}\) is the velocity that contact \(c\) predicts
for source particle \(i\) if that contact were evaluated by itself.

The same contact state also stores the source velocity at first contact:
\[
    \mathbf{v}_{i,c}^{0}.
\]

The per-contact velocity delta is:
\[
    \Delta \mathbf{v}_{i,c}
    =
    \mathbf{v}_{i,c}^{\ast}
    -
    \mathbf{v}_{i,c}^{0}.
\]

\subsection{Algorithm}

For each source particle \(i\), the current schedule-wise algorithm is:

\begin{enumerate}
    \item Start with the particle's current velocity:
    \[
        \mathbf{v}_{base} = \mathbf{v}_i .
    \]

    \item Initialize the pair-contact accumulator:
    \[
        \Delta \mathbf{v}_{sum} = (0,0),
        \qquad
        N = 0 .
    \]

    \item For each active particle-particle contact owned by source particle
    \(i\):
    \begin{enumerate}
        \item Compute the single-contact predicted source velocity
        \(\mathbf{v}_{i,c}^{\ast}\).

        \item Read the first-contact source velocity
        \(\mathbf{v}_{i,c}^{0}\).

        \item If this is the first active pair contact, set:
        \[
            \mathbf{v}_{base} = \mathbf{v}_{i,c}^{0}.
        \]

        \item Accumulate the contact delta:
        \[
            \Delta \mathbf{v}_{sum}
            \leftarrow
            \Delta \mathbf{v}_{sum}
            +
            \left(
                \mathbf{v}_{i,c}^{\ast}
                -
                \mathbf{v}_{i,c}^{0}
            \right).
        \]

        \item Increment the active pair-contact count:
        \[
            N \leftarrow N + 1 .
        \]
    \end{enumerate}

    \item If \(N > 0\), combine the particle-particle contact deltas by
    averaging:
    \[
        \mathbf{v}_{pair}
        =
        \mathbf{v}_{base}
        +
        \frac{\Delta \mathbf{v}_{sum}}{N}.
    \]
    If \(N = 0\), use:
    \[
        \mathbf{v}_{pair} = \mathbf{v}_i .
    \]

    \item Apply wall contacts sequentially to \(\mathbf{v}_{pair}\). For each
    active wall contact \(w\), compute the wall-predicted normal velocity and
    replace only the normal component of the current accumulator:
    \[
        \mathbf{v}_{wall}
        =
        \mathbf{v}_{acc}
        +
        \left(
            v_{n,w}^{\ast}
            -
            \mathbf{v}_{acc}\cdot\mathbf{n}_w
        \right)
        \mathbf{n}_w .
    \]
    After each wall contact:
    \[
        \mathbf{v}_{acc} \leftarrow \mathbf{v}_{wall}.
    \]

    \item Apply the pair rebound escape scheduler. For each active rebound pair
    contact, compute the desired outward relative normal velocity. If the
    current relative normal velocity is too small, adjust only the source
    particle accumulator:
    \[
        \mathbf{v}_{acc}
        \leftarrow
        \mathbf{v}_{acc}
        -
        \frac{1}{2}
        \left(
            u_{desired}
            -
            u_{current}
        \right)
        \mathbf{n}_{ij}.
    \]

    \item Write the final accumulator back to the source particle:
    \[
        \mathbf{v}_i \leftarrow \mathbf{v}_{acc}.
    \]
\end{enumerate}

\subsection{Pseudocode}

\begin{verbatim}
for each source particle i:
    base_velocity = current_velocity(i)
    accumulated_delta = (0, 0)
    prediction_count = 0
    has_pair = false

    for each active particle contact c owned by i:
        prediction = pair_prediction(i, c)
        if prediction is invalid:
            continue

        start_velocity = prediction.start_velocity
        predicted_velocity = prediction.predicted_velocity

        if not has_pair:
            base_velocity = start_velocity
            has_pair = true

        accumulated_delta += predicted_velocity - start_velocity
        prediction_count += 1

    if has_pair:
        pair_velocity =
            base_velocity + accumulated_delta / prediction_count
    else:
        pair_velocity = current_velocity(i)

    final_velocity = pair_velocity

    for each active wall contact w owned by i:
        final_velocity = wall_prediction(i, w, final_velocity)

    for each active rebound pair contact c owned by i:
        final_velocity =
            apply_pair_escape_if_needed(i, c, final_velocity)

    velocity(i) = final_velocity
\end{verbatim}

\subsection{Known Question}

The current combination rule uses an average:
\[
    \frac{\Delta \mathbf{v}_{sum}}{N}.
\]

This is GPU-friendly and prevents the response from growing linearly with the
number of simultaneous contacts. However, it is not a coupled contact solve. In
dense contact groups the average can under-correct, over-correct, or interact
poorly with the later wall scheduler and pair escape scheduler.

This averaging rule is the next item to evaluate. In particular, we need to
decide whether dense contacts should use:
\begin{itemize}
    \item the current average of predicted deltas,
    \item a weighted average,
    \item a limited sum,
    \item a strongest-contact rule,
    \item or another source-owned combination rule.
\end{itemize}
