% !TeX root = GeoDynamics.tex
\section{Conventions}

This section records the working conventions used by the current geometry
dynamics model. The purpose is to make assumptions explicit so that later code,
diagnostics, and GLSL porting decisions can be checked against the same set of
rules.

\subsection{Particle Ownership}

\begin{itemize}[leftmargin=*]
    \item Contact processing is source-owned.
    \item A source particle may read target particle state.
    \item A source particle writes only its own velocity, contact state, and
          diagnostics.
    \item A source particle does not write the target particle.
    \item Pair contacts are therefore represented twice in source-owned
          storage: once from \(i\) to \(j\), and once from \(j\) to \(i\).
    \item This ownership rule is used because it is compatible with the GLSL
          execution model.
\end{itemize}

\subsection{Memory Ownership}

\begin{itemize}[leftmargin=*]
    \item Dynamics code should not depend on writable global memory shared by
          all particles.
    \item Global configuration values, such as wall locations and model
          constants, may be shared only as read-only inputs.
    \item A source particle may update only memory that it owns.
    \item Per-particle memory may be updated only when there is an explicit
          ownership rule for that field.
    \item Source-owned contact state is allowed because each source particle
          owns its own contact slots.
    \item Source-owned wall ghost momentum ledger state is allowed because it
          belongs to one real particle, is updated only by that particle's wall
          contact exit, and is not shared wall memory.
    \item A contact update must not write a target particle's contact state,
          velocity, internal momentum, or diagnostics.
    \item Reductions, aggregate diagnostics, and UI totals may be computed
          after the source-owned step, but they are not dynamics inputs.
    \item Any proposed persistent reservoir or shared accumulator must be
          rejected unless it has a GLSL-portable ownership rule.
\end{itemize}

\subsection{Contact Geometry}

\begin{itemize}[leftmargin=*]
    \item A particle-particle contact exists when the center distance is less
          than the sum of the radii:
    \[
        d < r_i + r_j .
    \]
    \item The contact normal \(\mathbf{n}_{ij}\) points from the source
          particle \(i\) to the target particle \(j\).
    \item Relative normal velocity is:
    \[
        u_{ij}
        =
        \left(
            \mathbf{v}_j - \mathbf{v}_i
        \right)
        \cdot
        \mathbf{n}_{ij}.
    \]
    \item Under this sign convention, \(u_{ij} < 0\) means closing or
          compression, and \(u_{ij} > 0\) means separating or rebound.
    \item Overlap area is a diagnostic and force-law input. It is not a hard
          clipping target.
\end{itemize}

\subsection{Contact-Normal Resolution}

\begin{itemize}[leftmargin=*]
    \item Contact dynamics are normal-owned.
    \item A contact may change only the source velocity component along that
          contact normal.
    \item The source tangential velocity must be carried from the current
          source velocity.
    \item Stored first-contact velocity may be used to determine the contact's
          normal compression and rebound history, but it must not restore a
          stale tangential velocity component.
    \item A contact may store full velocity vectors for diagnostics, but the
          dynamics transfer operation must apply only the normal component.
    \item This rule applies equally to particle-particle contacts and
          particle-wall ghost contacts.
    \item A tangential change is allowed only when a separate documented rule,
          such as friction, bonding, or surface deformation, owns that
          tangential component.
\end{itemize}

\subsection{Wall Contacts}

\begin{itemize}[leftmargin=*]
    \item Walls are detection objects.
    \item Once a wall contact is detected, the wall is represented as a ghost
          particle.
    \item The ghost particle is instantaneous. It exists only for the current
          contact evaluation and is rebuilt from the current source particle
          state.
    \item The ghost particle is not stored as a persistent particle and is not
          integrated as a moving particle.
    \item The ghost particle has the same radius and mass as the source
          particle.
    \item The ghost particle center is the mirror of the source particle center
          across the wall contact plane.
    \item For frictionless wall response, the ghost velocity mirrors only the
          wall-normal component of the source velocity. The tangential ghost
          velocity is zero unless a separate friction/deformation rule is
          introduced.
    \item After the instantaneous ghost has been constructed, wall contact
          dynamics should use the same source-owned contact transfer operation
          used for particle-particle contacts.
    \item Wall-specific reflection logic should not be used as a separate
          dynamics path after the ghost particle has been constructed.
    \item \textbf{wall\_boundary} is the actual physical wall plane.
    \item \textbf{wall\_contact\_offset} is an intentional offset from the
          physical wall plane.  A zero offset means the contact plane is the
          physical wall plane.
    \item \textbf{wall\_ghost\_center} is mirrored across the effective contact
          plane.  For the top wall:
    \[
        y_{ghost}
        =
        2 y_{contact\ plane} - y_{source} .
    \]
    \item \textbf{distance\_to\_wall} is always measured to the physical wall
          plane.
    \item \textbf{distance\_to\_contact\_plane} is measured to the effective
          contact plane after applying the wall contact offset.
\end{itemize}

\subsection{Compression, Rebound, and Zero Velocity}

\begin{itemize}[leftmargin=*]
    \item Compression means the relative normal velocity is closing:
    \[
        u_{ij} < 0 .
    \]
    \item Rebound means the relative normal velocity is separating:
    \[
        u_{ij} > 0 .
    \]
    \item The zero-velocity overlap is the overlap at which the normal kinetic
          momentum for the compression episode has been converted into internal
          momentum.
    \item The zero-velocity overlap is not necessarily a hard geometric stop.
          Other contacts may continue pushing the particles deeper.
    \item Once the current overlap reaches or exceeds the zero-velocity overlap,
          the contact must be eligible to return internal momentum.
    \item Passing the zero-velocity overlap is therefore a bookkeeping phase
          transition, not necessarily an immediate separation event.
    \item A contact should not silently move its zero-velocity target forward
          simply because the current overlap passed the old target.
\end{itemize}

\subsection{Energy Terminology}

\begin{itemize}[leftmargin=*]
    \item Energy calculations and energy-based names should not be used for
          the dynamics method except as diagnostics.
    \item The method should be described in terms of geometry, velocity,
          momentum, internal momentum, and contact state.
    \item Even when a diagnostic or derivation uses an energy-style calculation,
          that terminology should not become the name of a persistent dynamics
          state.
    \item For example, use internal momentum naming rather than energy naming
          for storage and return phases.
\end{itemize}

\subsection{Geometric Preference}

\begin{itemize}[leftmargin=*]
    \item When more than one solution is possible, prefer the solution that
          moves the model toward a purely geometric method.
    \item A preferred rule should be expressible using contact geometry,
          relative motion, source-owned state, and explicit diagnostics.
    \item Avoid adding non-geometric hidden reservoirs or correction terms
          unless they are needed for diagnostics or accounting and are clearly
          documented as such.
    \item If a temporary non-geometric rule is introduced, it should be marked
          as temporary and should not obscure the geometric interpretation of
          the contact.
\end{itemize}

\subsection{Internal Momentum}

\begin{itemize}[leftmargin=*]
    \item Internal momentum is the momentum stored in contact compression.
    \item During compression, internal momentum accumulates.
    \item During rebound, internal momentum is returned.
    \item Particle-side internal momentum is reported as a vector contribution
          to the particle momentum accounting.
    \item In a particle-particle contact, the two particle-side internal
          momentum vectors should cancel as a pair contribution.
    \item In a wall contact, the wall ghost internal momentum is equal and
          opposite to the particle-side wall-contact internal momentum.
\end{itemize}

\subsection{Momentum Accounting}

\begin{itemize}[leftmargin=*]
    \item Drift is a diagnostic, not a dynamics input.
    \item Current modeled total momentum is:
    \[
        \mathbf{p}_{current}
        =
        \mathbf{p}_{particle\ velocity}
        +
        \mathbf{p}_{particle\ side\ internal}
        +
        \mathbf{p}_{wall\ ghost\ internal}.
    \]
    \item Drift is:
    \[
        \mathbf{p}_{drift}
        =
        \mathbf{p}_{initial}
        -
        \mathbf{p}_{current}.
    \]
    \item When no contacts are active, drift should return to zero, up to
          numerical error.
    \item A nonzero drift during active contact may indicate stale contact
          state, incorrect source weighting, tangent/normal contamination, or
          an actual dynamics bug.
\end{itemize}

\subsection{Reservoirs}

\begin{itemize}[leftmargin=*]
    \item Persistent wall velocity reservoirs are not part of the current
          dynamics model.
    \item A reservoir shared by multiple particles is writable global memory
          and is not GLSL-portable without atomics or another explicit
          ownership rule.
    \item Wall ghost momentum is instantaneous contact state, not persistent
          wall memory.
    \item A wall ghost momentum ledger may persist on the source particle for
          accounting after a wall contact exits.  It is not a global wall
          reservoir and it must not feed back into dynamics.
    \item Wall-side internal/contact momentum may still be reported as a
          diagnostic equal and opposite to the particle-side wall-contact
          internal momentum.
    \item Diagnostic reservoirs may be computed as post-step reductions, but
          they must not feed back into the dynamics.
\end{itemize}

\subsection{Maximum Overlap}

\begin{itemize}[leftmargin=*]
    \item Particles may numerically tunnel or exceed the intended model overlap.
    \item The simulation should not clip this automatically during debugging,
          because clipping can hide bugs.
    \item For diagnostics, a particle-particle overlap deeper than \(50\%\) is
          considered an error.
    \item For equal-radius particles, this means:
    \[
        d < r .
    \]
    \item More generally, the maximum allowed penetration depth is:
    \[
        \alpha_{max} = \min(r_i, r_j).
    \]
    \item The guardrail reports this condition and pauses the UI, but it does
          not change positions or velocities.
\end{itemize}

\subsection{Heuristics and Diagnostics}

\begin{itemize}[leftmargin=*]
    \item Heuristics that silently repair state can hide dynamics bugs.
    \item Diagnostics should prefer reporting impossible or suspicious states
          over changing the dynamics.
    \item Recovery rules must be documented separately from the physical model.
    \item Any rule that promotes a zero-velocity overlap to the current overlap
          should be treated as a diagnostic warning or explicit policy, not as
          an invisible correction.
\end{itemize}

\subsection{Schedule-Wise Resolution}

\begin{itemize}[leftmargin=*]
    \item The GLSL-portable path uses schedule-wise source-owned updates.
    \item Pair contacts produce per-contact source velocity predictions.
    \item Multiple contacts for one source must be combined by an explicit
          source-owned rule.
    \item The current geometric direction is area-weighted source-owned contact
          resolution, not unweighted averaging of independent full-contact
          predictions.
    \item Per-contact predictions must be combined as normal-component
          transfers. They must not be combined as full-vector velocity restores.
    \item Pair escape correction must not apply a second, hidden dynamics rule
          after source-owned contact resolution.
    \item The old cluster solver is retained only as a Python diagnostic
          comparator.
    \item Any rule that depends on writing both particles in a pair at once is
          not GLSL-portable under the current source-owned convention.
\end{itemize}

\subsection{Snapshot and Capture Diagnostics}

\begin{itemize}[leftmargin=*]
    \item Capture files are diagnostic artifacts.
    \item Contact snapshots preserve source-owned contact state for reproducing
          a specific frame.
    \item Particle-data snapshots preserve positions and velocities so a frame
          can be turned into a focused configuration.
    \item Guardrail errors should be written to capture files and shown in the
          UI.
    \item The UI should pause when a guardrail error is present.
\end{itemize}

\subsection{Python and GLSL Porting}

\begin{itemize}[leftmargin=*]
    \item Python is the reference implementation while the model is changing.
    \item GLSL should be regenerated from the reference behavior rather than
          patched by hand.
    \item Python structures may contain extra diagnostics that are not required
          in GLSL.
    \item Persistent state needed by the dynamics should be distinguished from
          reporting-only state before porting.
    \item A dynamics rule should be expressible as source-owned operations
          before it is considered ready for GLSL.
\end{itemize}
