% !TeX root = GeoDynamics.tex
\section{Drift calculations}

Drift calculation is an accumulation of momentum changes during a simulation. The drift is:


\begin{equation}
\begin{aligned}
\mathbf{p}_{CurrentTotalMomentum} = \mathbf{p}_{ParticleVelocityMomentum} \\ + \mathbf{p}_{ ParticleSideInternalMomentum} + \mathbf{p}_{WallGhostInternalMomentum} + \mathbf{p}_{WallVelocityMomentumReservoir}
\end{aligned}
\end{equation}

Where velocity momentum, $\mathbf{p}_{ParticleVelocityMomentum} $, is the sum of all the particles translational momentum. Internal momentum, $\mathbf{p}_{ParticleSideInternalMomentum}$. Wall momentum \\ $\mathbf{p}_{ WallGhostInternalMomentum}$ is the value of the wall internal momentum. And \\ $\mathbf{p}_{WallVelocityMomentumReservoir}$ is the pseudo wall reaction velocity.

The total starting momentum $\mathbf{p}_{InitialTotalMomentum}$, minus the current total momentum \\  $\mathbf{p}_{CurrentTotalMomentum}$ is drift $\mathbf{p}_{Drift}$.

Reporting variable perform a running sum where as each particles momentum values are recorded by adding them to previous reporting values. This means for each particle, at a time step, the kinetic momentum is added to the global variable $\mathbf{p}_{ ParticleVelocity Momentum}$, and likewise for the other variables. They are independently assigned to reporting variables. At the end of the time step the filled reporting variables are used to calculate the drift.  

Drift tells us how much the total system is varying from the current system and shows us the accumulation/system errors. The  accumulation/system errors become very important in large scale simulation of millions of particles.

The model of wall internal momentum may not match the actual dynamics but it is viewed as follows: The wall/ghost internal momentum ,$\mathbf{p}_{WallGhostInternalMomentum}$ is equal and opposite to the particle-side,$\mathbf{p}_{ParticleSideInternalMomentum}$  wall-contact internal momentum. 

The model of particle internal momentum may not match the actual dynamics but it is viewed as follows:Each particles internal momentum accumulates during compression and feeds back during rebound. During compression it is a vector opposite the velocity vector. In this way the internal momentum is fed back in the same direction as rebound. 

Particle-particle collisions result in four offsetting vectors, two opposing velocity momentum vectors and two opposing internal momentum vectors. Particle-wall collisions differ mathematically since the wall does not move. There are then only three vectors - two internal momentum vectors and only one velocity momentum vector. This leaves an imbalance of one velocity momentum vector. To account for this we create a wall-side velocity momentum reservoir $\mathbf{p}_{WallVelocityMomentumReservoir}$.

\subsection{Reporting Values}

The UI reports the momentum accounting values as vector rows.  Each row shows
the x component, the y component, and the vector magnitude:

\begin{equation}
\begin{aligned}
\mathbf{p} = (p_x, p_y), \qquad |\mathbf{p}| = \sqrt{p_x^2+p_y^2}
\end{aligned}
\end{equation}

The UI rows are:

\begin{itemize}
\item \textbf{start}: The initial total translational momentum of all active
particles at the start of the run.  This is the reference value used for drift
calculation.

\item \textbf{particle\_velocity}: The current translational momentum of all
active particles.  For each particle,
\begin{equation}
\begin{aligned}
\mathbf{p}_{particle} = m\mathbf{v}
\end{aligned}
\end{equation}
and the row is the vector sum over all active particles.

\item \textbf{particle\_side\_int}: The particle-side internal momentum vector
currently stored in active contacts.  For a particle-particle contact, this is
reported on the particle side of the contact normal.  For a wall contact, this
is the particle-side contact internal momentum, opposite the wall/ghost side.
This value is a signed vector sum, not a scalar sum.

\item \textbf{wall\_ghost\_int}: The wall/ghost-side internal momentum vector
for active wall contacts.  Under the current reporting convention, this is
equal and opposite to the particle-side wall-contact internal momentum.  It is
included so that wall contacts can be reported as if the wall contact has a
ghost-side internal state.

\item \textbf{wall\_vel\_reserv}: The wall velocity momentum reservoir.  This
is the pseudo wall reaction velocity momentum used to account for momentum that
would belong to the non-moving wall side of a wall collision.  It is zero when
no wall velocity reservoir has been accumulated.

\item \textbf{modeled\_total}: The current reported system momentum after
adding all reported momentum components:
\begin{equation}
\begin{aligned}
\mathbf{p}_{modeled}
=
\mathbf{p}_{particle\_velocity}
+ \mathbf{p}_{particle\_side\_int}
+ \mathbf{p}_{wall\_ghost\_int}
+ \mathbf{p}_{wall\_vel\_reserv}
\end{aligned}
\end{equation}

\item \textbf{drift}: The difference between the starting total momentum and
the modeled total momentum:
\begin{equation}
\begin{aligned}
\mathbf{p}_{drift}
=
\mathbf{p}_{start}
-
\mathbf{p}_{modeled}
\end{aligned}
\end{equation}
When the accounting and dynamics are consistent for the selected model, this
row should be zero after all contacts have separated and all internal momentum
has returned to translational momentum.
\end{itemize}

The capture files write these same values in the \texttt{[momentum\_balance]}
section.  They also write a more detailed \texttt{[momentum\_diagnostics]}
section.  The diagnostic names are more explicit and include aliases used while
debugging:

\begin{itemize}
\item \texttt{start\_px}, \texttt{start\_py}, and \texttt{start\_p}: component
and magnitude form of the UI \textbf{start} row.

\item \texttt{current\_px}, \texttt{current\_py}, and \texttt{current\_p}:
component and magnitude form of the current particle translational momentum.
These are equivalent to \texttt{particle\_velocity\_*}.

\item \texttt{particle\_velocity\_px}, \texttt{particle\_velocity\_py}, and
\texttt{particle\_velocity\_p}: component and magnitude form of the UI
\textbf{particle\_velocity} row.

\item \texttt{particle\_side\_internal\_px},
\texttt{particle\_side\_internal\_py}, and
\texttt{particle\_side\_internal\_p}: component and magnitude form of the UI
\textbf{particle\_side\_int} row.

\item \texttt{particle\_pair\_internal\_*} and \texttt{pair\_internal\_*}: the
particle-particle portion of particle-side internal momentum.  These exclude
wall contacts.

\item \texttt{wall\_internal\_*} and \texttt{wall\_ghost\_internal\_*}: the
wall/ghost-side internal momentum for active wall contacts.  These are aliases
for the same reported vector.

\item \texttt{contact\_internal\_*}: the vector sum of particle-side internal
momentum and wall/ghost internal momentum.  Because the current wall convention
uses equal and opposite particle-side and wall/ghost-side internal vectors,
wall-contact internal momentum can cancel inside this value.

\item \texttt{wall\_velocity\_reservoir\_*} and \texttt{wall\_ghost\_*}: the
wall velocity momentum reservoir.  These are aliases for the same vector.

\item \texttt{modeled\_total\_without\_reservoir\_*}: the current particle
momentum plus contact internal momentum, before adding the wall velocity
momentum reservoir.

\item \texttt{modeled\_total\_*} and
\texttt{modeled\_total\_with\_wall\_*}: the full modeled total used by the UI
\textbf{modeled\_total} row.

\item \texttt{drift\_modeled\_*} and \texttt{drift\_with\_wall\_ghost\_*}: the
full drift vector used by the UI \textbf{drift} row.

\item \texttt{drift\_without\_reservoir\_*}: the drift that would be reported
if the wall velocity momentum reservoir were omitted.

\item \texttt{pair\_internal\_scalar}: scalar sum of active particle-particle
contact internal momentum magnitudes.  This is a diagnostic scalar and is not
used directly in the vector drift calculation.

\item \texttt{wall\_internal\_scalar}: scalar sum of active wall-contact
internal momentum magnitudes.  This is a diagnostic scalar and is not used
directly in the vector drift calculation.

\item \texttt{contact\_internal\_scalar}: scalar sum of pair and wall internal
momentum magnitudes.  This is useful for seeing whether internal momentum is
present too briefly to observe in the UI.

\item \texttt{contact\_stored\_scalar},
\texttt{contact\_released\_scalar}, and
\texttt{contact\_remaining\_scalar}: scalar diagnostic totals for stored,
released, and remaining internal contact momentum.

\item \texttt{active\_pair\_contacts} and \texttt{active\_wall\_contacts}: the
number of active particle-particle and particle-wall contacts in the current
frame.
\end{itemize}

The vector rows are the authoritative values for drift accounting.  The scalar
rows are diagnostics only; they help show whether a contact is accumulating,
returning, or clear, but they are not substituted into the drift vector
equation.

