
% !TeX root = main.tex
\section{Dynamics Conventions}

\subsection{Purpose}

The project maintains independently selectable dynamics models.  Each model is
a readable Python implementation whose source-complete structure is intended
to transfer to GLSL with minimal translation.

\subsection{Source-Owned Execution}

Each source particle reads immutable position and velocity snapshots,
processes all of its current contacts, vector-sums the planned contact
response, and writes only its own state.  A source must not write a target
particle.

The opposing source evaluation reads the same frame-start geometry.  Pair
compatibility and whole-system conservation must therefore emerge from the
contact rule used independently by both sources.

\subsection{Contact Geometry Convention}

Contact overlap is a geometric property of the current frame-start particle
positions and radii.  For a particle-particle contact, the source-side and
target-side contact geometry is the same geometry, even when the two particles
have different velocities.  Therefore, once contact geometry is computed for a
source-target pair, the target particle's contact geometry is known from the
same calculation.

Velocity may affect momentum transfer, phase, or future position, but it does
not make the instantaneous geometric overlap different for the two particles in
the same contact.

\texttt{ForceDynamics} currently supports two scalar contact measures:

\begin{itemize}
    \item penetration depth, \(\delta_{ij}\), the default force measure;
    \item overlap area, \(A_{ij}\), retained for diagnostics and optional
    area-force comparison.
\end{itemize}

Both measures are geometric and source-target symmetric for a particle pair.
Documentation must not assume that overlap area is the active force measure
unless \texttt{contact\_force\_measure = "area"} is explicitly selected.

For depth-mode particle contacts, the project uses the following vector names.
Let \(\mathbf n_{ij}\) be the unit collision normal from the source particle
center to the target particle center:

\begin{itemize}
    \item source orientation vector:
    \(\mathbf O_i = r_i\mathbf n_{ij}\);
    \item source-to-target-leading-edge proximity vector:
    \(\mathbf P_{ij}=(\mathbf x_j-r_j\mathbf n_{ij})-\mathbf x_i\);
    \item penetration vector:
    \(\mathbf D_{ij}=\mathbf O_i-\mathbf P_{ij}\).
\end{itemize}

The magnitude of the source orientation vector is the source radius.  The
magnitude of the proximity vector is the distance from the source center to the
leading edge of the target particle along the collision line.  The penetration
depth is the orientation magnitude minus the proximity magnitude:

\begin{equation}
\delta_{ij}=\lVert\mathbf O_i\rVert-\lVert\mathbf P_{ij}\rVert
=r_i-(d_{ij}-r_j)=r_i+r_j-d_{ij}.
\end{equation}

Thus the proximity vector points to the intruding target edge, but the depth of
penetration is the distance from the source boundary to that target leading
edge.  The penetration vector has this depth as its magnitude when the vectors
are collinear.

\subsection{Frame State}

Current contact geometry and relative motion are calculated from the immutable
frame-start snapshot:

\begin{itemize}
    \item current positions;
    \item current velocities;
    \item current radii and masses;
    \item current overlap geometry;
    \item the current source-owned contact representation.
\end{itemize}

\texttt{ForceDynamics} immediately accumulates each detected contact into a
source-local force sum; its contact slots are diagnostic only.

\subsection{Force-Field Contact Response}

\texttt{ForceDynamics} treats contact geometry as a central repulsive pair
force:

\begin{equation}
F_{ij}=q_{ij}M_{ij}.
\end{equation}

Here \(M_{ij}\) is the selected scalar contact measure.  In the current default
\texttt{depth} mode, \(M_{ij}=\delta_{ij}\).  In optional \texttt{area} mode,
\(M_{ij}=A_{ij}\).

Every overlapping frame applies:

\begin{equation}
\Delta\mathbf p_i=-F_{ij}\mathbf n_{ij}\Delta t.
\end{equation}

There are no force-law compression/release phases and no available-momentum
caps.  Source-owned internal momentum may be recorded as diagnostic
bookkeeping, but it does not currently feed back into the force calculation or
the velocity update.  The corresponding diagnostic pair potential is:

\begin{equation}
U_{ij}(d)=q_{ij}\int_d^{r_i+r_j}M_{ij}(s)\,ds.
\end{equation}

The ordinary Force contact law is instantaneous and purely geometric.  Each
timestep scans current particle and wall contacts once and immediately
vector-sums their contributions into the source-local \texttt{totalForce}.
After the source force sum is complete, semi-implicit Euler first updates
velocity and then updates position using the new velocity.  An end-position
snapshot is created only for diagnostics; it is not a second force evaluation.
Relative velocity is diagnostic only and does not affect force magnitude.

Starting contacts are the exception to memoryless operation.  When particles
or wall contacts are already overlapping at initialization, the starting
separation is recorded as an effective zero-force contact limit.  The model
applies force only if the contact compresses deeper than that starting
separation.  After rebound to the starting separation, the physical overlap is
suppressed until the actual radii separate, preventing a second rebound from
the configured initial overlap.

\subsection{Diagnostics}

Momentum, kinetic energy, potential energy, and total energy are reported
diagnostics.  Diagnostic calculations evaluate the contact rule but do not
affect particle motion.  Force potential energy is calculated from the same
physical or effective geometry used by the force calculation.  Documentation
must distinguish actual vector momentum drift from any scalar diagnostic
totals.  Diagnostic return to zero does not, by itself, prove momentum or
kinetic-energy conservation.

\subsection{Momentum-First Causality}

The dynamics model is force and momentum governed.  Energy is diagnostic
bookkeeping, not an input correction.  No particle is assumed to know the
future collision arc, the final rebound state, or a global energy target.  A
model update must be expressible from current geometry, current force, current
momentum, and stored local contact state.

Therefore global or after-the-fact energy repair must not be used to drive
motion.  If a crowded contact cannot resolve all compression as translational
motion during the current timestep, the unresolved part belongs in a local
internal momentum or stress ledger, not in an energy correction.  Later release
must occur only through current contact geometry and current contact normals.

Energy diagnostics may expose whether the bookkeeping is consistent, but the
governing question is where the impulse went: translational particle momentum,
stored contact/internal momentum, or wall/constraint momentum.

\subsection{Internal Momentum And Heat}

Internal momentum is the local bookkeeping layer for compression that cannot be
fully expressed as translational motion during the current timestep.  It is not
a global energy repair, and it must not require knowledge of the future contact
arc.  A particle may only record or release internal momentum from current
contact geometry, current relative motion, current contact normal, and its own
source-owned state.

The current \texttt{ForceDynamics} implementation records diagnostic
recoverable internal momentum in \texttt{parms.yzw}.  \texttt{parms.x} remains
particle mass.  During compression, contact impulse is added along the current
normal.  During rebound, the stored component along the current normal is
drained.  The scalar magnitude is mirrored to \texttt{Data.z} for reporting.
This value is diagnostic only and is not read by \texttt{CalcVelocity}.

\texttt{Data.z} has a type-dependent meaning.  For mobile particles
(\texttt{ptype=0}), it is the internal-momentum magnitude described above.  A
boundary marker is never processed as a mobile source.  Its positive
\texttt{pdata.ptype} value is the wall-evaluator ID: horizontal is 1, vertical
is 2, and CD nozzle is 3.  Python and the Vulkan loader copy that value into
runtime \texttt{ptype} and \texttt{Data.z}.  The binary
\texttt{pdata.temp\_vel} field remains reserved and is not part of boundary
dispatch.  Per-frame contact reset and internal-momentum synchronization must
not overwrite \texttt{Data.z} for boundary particles.

Heat is a later material model layer.  Real compressed matter can convert some
stored mechanical compression into thermal state and later transfer that heat
to surrounding particles or boundaries according to a heat-transfer
coefficient.  That future model must still be local and causal: heat may be
derived from force, momentum, compression, and material coefficients, but it
must not be introduced as an after-the-fact command to make total energy match
a desired value.

\subsection{Visual Diagnostics}

Particle color is diagnostic display state.  It must not affect collision
detection, force accumulation, integration, lifecycle state, or energy and
momentum reports.

When \texttt{RUN\_CONFIGURATION.hsv\_color} is enabled in the Python viewer,
or when \texttt{HSV\_ON} is enabled in a Vulkan scenario
\texttt{params.glsl}, particle color is derived from the velocity angle stored
in \texttt{VelRad.w}.  The hue maps angle to the unit interval with
\(h=\theta/(2\pi)\), wrapped modulo one.  The configuration values
\texttt{hsv\_sat}/\texttt{HSV\_SAT} and \texttt{hsv\_val}/\texttt{HSV\_VAL}
control saturation and value and must remain in the interval \([0,1]\).

This colorization is a visualization aid for flow direction and particle
motion.  The dynamics model owns the velocity angle; the renderer owns the
conversion from angle to RGB.

For ordinary mobile particles, the collision-color flag \texttt{colFlg} is
diagnostic display state.  Shader paths that use it must clear it once per
source invocation and set it only when a valid particle or wall contact is
accumulated for the current frame.  The retired deferred-birth investigation
reserved additional values, but the live model no longer uses
\texttt{colFlg} as lifecycle state.

Rendered sphere radius must be derived from the particle radius stored in
\texttt{Data.x}.  Mesh files may provide a convenient unit sphere, but they
must not bake a separate configuration radius into the rendered particle size.

\subsection{Reservoir Deferred-Birth Investigation}

This subsection records the deferred-birth design investigated for the staged
reservoir.  It was not implemented and is no longer the selected next
reservoir model.  It remains here because it defines the failure discovered in
the staged-birth system and the synchronization cost of repairing that system.

A reservoir boundary is a lifecycle boundary rather than a solid contact
wall.  Reservoir particles are created with scheduled birth frames and remain
outside ordinary dynamics until successfully admitted into the live particle
population.  The authoritative lifecycle field is \texttt{Data.w}:

\begin{itemize}
    \item \(\texttt{Data.w}>0\): waiting particle; the value is its next birth
    attempt frame;
    \item \(\texttt{Data.w}=0\): live particle;
    \item \(\texttt{Data.w}<0\): dead particle.
\end{itemize}

The reservoir uses the following reserved \texttt{colFlg} values during birth:

\begin{itemize}
    \item \(\texttt{colFlg}=2\): waiting for its scheduled or deferred birth
    attempt;
    \item \(\texttt{colFlg}=3\): birth candidate whose scheduled frame has
    arrived and which wants to become live;
    \item \(\texttt{colFlg}=0\): successfully admitted live particle with no
    current diagnostic collision color.
\end{itemize}

When the current frame reaches a waiting particle's value in
\texttt{Data.w}, the source becomes a birth candidate.  It is placed at its
configured birth position and tested for particle overlap before it is allowed
to participate in force accumulation or motion.  This test concerns only
particle starting overlap; wall starting overlap is a separate problem.

A birth candidate is compared only with already-live mobile targets.  A valid
birth-blocking target must satisfy all of the following:

\begin{itemize}
    \item the target is not the source;
    \item the target is a mobile particle, not a boundary marker;
    \item \(\texttt{Target.Data.w}=0\);
    \item the target is not another waiting or birth-candidate particle.
\end{itemize}

Dead particles, unborn particles, other birth candidates, and boundary marker
particles do not block this birth test.  Consequently, particles scheduled for
the same frame must already have nonoverlapping configured birth positions;
the live-target test does not resolve overlap among simultaneous candidates.

The contact search must distinguish an ordinary live-particle contact from a
birth-blocking starting collision.  A boolean result is insufficient for the
wrapper because it cannot indicate whether ordinary force processing should
continue.  The motion contact function should therefore return distinct
results equivalent to:

\begin{verbatim}
NO_CONTACT = 0
CONTACT = 1
STARTING_COLLISION = 2
\end{verbatim}

On the first \texttt{STARTING\_COLLISION}, the source invocation must stop its
remaining neighbor search.  The candidate is returned to waiting state:

\begin{equation}
\texttt{Data.w}=\texttt{frameNum}+1,
\qquad
\texttt{colFlg}=2.
\end{equation}

It receives no contact force, velocity update, or position update during that
frame.  On the next frame it attempts birth again from the configured birth
position.  Each blocked particle is delayed independently by one frame.  The
original scheduled frames of later particles and waves are not shifted, so the
generator's original release spacing remains unchanged.

If the complete live-target search finds no starting collision, birth
succeeds:

\begin{equation}
\texttt{Data.w}=0,
\qquad
\texttt{colFlg}=0.
\end{equation}

The particle may then enter ordinary contact, force, velocity, and position
processing as a live source.

GPU admission must not depend on invocation execution order.  If one birth
candidate writes \texttt{Data.w}=0 while another candidate is still scanning,
the second candidate may observe either the old or new state.  Therefore birth
testing and birth-state commitment require separate phases, separate snapshots,
or an equivalent synchronization design.  During the test phase, target
liveness must come from an immutable frame-start lifecycle state.  This keeps
the rule source-owned and deterministic without allowing one source invocation
to modify another particle.

\subsection{Prefilled Moving-Piston Reservoir Convention}

The selected next reservoir model removes runtime particle birth.  A finite
reservoir is prefilled with nonoverlapping mobile particles.  At the release
frame the mobile population and rear wall begin moving in the flow direction.
The rear wall then moves with the configured inlet velocity and acts as a
piston.  Particle and piston motion must begin together; otherwise an
equal-speed piston cannot close a gap created before it starts.

The piston supplies external impulse, so particle momentum alone is not a
closed-system conservation diagnostic.  Runtime diagnostics must record
particle momentum together with piston impulse.  Contact response must use
particle velocity relative to piston velocity rather than stationary-wall
relative velocity.

The useful run duration is finite and is set by reservoir length, piston
speed, and distance to the nozzle entrance.  Downstream restriction may slow
and compress particles while the piston maintains supply without creating
geometric overlap through birth.

The cell domain should be rectangular rather than cubic.  Independent
\texttt{CellAryW}, \texttt{CellAryH}, and \texttt{CellAryL} dimensions bound
the long flow axis, narrow cross-axis span, and shallow 2D depth.  Allocation,
index validation, cell guards, and loops must use the corresponding axis
dimension.  Linear indexing remains:
\begin{equation}
i=x+W(y+Hz).
\end{equation}

\subsection{Boundary Convention}

Force-model walls must use a conservative geometric response compatible with
the model's integration method.  A fixed wall may contribute force and
potential energy to a source particle, but it is not integrated and must not
receive a particle-state write.  Current \texttt{ForceDynamics} scans each of
the four configured walls exactly once for each source and accumulates wall
force into the same source-local \texttt{totalForce} used for particle
contacts.  A wall contact uses stationary equal-particle ghost geometry and
the configured \texttt{wall\_contact\_offset}.  Starting wall contacts retain
the physical wall ghost for clearance but use an effective ghost radius while
the starting overlap is active.

Long-running closed-box energy statistics are meaningful only after wall work
and wall potential energy are included in the same diagnostics as
particle-particle contacts.
