% !TeX root = main.tex
\newpage
\section{Simple Reservoir Flow}

\subsection{Purpose}

\texttt{SimpleReservoir} is a particle birth/life/death process used with
\texttt{ForceDynamics}.  It does not change the ForceDynamics contact-force
law or the configured \texttt{contact\_force\_measure}.  It decides when
particles become active, which particles are eligible for dynamics, and when
particles leave the active simulation.

The scheduled-birth reservoir documented below is the current implementation,
not the selected next architecture.  Capture diagnostics showed that a slowed
particle can remain at the inlet when a later wave is admitted.  The resulting
deep starting overlap preserves pair momentum but injects artificial contact
compression and kinetic energy.  The selected replacement is the prefilled
moving-piston reservoir defined in \texttt{DynamicsConventions.tex}.

\subsection{State Field}

For \texttt{SimpleReservoir}, the model-owned particle state is stored in
\texttt{Particle.Data.w}.  When particles are loaded from a binary
\texttt{pdata} file, the input field \texttt{state\_flg} is copied into
\texttt{Particle.Data.w}.

The meaning of \texttt{Data.w} is:

\begin{itemize}
    \item \(\texttt{Data.w} > 0\): unborn particle.  The value is the frame on
    which the particle becomes active.
    \item \(\texttt{Data.w} = 0\): live particle.  The particle participates in
    ForceDynamics.
    \item \(\texttt{Data.w} < 0\): dead particle.  The particle is removed from
    ForceDynamics.
\end{itemize}

These meanings apply only when the active configuration selects
\texttt{SimpleReservoir}.  Other models may assign different meanings to
\texttt{Data.w}.

\subsection{Birth Rule}

At the beginning of each frame, before contact detection, every unborn
particle is tested against the current frame number:

\[
\texttt{if Data.w > 0 and frameNum >= Data.w}
\]

then the particle is born.  Birth performs:

\begin{itemize}
    \item set \(\texttt{Data.w}=0\);
    \item place the particle at the inlet birth position;
    \item assign the inlet velocity;
    \item make both position buffers agree on the birth position.
\end{itemize}

A particle is born before the frame-start snapshot is built, so the newly born
particle participates in the current frame using ordinary ForceDynamics rules.

This is the current implementation.  It does not test whether the proposed
birth position overlaps an already-live mobile particle.  Deferred admission
was analyzed as one repair, but the selected next model removes runtime birth
and uses a prefilled moving-piston reservoir.

In the GLSL compute path this rule is owned by:

\begin{verbatim}
StartReservoir(SourceID)
\end{verbatim}

The compute shader calls this function before particle contact scans, wall
contact scans, velocity update, or position update.  If the function returns
\texttt{false}, the source invocation returns immediately.  This means unborn
and dead particles do not participate in the current frame.  If the source
particle reaches its stored birth frame, \texttt{StartReservoir} resets the
active position to the inlet birth location, updates both position buffers,
sets \texttt{Data.w=0}, and returns \texttt{true}.

\subsection{Live Rule}

A particle is live when:

\[
\texttt{Data.w = 0}.
\]

Only live particles are eligible for:

\begin{itemize}
    \item particle-particle contact detection;
    \item wall contact detection;
    \item force accumulation;
    \item velocity update;
    \item position update;
    \item energy and momentum diagnostics.
\end{itemize}

Unborn and dead particles are skipped by the dynamics pass.

\subsection{Death Rule}

After position update, each live particle is tested against the outlet
boundary.  If the particle has crossed the outlet boundary, then:

\[
\texttt{Data.w = -1}.
\]

A dead particle no longer participates in ForceDynamics and is not reused by
\texttt{SimpleReservoir}.

In the GLSL compute path this rule is owned by:

\begin{verbatim}
RetireParticlePastXMax(SourceID)
\end{verbatim}

The compute shader calls this function after \texttt{CalcPosition(SourceID)}.
It reads the next position buffer, the buffer just written by
\texttt{CalcPosition}, and sets \texttt{Data.w=-1} when the particle has
passed the selected outlet boundary.  For horizontal reservoir flow this is
the \(x_{\max}\) outlet.  For vertical reservoir flow this is the
\(y_{\max}\) outlet.  The legacy function name is retained for shader-wrapper
compatibility even though the implementation now respects the generated
\texttt{PERIODIC\_DIRECTION} selector.

\subsection{Boundary Rule}

For \texttt{SimpleReservoir}, the inlet and outlet are not solid walls.  They
are lifecycle planes.  Only the pipe walls are solid ForceDynamics walls.

For the current two-dimensional pipe convention:

\begin{itemize}
    \item \(x_{\min}\): inlet birth plane, not a wall;
    \item \(x_{\max}\): outlet death plane, not a wall;
    \item \(y_{\min}\): solid lower wall;
    \item \(y_{\max}\): solid upper wall.
\end{itemize}

Therefore the wall scan for \texttt{SimpleReservoir} includes only wall flags
3 and 4.

\subsection{Boundary Particle Wall Rule}

Reservoir configurations may select the wall representation with:

\begin{verbatim}
wall_type = 0;  // manual configured wall planes
wall_type = 1;  // particle boundary markers
\end{verbatim}

When \texttt{wall\_type=0}, the configured wall fields retain their legacy
meaning as physical wall planes.  When \texttt{wall\_type=1}, boundary
particles define the solid wall broad phase.  The configured manual wall
planes must not be used as the physical wall-collision geometry for those
pipe walls.

A boundary particle is an occupancy marker, not a moving dynamics particle.
It is stored in the same particle buffer with a positive evaluator-valued
\texttt{ptype} and
\texttt{Data.w=0.0} so the vertex/cell placement pass includes it.  The
compute/source dynamics dispatch still processes only mobile particles as
sources.  When a mobile source encounters a boundary marker as a target, it
must evaluate the wall segment represented by that marker rather than process
the marker as an ordinary particle-particle collision.

Each newly generated boundary marker is self-describing.  Its evaluator ID is
written to \texttt{pdata.ptype} and copied into runtime \texttt{ptype} and
\texttt{Data.z}:
\begin{verbatim}
1 = horizontal_wall
2 = vertical_wall
3 = cd_nozzle_wall
\end{verbatim}
\texttt{EvaluateWallSegment(SourceID, BoundaryID)} dispatches from this target-
owned value.  There is no runtime global-function fallback.  Binary files whose
boundary markers have evaluator ID zero must be regenerated.

The current straight-wall implementation is:

\begin{verbatim}
EvaluateWallSegment(SourceID, BoundaryID)
\end{verbatim}

For now, each boundary marker represents either a straight horizontal or
straight vertical wall segment through the center of the marker's cell.  The
generator configuration selects which evaluator ID is written into each
marker; runtime Python dispatch reads that marker-owned ID from
\texttt{Data.z}.  Horizontal segments classify markers as lower or upper
walls.  Vertical segments classify markers as left or right walls.  The
evaluator calculates the current source-to-segment wall ghost geometry and
returns the normal, overlap area, center distance, and wall flag needed by the
ordinary wall contact slot.  Later sloped or curved boundaries should extend
\texttt{EvaluateWallSegment} without changing the source-owned force
accumulation rule.

Python and Vulkan have completed this marker-owned dispatch migration.  The
Vulkan loader copies positive \texttt{pdata.ptype} values into
\texttt{Data.z}, and generated \texttt{EvaluateWallSegment} dispatches from
the target marker's value.  The old global \texttt{WALL\_FUNC} evaluator
decision has been removed.  \texttt{WALL\_MODEL\_BOUNDARY\_PARTICLE} remains
the wrapper-level selection that enables marker-target wall processing.

The closed horizontal reservoir test emits horizontal-evaluator markers along
the derived \(y_{\min}\) and \(y_{\max}\) walls and vertical-evaluator markers
along \(x_{\min}\) and \(x_{\max}\).  Marker coordinates are deduplicated, so
an exact corner coordinate is stored once.  The inlet wall is
\texttt{reservoir\_inlet\_x}; the outlet wall is that inlet plus
\texttt{wall\_boundary\_length}.  The lifecycle retirement plane is configured
separately beyond the physical outlet wall, so a normal wall rebound is
attempted before a particle can be retired as an escape.

The vertical periodic test uses the rotated form of the same construction:
exact horizontal inlet/outlet caps, exact vertical side walls, and one marker
per corner.  Mobile row clearance includes particle radius, wall-contact
offset, maximum configured position jitter, and a small numerical epsilon.

\subsection{ForceDynamics Invariance}

\texttt{SimpleReservoir} does not modify the ForceDynamics equations:

\[
F_{ij}=q_{ij}M_{ij}
\]

and

\[
\mathbf v_{n+1}
=
\mathbf v_n+\frac{\mathbf F_n}{m}\Delta t.
\]

\(M_{ij}\) is selected by \texttt{contact\_force\_measure}.  The current
default is penetration depth; overlap area remains available as the optional
\texttt{area} mode.  \texttt{SimpleReservoir} only controls particle
eligibility and boundary interpretation.


\subsection{Unborn Particle Rule}

Reservoir particles may be stored in repeated columns of length
\texttt{col\_length}.  Many unborn particles may therefore share the same
stored position.  These stored positions are not physical particle positions
until the particle is born.

For \texttt{SimpleReservoir}, if:

\[
\texttt{Data.w > frameNum}
\]

then the particle is unborn.  An unborn particle is ignored by all
ForceDynamics operations:

\begin{itemize}
    \item no particle-particle contact check;
    \item no wall contact check;
    \item no tunneling check;
    \item no starting-contact check;
    \item no force accumulation;
    \item no velocity update;
    \item no position update;
    \item no energy or momentum diagnostics.
\end{itemize}

This prevents staged reservoir particles that share the same stored location
from producing artificial overlap, tunneling, or starting-collision errors.

\subsection{Birth Frame Rule}

When:

\[
\texttt{Data.w <= frameNum}
\]

and \(\texttt{Data.w > 0}\), the particle becomes live.  The birth process
sets:

\[
\texttt{Data.w = 0}
\]

and assigns the particle's active inlet position and velocity before the
ForceDynamics frame snapshot is built.

The corresponding GLSL source invocation order is:

\begin{verbatim}
if (!StartReservoir(SourceID))
    return;

// particle contact scan
// wall contact scan
CalcVelocity(SourceID, totalForce);
CalcPosition(SourceID);
RetireParticlePastXMax(SourceID);
\end{verbatim}

This preserves source ownership.  The reservoir state gates only the current
source invocation; it does not write target particles and does not introduce a
separate global reservoir pass.

The generated \texttt{PERIODIC\_DIRECTION} selector controls whether the inlet
and outlet are interpreted along \(x\) or \(y\).  When
\texttt{PERIODIC\_DIRECTION} is \texttt{vertical}, birth preserves the
generated \(x\) coordinate and offsets the \(y\) coordinate from
\(y_{\min}\).  Retirement then checks \(y_{\max}\).  When the selector is not
vertical, the historical horizontal \(x\)-axis behavior is used.

\subsection{Planned Deferred Birth Admission}

When a scheduled birth frame arrives, the particle will become a birth
candidate rather than becoming live immediately.  The reserved lifecycle
states are:

\begin{verbatim}
colFlg = 2  waiting for a birth attempt
colFlg = 3  testing a birth attempt
colFlg = 0  admitted live particle
\end{verbatim}

The candidate birth position is tested only against already-live mobile
particles.  Dead particles, unborn particles, simultaneous birth candidates,
and boundary markers are excluded.  If the motion contact query reports a
starting collision, the source search terminates immediately, the particle
returns to \texttt{colFlg=2}, and \texttt{Data.w} is changed to
\texttt{frameNum+1}.  It receives no force or motion update in that frame.

If no live target overlaps the candidate position, the particle is admitted by
setting \texttt{Data.w=0} and \texttt{colFlg=0}.  Delaying one particle does
not move the scheduled birth frames of later particles, so the generated wave
spacing is preserved.

The GPU implementation must test target liveness from immutable frame-start
state.  Birth testing and activation cannot depend on the order in which source
invocations happen to execute; separate test and commit phases, or an
equivalent synchronized design, are required.


\subsection{Birth Placement Rule}

When an unborn particle becomes live, its stored reservoir position is not used
as its active physical position.  The particle is moved to the inlet birth
position before ForceDynamics contact detection.

The birth position is offset from the reservoir/staging column by at least

\[
1.1(2r)
\]

where \(r\) is the particle radius.  Equivalently, the minimum birth offset is

\[
2.2r.
\]

This places the newly live particle slightly more than one particle diameter
away from particles that remain in the reservoir column.  The factor \(1.1\)
adds a small clearance margin so the particle is not born in contact due to
rounding or exact-diameter placement.

\texttt{reservoir\_birth\_spacing\_factor} may request a larger offset, but
the runtime clamps the factor to a minimum of \(1.1\).  For a closed
boundary-particle reservoir, the inlet wall is placed at
\texttt{reservoir\_inlet\_x}; positive-x birth therefore occurs at least
\(1.1(2r)\) inside that wall.

After birth:

\[
\texttt{Data.w}=0
\]

and both position buffers are set to the active birth position.

\subsection{Death Rule}

After position update, every live particle is tested against the outlet
boundary using its center position.  If the particle center has passed the
outlet boundary, then:

\[
\texttt{Data.w}=-1.
\]

A particle with \(\texttt{Data.w}<0\) is dead.  Dead particles are no longer
processed by ForceDynamics and are not checked for collision, wall contact,
tunneling, starting overlap, force, motion, or ordinary energy/momentum
diagnostics.

\subsection{Vertex Lifecycle Rule}

The graphics vertex stage must apply the same lifecycle rule before building
particle cell occupancy or drawing particle markers.  A vertex invocation for
an unborn or dead reservoir particle must not write cell-array occupancy and
must not leave vertex outputs undefined.

The shared GLSL vertex helper is:

\begin{verbatim}
ParticleLifeActive(SourceID, frameNum)
\end{verbatim}

For inactive particles, the vertex shader must write a defined offscreen
position before returning:

\begin{verbatim}
if (!ParticleLifeActive(index, uint(ShaderFlags.frameNum))) {
    gl_Position = vec4(2.0, 2.0, 2.0, 1.0);
    gl_PointSize = 0.0;
    fragColor = vec3(0.0);
    return;
}
\end{verbatim}

A bare vertex-stage \texttt{return} is not sufficient because it prevents
occupancy processing but can leave \texttt{gl\_Position} and color outputs
undefined.  \texttt{discard} is not available in vertex shaders, so moving the
vertex outside clip space is the required suppression method.

\subsection{CD Nozzle Reservoir Convention}

\texttt{BoundaryCDNozzleReservoir} is the reservoir generator for
converging-diverging nozzle studies.  The nozzle is described by a piecewise
radius profile with independently configured inlet, converging, throat,
diverging, and exit lengths.  For \texttt{dim=2}, the profile is generated in
the \(xy\) plane at a fixed \texttt{nozzle\_profile\_z}, which defaults to
\(2.0\).  The \(x\) axis is the default axial flow direction.

The generator writes mobile particles first and boundary marker particles
after them.  Mobile particles have \texttt{ptype=0.0}.  The inlet cap uses
\texttt{ptype=2.0} for the vertical-wall evaluator, and curved nozzle markers
use \texttt{ptype=3.0}.  The generated \texttt{dispatchx} covers only the null
particle plus mobile source particles.  The generated \texttt{num\_particles}
includes mobile and boundary particles.

The nozzle entrance must be anchored to the reservoir inlet plane.  For
horizontal flow, \texttt{nozzle\_start\_x} defaults to \(1.0\).  For vertical
flow, \texttt{nozzle\_start\_y} defaults to \(1.0\).  These start-plane
values are also the default reservoir inlet coordinates and the process
minimum written to the generated test file.  The process maximum is the start
plane plus the nozzle axial length.  These process bounds are lifecycle
planes; they are not the physical curved wall.

The generated CD nozzle OBJ is a visualization aid:
\begin{verbatim}
vulkan/sim/BoundaryCDNozzleReservoir/CDNozzleGenerated.obj
\end{verbatim}
It is generated from the same profile as the boundary particles, but it does
not define dynamics.  Contact response remains owned by boundary marker
particles and the selected wall-segment evaluator.

The current two-dimensional CD nozzle wall function is selected with:
\begin{verbatim}
boundary_particle_function = "cd_nozzle_wall";
\end{verbatim}
For horizontal CD nozzle flow, boundary particles are locality markers.  The
evaluator calculates the actual wall point and wall normal from the analytic
nozzle profile at the source particle's current \(x\) coordinate.  The wall
profile uses the same inlet, converging, throat, diverging, and exit lengths
and radii as the generator.  The source-owned force accumulation rule does not
change; only the wall-segment geometry evaluator changes.

The current Python 2D analytic profile uses straight inlet, throat, and exit
sections with throat-centered quadratic converging and diverging radius
sections.  For a transition of axial length \(L\), endpoint radius \(r_e\),
throat radius \(r_t\), and throat-centered coordinate \(\xi\), the implemented
form is

\begin{equation}
r(\xi)=r_t+(r_e-r_t)\left(\frac{\xi}{L}\right)^2,
\qquad
\frac{dr}{dx}=\frac{2(r_e-r_t)\xi}{L^2}.
\end{equation}

The converging side uses a signed coordinate whose derivative is negative; the
diverging side uses a positive coordinate.  Radius is continuous at every
section boundary and the slope is zero through the throat.  The generated GLSL
radius and slope templates use the same equations and the CD shader set
compiles with the quadratic profile.

The 2D inlet is closed by a vertical-evaluator marker column at
\texttt{nozzle\_start\_x}, spanning the inlet lower and upper radii.  Its two
corner coordinates are shared with the nozzle profile and stored only once.
The outlet remains open.  Reservoir birth places a mobile center at least
\(1.1\) diameters inside the inlet cap before dynamics processing.
