% !TeX root = main.tex
\newpage
\section{ForceDynamics Source Flow}

\newcommand{\ForceFunctionSection}[1]{%
  \subsection{ForceDynamics:#1}%
  \label{force:#1}%
}

\subsection{High-Level Model}

\texttt{ForceDynamics} is a source-owned, current-state contact model.  Each
mobile source reads immutable frame-start state, evaluates every contact that
applies to that source, sums the contact forces, and writes only its own
velocity and position.  A source never writes a target particle.  No global
pair solver reaches backward to revise a completed collision.

The frame has four conceptual stages:
\begin{enumerate}
    \item snapshot immutable frame-start particle state;
    \item detect particle and wall contacts and build one source-local force sum;
    \item calculate a candidate source velocity, then apply the current-state
          maximum-compression constraint to the source's complete contact set;
    \item integrate the source position and record diagnostics.
\end{enumerate}

Python calculates penetration depth and overlap area.  The default force
measure is penetration depth:
\begin{verbatim}
contact_force_measure = "depth";
\end{verbatim}
The alternate supported mode is:
\begin{verbatim}
contact_force_measure = "area";
\end{verbatim}
The source-owned schedule is exported to one GLSL compute invocation per
source particle.  The horizontal boundary-particle reservoir now runs the
generated force, maximum-depth, internal-momentum, and position paths in
Vulkan.  Position and velocity are both double buffered so every invocation
reads immutable frame-start state and writes only the alternate source state.

\texttt{ForceDynamics} has one stationary-wall representation: finite cubic
parametric segments in \texttt{curve\_wall\_segments}.  Straight lines are
represented by zero quadratic and cubic coefficients.  Boundary particles
with evaluator ID 5 are occupancy/locality markers; they are not mobile
dynamics sources and do not own separate horizontal, vertical, linear, or CD
wall equations.

When any evaluator-5 marker is local to a mobile source, Python evaluates the
configured parametric segments directly.  It retains the deepest valid
contact for each positive physical \texttt{wall\_flag}.  Marker target IDs do
not identify physical walls, which removes marker-order and shared-corner
ambiguity.

\clearpage

\ForceFunctionSection{CollisionRun}

\texttt{CollisionRun} is the Python harness for the shader-shaped execution
schedule:

\begin{verbatim}
snapshot frame-start positions and velocities
initialize starting-contact effective geometry and diagnostics
initialize totalForce and source contact sets
for each SourceID:
    ProcessPistonCollision(SourceID, totalForce)
    scan possible targets
        if TargetID is a mobile particle:
            ProcessParticleCollision(TargetID, SourceID, totalForce)
        if TargetID is a local evaluator-5 boundary marker:
            remember that parametric wall evaluation is local
    evaluate every curve_wall_segment once
    retain deepest contact per wall_flag
    ProcessParametricWallCollision(SourceID, contact, totalForce)
capture frame-start pair diagnostics
for each SourceID:
    calculate force-driven candidate velocity
    ApplySourceMaximumDepth(SourceID)
for each SourceID:
    CalcPosition(SourceID)
snapshot end positions for diagnostics
\end{verbatim}

The Python harness uses a direct particle scan where GLSL uses cell occupancy
lists.  Both schedules complete one source-local force sum before integration.
Maximum compression is evaluated from observed frame state, not from a
predicted future position.
Starting-contact state is initialized after the frame-start snapshot so
particle-particle and particle-wall overlaps present in the initial
configuration can be treated as zero-force contact limits.  The end-position
snapshot is diagnostic only and is not a second force scan.

The Python harness uses a direct scan over particles where GLSL uses cell
occupancy lists.  Boundary markers gate locality only.  Actual wall points,
tangents, normals, penetration, and force come from the configured parametric
segments.

The horizontal prefilled-reservoir scenario also evaluates one analytic
piston contact per mobile source before scanning particle and boundary-marker
targets.  The piston does not depend on marker locality and does not duplicate
the stationary boundary-marker wall path.

\ForceFunctionSection{GetPistonPosition}

\texttt{GetPistonPosition(frame)} returns the analytic piston plane position.
Before \texttt{piston\_start\_frame} it remains at the reservoir start.  It
then advances by \texttt{reservoir\_velocity.x} and the frame timestep until
it reaches the reservoir end, where it is clamped.

\ForceFunctionSection{GetPistonVelocity}

\texttt{GetPistonVelocity(frame)} returns zero before motion and after the
stop plane is reached.  During travel it returns the configured reservoir
velocity.  This is the target-wall velocity used by relative-normal motion,
the penetration-step guard, internal-momentum compression and release, and
maximum-depth containment.

\ForceFunctionSection{EvaluatePistonWall}

\texttt{EvaluatePistonWall(SourceID)} constructs a vertical equal-radius
ghost plane at the current analytic piston position.  It applies across the
reservoir transverse bounds and returns ordinary wall-contact geometry.

\ForceFunctionSection{ProcessPistonCollision}

\texttt{ProcessPistonCollision(SourceID,totalForce)} evaluates the piston once
for each active mobile source, records its moving target velocity, performs
wall penetration and maximum-depth checks, appends a source-owned wall
contact, and accumulates force.  No piston boundary-marker row is emitted.

\ForceFunctionSection{InitializeStartingContactState}

\texttt{InitializeStartingContactState()} records mobile particle contacts
that already overlap when the model starts.  Boundary marker particles are
not recorded as mobile starting contacts.  For each starting contact, the
current center distance becomes the effective separation limit.
The physical radii are still retained, but force and potential energy use the
effective geometry until the contact rebounds to its starting separation and
then clears the physical overlap.

\ForceFunctionSection{ProcessParticleCollision}

\texttt{ProcessParticleCollision(TargetID, SourceID, totalForce)} initializes
the current contact geometry, records a diagnostic contact slot, and
immediately adds the contact force to the source-local \texttt{totalForce}.
It does not write the target particle.  Boundary markers are not processed by
this function; they are routed to the boundary-particle wall path.

\ForceFunctionSection{ProcessParametricWallCollision}

\texttt{ProcessParametricWallCollision(SourceID,contact,totalForce)} accepts
already evaluated transient wall geometry, appends the diagnostic wall slot,
and immediately accumulates force into the source-local sum.

\ForceFunctionSection{EvaluateParametricWallSegment}

\texttt{EvaluateParametricWallSegment(SourceID,segment)} is the sole
stationary-wall evaluator.
Each finite segment stores cubic coefficients for \(x(t)\) and \(y(t)\), with
\(0\leq t\leq1\), plus its physical wall flag.  Boundary markers provide only
broad-phase locality.  The evaluator selects the applicable segment, performs
a bounded closest-point search, evaluates the local tangent, and derives the
outward right-hand normal from segment direction.  Straight, quadratic, and
cubic sections therefore share one wall-contact implementation.

Python and the Vulkan wrappers may encounter several marker targets
representing the same physical wall side.  They evaluate marker geometry and
retain the deepest valid candidate for each wall flag before force is applied.
This removes marker scan-order dependence at segment transitions.  Ordinary
mobile-particle duplicates remain keyed by target ID.

\ForceFunctionSection{ParametricBoundaryMarkerApplies}

\texttt{ParametricBoundaryMarkerApplies(SourceID,BoundaryID)} is the Python
locality guard.  It accepts only evaluator-5 markers within one cell of the
source in each axis.  The GLSL path obtains the equivalent locality from cell
occupancy.

\ForceFunctionSection{IsBoundaryParticle}

\texttt{IsBoundaryParticle(ParticleID)} returns true when the particle record
has positive evaluator-valued \texttt{ptype}.  Boundary particles are
occupancy and wall-geometry markers, not mobile dynamics sources.

\ForceFunctionSection{IsMobileParticleActiveForDynamics}

\texttt{IsMobileParticleActiveForDynamics(ParticleID)} excludes boundary
markers from source integration.  Scheduled reservoir lifecycle filtering has
been removed; mobile particles are active from their generated initial state.

\ForceFunctionSection{InitializeContactState}

\texttt{InitializeContactState(SourceID, TargetID)} obtains the current
particle-particle geometry and records one source-owned diagnostic contact
slot.

\ForceFunctionSection{GetContactState}

\texttt{GetContactState(SourceID, TargetID)} obtains a reusable current-frame
contact slot.  The slot contains geometry and diagnostic values only; it is
not used later to calculate force.

\ForceFunctionSection{AppendContactSlot}

\texttt{AppendContactSlot(SourceID, TargetID)} records an active
particle-contact slot owned by the source.

\ForceFunctionSection{GetContactSlots}

\texttt{GetContactSlots(SourceID)} returns the fixed diagnostic contact-slot
array owned by the source particle.

\ForceFunctionSection{GetParticleGeometry}

\texttt{GetParticleGeometry(SourceID, TargetID)} calculates overlap area,
center distance, penetration depth, and the source-to-target normal from the
immutable frame-start position snapshot.  In depth mode, the source orientation
vector points from the source center to the source boundary in the collision
direction, the proximity vector points from the source center to the target
leading edge, and penetration depth is the orientation magnitude minus the
proximity magnitude.  Ordinary contacts use the physical particle radii.  If
the pair belongs to a starting contact,
\texttt{GetParticleGeometry} obtains effective radii from
\texttt{GetParticleEffectiveContactGeometry}.  The effective radii make the
starting separation a zero-force limit.  If a resolved starting contact is
still physically overlapping, physical force is suppressed until the actual
radii separate.

\ForceFunctionSection{GetParticleEffectiveContactGeometry}

\texttt{GetParticleEffectiveContactGeometry(SourceID, TargetID,
center\_distance)} selects the radii used by both force and diagnostic
potential energy for one particle contact.  It returns physical radii for
ordinary contacts, effective radii for active starting contacts, and a
suppression flag for resolved starting contacts that have not yet cleared the
physical-radius overlap.

\ForceFunctionSection{GetParticlePosition}

\texttt{GetParticlePosition(ParticleID)} returns a particle position from the
selected immutable position snapshot.

\ForceFunctionSection{MaximumCompression}

Maximum compression applies to both particle and wall contacts.  With source
radius \(r_s\), the first-version rule is:
\begin{equation}
\delta_{\max}=0.5r_s.
\end{equation}
At or beyond this observed depth, further inward source-normal motion is
removed.  Tangential velocity is preserved.  The removed source translational
momentum is stored in the source-owned internal-momentum vector
\texttt{parms.yzw}.

All maximum-depth contacts for one source are collected before the source is
updated.  \texttt{ProjectSourceVelocityToContactSet} deterministically projects
the candidate source velocity against that complete local constraint set.  It
does not mutate targets and does not iterate over global particle pairs.

The model also enforces a timestep-resolution condition:
\begin{equation}
\text{relative inward displacement per step}
\leq (1-0.5)r_s.
\end{equation}
A violation raises \texttt{ERROR\_PENETRATION\_STEP\_TOO\_LARGE}; the simulation
does not silently alter \(\Delta t\).

For a moving piston, inward displacement and maximum-depth containment are
relative to piston velocity.  At maximum depth the source normal velocity is
limited to the piston normal velocity rather than to zero.  Removed relative
inward source momentum remains source-owned internal momentum.

Error reporting distinguishes \texttt{ERROR\_PARTICLE\_TUNNELING},
\texttt{ERROR\_WALL\_TUNNELING}, \texttt{ERROR\_MAX\_DEPTH\_CONSTRAINT}, and
\texttt{ERROR\_PENETRATION\_STEP\_TOO\_LARGE}.  Reported source and target IDs
are particle numbers because Python preserves the null particle at index zero.

\ForceFunctionSection{VulkanBufferedState}

Vulkan uses the same selector for position and velocity buffers.  When
\texttt{positionBuffer=0}, contact processing reads \texttt{PosLocA} and
\texttt{VelRadA}, while integration writes \texttt{PosLocB} and
\texttt{VelRadB}.  The roles reverse when the selector is one.
\texttt{GetParticleVelocity} reads the immutable current velocity,
\texttt{GetNextParticleVelocity} reads the candidate/output velocity, and
\texttt{SetNextParticleVelocity} writes only that alternate buffer.

After \texttt{CalcPosition}, \texttt{ApplyParticleDeathBounds} tests the newly
written position against the six configured death planes.  A mobile particle
outside any plane receives \texttt{Data.w=-1}.  Dead particles are excluded as
compute sources, contact targets, cell-occupancy entries, and rendered
particles.  Boundary markers are exempt from death-bound processing.

\ForceFunctionSection{GetStartFrameVelocity}

\texttt{GetStartFrameVelocity(ParticleID)} returns the immutable frame-start
velocity used by integration and relative-normal-velocity diagnostics.

\ForceFunctionSection{AccumulateContactForce}

For every current particle or wall contact,
\texttt{AccumulateContactForce(SourceID, contact\_state, totalForce)} applies:

\begin{equation}
F_{ij}=q_{ij}M_{ij},
\end{equation}

\begin{equation}
\mathbf F_i \mathrel{+}= -F_{ij}\mathbf n_{ij}.
\end{equation}

\(M_{ij}\) is selected by \texttt{contact\_force\_measure}.  In the default
\texttt{depth} mode, \(M_{ij}\) is penetration depth.  With
\(\mathbf O_i\) as the source orientation vector and \(\mathbf P_{ij}\) as the
source-to-target-leading-edge proximity vector:
\begin{equation}
M_{ij}=\delta_{ij}=\lVert\mathbf O_i\rVert-\lVert\mathbf P_{ij}\rVert,
\end{equation}
which reduces to:
\begin{equation}
M_{ij}=r_i+r_j-\lVert \mathbf x_j-\mathbf x_i\rVert.
\end{equation}
For wall contacts the equal-radius wall ghost gives
\(M_{ij}=2r_i-d_{\mathrm{wall}}\).  In \texttt{area} mode, \(M_{ij}\) is the
circle-overlap area \(A_{ij}\), preserving the earlier force law.  Python
continues to calculate overlap area in both modes for diagnostics and for
area-mode comparison.

The force is applied during both compression and rebound.  Starting contacts
use effective geometry measured from their starting separation.  There is no
contact weighting, available-momentum cap, or collision phase.  Maximum-depth
containment is a separate source-velocity constraint applied after the
force-driven candidate velocity is calculated.

\ForceFunctionSection{RecoverableInternalMomentum}

Each source particle owns a recoverable internal momentum vector in
\texttt{parms.yzw}.  \texttt{parms.x} remains particle mass.  During
compression, \texttt{AccumulateContactForce} adds the current contact impulse
along the contact normal to this vector.  During rebound, it releases the
stored component along the same normal, limited by the current contact
impulse.  The scalar magnitude is mirrored to \texttt{Data.z} and to the
Python \texttt{internal\_momentum} report field.

Force-contact bookkeeping records compression and rebound.  In addition,
maximum-depth containment stores inward source momentum that cannot be
expressed as translational motion during the current frame.  The storage is
therefore no longer purely diagnostic, although a complete release and thermal
transfer model is still deferred.  Heat, thermalization, and heat-transfer
coefficients remain later model layers.

Piston contacts use piston velocity when determining the sign of relative
normal motion.  They therefore store momentum while the piston closes on a
source and release it when source and piston separate, rather than treating
the piston as a stationary wall.

\ForceFunctionSection{PistonDiagnostics}

Python records current piston position and velocity, current-frame contact
count, frame and cumulative scalar impulse, and cumulative vector momentum
transferred from the piston to particles.  The game UI displays this state and
\texttt{momentum.csv} carries the same diagnostic columns.

\ForceFunctionSection{GetContactStiffness}

\texttt{GetContactStiffness(SourceID, TargetID, contact\_type)} selects
particle-pair or wall stiffness for the current contact.

\ForceFunctionSection{GetPairStiffness}

\texttt{GetPairStiffness(SourceID, TargetID)} returns the nonnegative mean of
the two particle-owned stiffness values:

\begin{equation}
q_{ij}=\max\left(0,\frac{q_i+q_j}{2}\right).
\end{equation}

The units of \(q\) depend on \texttt{contact\_force\_measure}.  In
\texttt{depth} mode, \(q\) is depth stiffness.  In \texttt{area} mode, \(q\)
is area stiffness.

\ForceFunctionSection{CalcVelocity}

\texttt{CalcVelocity(SourceID, totalForce)} first calculates the force-driven
candidate source velocity:

\begin{equation}
\mathbf v_{n+1}
=
\mathbf v_n+\frac{\mathbf F_n}{m}\Delta t.
\end{equation}
It then calls \texttt{ApplySourceMaximumDepth(SourceID)} before the velocity is
accepted.  This second step may remove inward normal velocity and store the
corresponding source-owned internal momentum.  Position integration sees only
the accepted velocity.

\ForceFunctionSection{GetParticleMass}

\texttt{GetParticleMass(ParticleID)} returns particle mass from
\texttt{parms.x} with an epsilon lower bound.

\ForceFunctionSection{CalcPosition}

\texttt{CalcPosition(SourceID)} performs the semi-implicit Euler position
write using the newly calculated velocity:

\begin{equation}
\mathbf x_{n+1}
=
\mathbf x_n+\mathbf v_{n+1}\Delta t.
\end{equation}

This velocity-first update is the symplectic Euler method.  Symplectic Euler
does not conserve the system's exact total energy at every timestep.  Instead,
it typically produces a bounded energy excursion in which the calculated
total energy oscillates above and below the exact value.  Unlike standard
Euler integration, this error generally does not accumulate into continuous
energy growth or decay.  Decreasing the timestep reduces the excursion's
magnitude, making symplectic Euler well suited for long-running conservative
dynamics simulations.

\subsection{Potential Energy and Diagnostics}

In depth mode the pair potential is:
\begin{equation}
U_{ij}(d)=\frac{1}{2}q_{ij}\delta_{ij}^{2}.
\end{equation}
In area mode the earlier overlap-area potential remains
\(U_{ij}(d)=q_{ij}\int_d^{r_i+r_j}A_{ij}(s)\,ds\).

For a starting contact, \(r_i+r_j\) is replaced by the effective starting
separation limit while the contact is active.  After the contact rebounds to
that limit, force and potential energy remain suppressed until the actual
physical radii separate.  This avoids a second rebound caused only by the
initial configured overlap.

Potential energy, relative normal velocity, contact slots, end-position
snapshots, and pair-phase measurements are diagnostic only.  They do not
affect the source-local force sum or particle integration.  System energy is:

\begin{equation}
E_{\mathrm{total}}=KE+\sum_{i<j}U_{ij}+\sum_{i,w}U_{iw}.
\end{equation}

\subsection{Current Numerical Results}

The central pair rule independently produces equal-and-opposite particle-pair
forces and preserves vector momentum.  Semi-implicit Euler introduces a small
bounded numerical energy excursion.  Current captures include:

\begin{verbatim}
ThreeParticle, 500 steps:
energy range = approximately [-9.18e-5, +9.98e-5]
final energy difference = approximately -1.69e-7

ParticlesInBox, 10000 steps:
energy range = approximately [-1.85e-5, +1.90e-5]
final energy difference = approximately -7.34e-7
\end{verbatim}

The long-running box result currently rises and falls within a bounded band
rather than showing monotonic energy drift.
