
% !TeX root = GeoDynamics.tex
\section{Geo Dynamics Conventions}

\subsection{Purpose}

The Geo path is a clean restart of the collision dynamics code.  Its purpose is
to build a Python implementation whose structure can be transferred to GLSL
with minimal translation.

\subsection{File Boundaries}

\texttt{GeoDynamics.py} is the GLSL-portable dynamics layer.  It should contain
only functions that correspond directly to shader-side dynamics functions.

\texttt{GeoBase.py} is the Python harness.  It may load configuration files,
create Python data structures, build particle arrays, set push constants,
iterate particles, and call the Geo dynamics functions.

\texttt{Reporting.py} is reserved for reporting and diagnostic output.  It
should not own dynamics decisions.

\subsection{Naming Convention}

Functions intended to map to GLSL should use GLSL-like names and signatures
where practical.  Current examples are:

\begin{verbatim}
GeoAddParticleContact(SourceID, TargetID)
GeoProcessCollisions(SourceID)
GeoMoveParticle(SourceID, positionBuffer, dt)
isParticleContact(Frame, SourceID, TargetID, positionBuffer)
\end{verbatim}

\texttt{GeoDynamics.py} should avoid Python-only helper names unless the helper
is also expected to exist in GLSL.

Before adding or renaming any Geo field, buffer, function, or constant, check
the GLSL source files first.  The GLSL names are the naming authority for the
Geo path:

\begin{verbatim}
C:\_DJ\gPCD\vulkan\sim\common\*.glsl
C:\_DJ\gPCD\vulkan\sim\BoundarySpheresMotion\BoundarySpheresMotionP.comp
\end{verbatim}

Python names should stay coordinated with those files whenever possible.  If a
Python-only harness name is needed, it belongs in \texttt{GeoBase.py},
\texttt{GeoRunner.py}, or \texttt{Reporting.py}, not in
\texttt{GeoDynamics.py}.

\subsection{Data Convention}

Particle records use field access:

\begin{verbatim}
Particle.vx
Particle.radius
\end{verbatim}

rather than dictionary access:

\begin{verbatim}
Particle["vx"]
\end{verbatim}

The Python structures should be shaped to resemble GLSL buffers and push
constants.  Python-only configuration parsing must not leak into
\texttt{GeoDynamics.py}.

When adding particle position, velocity, radius, collision-list, push-constant,
or error fields, first look for the corresponding GLSL structure or function
name in the common shader files.  For example, position double buffering should
prefer names coordinated with the GLSL particle buffers and
\texttt{ShaderFlags.positionBuffer}, rather than inventing unrelated Python-only
names.

\subsection{Overlap Area Convention}

Overlap area is a geometric contact property.  For a particle-particle contact,
the source-side and target-side overlap areas are the same geometric area, even
when the two particles have different velocities.  Therefore, once the overlap
area is computed for a source-target pair, the target particle's overlap area is
known from the same calculation.

Velocity may affect momentum transfer, phase, or future position, but it does
not make the instantaneous geometric overlap area different for the two
particles in the same contact.

\subsection{Instantaneous Geometric Goal}

The long-term Geo dynamics target is an instantaneous, purely geometric
collision model.

Instantaneous means the collision response for a frame is computed from the
current frame's particle state only:

\begin{itemize}
    \item current positions;
    \item current velocities;
    \item current radii and masses;
    \item current overlap geometry;
    \item current source-owned contact list for this frame.
\end{itemize}

Purely geometric means the response should be derived from contact geometry,
such as center distance, normal direction, penetration depth, overlap area, and
the distribution of overlap area across a source particle's simultaneous
contacts.  Energy calculations and energy-named dynamics variables should be
reserved for diagnostics only.

The model should not require persistent collision-episode memory except for
explicitly approved exceptions such as the current contact-owned internal
momentum ledger.  Unapproved examples include:

\begin{itemize}
    \item stored contact momentum reservoirs;
    \item immutable first-contact velocity;
    \item persistent compression/rebound phase state;
    \item remembered alpha-zero values;
    \item global wall or contact reservoirs shared by particles.
\end{itemize}

The only persistent particle state should be ordinary particle state, such as
position, velocity, radius, mass, and explicitly approved per-particle fields.
The current frame may build temporary contact information, but that temporary
contact information should be safe to rebuild on the next frame.

\subsection{Approved Contact-Owned Internal Momentum Exception}

The Geo path may carry contact-owned internal normal momentum ledgers under
the source particle.  This is an explicit exception to the fully stateless
geometric target because the current overlap-area impulse needs a ledger to
avoid releasing more momentum than it removed during compression.

The exception is constrained:

\begin{itemize}[leftmargin=*]
    \item each value is owned by the source particle and keyed by the current
    target particle;
    \item it is not a shared pair reservoir and does not allow target writes;
    \item it is not owned by the indexed contact slot;
    \item stale target keys must be pruned when the contact is no longer active;
    \item it must not write the target particle;
    \item it may be used only to cap source-owned rebound/release impulse;
    \item the source particle may report the sum of active ledgers as total
    internal momentum.
\end{itemize}

In the current Python layout, the contact ledgers are:

\begin{verbatim}
P[SourceID].contact_internal_momentum[TargetID]
P[SourceID].contact_internal_phase[TargetID]
\end{verbatim}

The Python diagnostics may also remember a source-owned compression direction
for the same target key:

\begin{verbatim}
P[SourceID].contact_compression_direction[TargetID]
\end{verbatim}

This value is diagnostic-only in the current model.  It may be used to compare
stored compression direction against rebound direction, but it must not change
the applied impulse unless a later project update explicitly promotes a new
dynamics rule.

The GLSL-shaped particle field:

\begin{verbatim}
P[SourceID].Data.z
\end{verbatim}

reports the source particle's total active internal normal momentum, computed
as the sum of its contact ledgers.  The current Python model has therefore
moved ahead of the GLSL storage shape; GLSL export must not be regenerated for
this change until an equivalent shader-side representation is designed.

\subsection{Contact Slot Convention}

Contact slots are current-frame indexed records.  When a contact is discovered,
the next slot is the source particle's current contact count:

\begin{verbatim}
slot = P[SourceID].contactCount
P[SourceID].contactCount += 1
\end{verbatim}

The slot records the current frame's source, target, normal, overlap area,
phase report, and diagnostics.  Slot index is not contact identity and must
not own persistent dynamics memory.  Persistent contact memory, when allowed
by the exception above, belongs under the source particle and is keyed by
target identity, not by slot index.

Slot order must not decide which contact receives internal momentum.  Any
frame-local contact response that uses persistent ledgers must first build the
current contact set and then plan the unordered-pair impulse from snapshot
data before source-owned writes are applied.

\subsection{Frame-Local Closing Momentum Convention}

A closing contact may compute frame-local reduced-mass closing momentum:

\begin{equation}
P_{close} =
\mu \max(0,-v_n),
\qquad
\mu = \frac{m_s m_t}{m_s + m_t}.
\end{equation}

This is allowed because it is computed from the frame-start velocity snapshot,
current contact normal, and particle masses.  It is not a persistent
\texttt{p\_zero} value, not first-contact memory, and not a pair reservoir.
It may be used as a source-owned compression availability value, with the same
current-frame area weighting rules used for multi-contact geometry.

\subsection{Model-Preservation Warning}

Before editing code, any proposed change that would move the Geo path away from
the instantaneous, purely geometric target must be called out explicitly.

Examples of changes that require warning before implementation:

\begin{itemize}
    \item adding persistent per-contact memory for dynamics;
    \item adding a reservoir whose value survives after the frame;
    \item using first-contact velocity as a long-lived dynamics input;
    \item using phase history to decide future dynamics;
    \item adding global writable memory shared by all particles;
    \item adding a wall-specific dynamics path instead of expressing the wall as
    contact geometry.
\end{itemize}

This warning does not forbid such changes forever, but it makes the model cost
visible before implementation.  The current internal momentum exception is
contact-owned memory stored under the source particle; it remains source-owned
for writes, but it is no longer a single source-total dynamics ledger.

\subsection{Control Flow Shape}

The Python control-flow harness should match the compute shader shape:

\begin{verbatim}
for each SourceID:
    build source-owned contact list

GeoPlanContactImpulses()

for each SourceID:
    GeoProcessCollisions(SourceID)

for each SourceID:
    GeoMoveParticle(SourceID, positionBuffer, dt)
\end{verbatim}

The current Python harness may use a direct all-pairs scan while the shader uses
cell/corner occupancy lookup.  That difference belongs in \texttt{GeoBase.py},
not \texttt{GeoDynamics.py}.

\subsection{No Pairwise Write-Back Shortcut}

The Geo Python implementation must not use a pairwise dynamics shortcut that
cannot be exported to the current GLSL execution model.

The active shader shape is source-owned: one invocation owns one source
particle.  Therefore a Geo dynamics function may write only the current source
particle's state.  It must not solve a pair by directly updating both
particles:

\begin{verbatim}
solve_pair(source, target):
    P[source].velocity = ...
    P[target].velocity = ...   // forbidden in GeoDynamics.py
\end{verbatim}

This rule applies even when a pairwise update would be physically cleaner in
Python.  Python is not allowed to become a separate reference model that fixes
collisions in a way Vulkan cannot reproduce.  If the shader cannot safely do
the write pattern, the Python Geo path cannot do it either.

Allowed alternatives must preserve source-owned execution.  Examples include:

\begin{itemize}[leftmargin=*]
    \item computing a source-owned response from source-owned contact data;
    \item using read-only frame-start snapshots of target state;
    \item accumulating source-owned diagnostic values;
    \item designing an explicit GLSL-compatible staged pass before adding the
    equivalent Python implementation.
\end{itemize}

Any proposal that needs one contact solve to write both particles must first
change the GLSL architecture, not just the Python model.

\subsection{Shadow Dynamics Convention}

Shadow dynamics are experimental dynamics, not applied live dynamics.  The
shadow path must run as a separate object from the live path:

\begin{verbatim}
GeoBase + GeoDynamics       -> live object
ShadowBase + ShadowDynamics -> shadow object
\end{verbatim}

The shadow object may write its own particle velocity, position, internal
momentum, contact phase, and contact ownership because those writes affect only
the shadow particle state.  It must not mutate live particles or live
contact ledgers.  Shadow reports must use the \texttt{*_s.csv} suffix so live
and shadow captures can be compared without sharing files.

The applied live dynamics remain authoritative.  A shadow result may be
promoted only after a project update records:

\begin{itemize}[leftmargin=*]
    \item what problem it solves;
    \item what tests support it;
    \item what convention changes are required;
    \item how the same rule can be represented in the GLSL source-owned
    execution model.
\end{itemize}

The \texttt{target\_elastic\_*} diagnostic remains a special case: it is a
durable reference compass rather than a proposed soft-response rule.  It may be
kept permanently to compare the current frame against an elastic hard-contact
target, but it should be reported through the shadow/reporting path rather
than embedded in live \texttt{GeoDynamics.py}.

Historical in-place shadow families such as \texttt{shadow\_*},
\texttt{vector\_shadow\_*}, \texttt{target\_ratio\_*},
\texttt{vector\_internal\_*}, and \texttt{parabolic\_*} have been removed from
the live dynamics path.  Future experiments belong in
\texttt{ShadowDynamics.py}.  The current shadow experiment uses a
velocity-reference progress variable instead of a geometric p-zero overlap
reference.

\subsection{Fixed P-Zero Shadow Convention}

A parabolic p-zero shadow may carry a contact-owned shadow reference:

\begin{verbatim}
particle.contact_parabolic_zero_reference[TargetID]
\end{verbatim}

The reference is set when the directed contact first has closing normal
momentum.  It stores the zero internal normal momentum and the corresponding
p-zero overlap area computed from the current-frame snapshot.  The shadow then
uses that fixed reference for the rest of the active contact instead of
recomputing the reference from the growing internal momentum ledger.

The current p-zero diagnostic target uses the fixed overlap fraction
\(f_k = \operatorname{clamp}(A_k/A_{0,k},0,1)\) and maps it through the direct
parabolic target:

\begin{equation}
I^{shadow}_{k} = P_{0,k}\left(1-\left(1-f_k\right)^2\right).
\end{equation}

The shadow impulse is the difference between this target and the stored shadow
internal momentum.  This memory is allowed because it belongs to the separate
shadow object.  It must be pruned when the contact is no longer active.

\subsection{Standalone Parabolic Response Implementation}

The next parabolic-response implementation belongs in
\texttt{ShadowDynamics.py}, not inside live \texttt{GeoDynamics.py}.  The live
and shadow simulations should be separate objects created from the same cfg
data:

\begin{verbatim}
GeoBase + GeoDynamics       -> live object
ShadowBase + ShadowDynamics -> shadow object
\end{verbatim}

The shadow object may write its own velocity, position, contact phase, and
internal momentum because those writes affect only the shadow particle state
and are reported to \texttt{*_s.csv} files.  The live object remains
authoritative for the UI and normal reports.

The parabolic response is a contact-owned target-ledger rule.  For each
directed contact \(k\), compute the current-frame normal velocity:

\begin{equation}
v_{n,k} = (v_{target}-v_{source})\cdot n_k.
\end{equation}

When a contact first has closing normal momentum, store the p-zero reference:

\begin{equation}
P_{0,k} = \mu_k \max(0,-v_{n,k}),
\end{equation}

and compute the corresponding p-zero overlap area \(A_{0,k}\) from the same
area/stiffness/\(\Delta t\) relation used by the geometry model.  This
reference is fixed for the active contact.  It must not be recomputed from the
current internal momentum ledger.

For each frame, compute:

\begin{equation}
f_k = \operatorname{clamp}(A_k/A_{0,k},0,1),
\qquad
I^{target}_{k} = P_{0,k}\left(1-\left(1-f_k\right)^2\right).
\end{equation}

The planned contact impulse is the ledger delta:

\begin{equation}
\Delta I_k = I^{target}_{k} - I^{current}_{k}.
\end{equation}

Positive \(\Delta I_k\) stores internal momentum and applies compression.
Negative \(\Delta I_k\) releases internal momentum and applies rebound.  The
velocity update still follows the source-owned rule:

\begin{equation}
\Delta p_{source,k} = -\Delta I_k n_k.
\end{equation}

The compression-to-rebound transition is phase-owned, not a separate geometry
branch.  A contact in \texttt{COMPRESSION} may flip to \texttt{RETURNING} when
its normal velocity reaches the p-zero tolerance:

\begin{equation}
|v_{n,k}| \leq \epsilon_v,
\qquad
\epsilon_v = (P_{0,k}/\mu_k)q_v.
\end{equation}

At that instant, the residual normal momentum is:

\begin{equation}
P^{residual}_{k} = \mu_k |v_{n,k}|.
\end{equation}

This residual is diagnostic until the exact ledger accounting is chosen.  The
important rule is that once a contact enters \texttt{RETURNING}, it remains
\texttt{RETURNING} until its contact-owned internal momentum drains or the
contact ends.  The rebound side must not request new compression merely because
the overlap fraction is still large.

The implementation must keep the functions narrow:

\begin{itemize}[leftmargin=*]
    \item build the shadow frame snapshot;
    \item build shadow contact lists;
    \item update p-zero references;
    \item compute parabolic target ledgers;
    \item compute source-owned impulse deltas;
    \item apply shadow velocity writes;
    \item move shadow particles;
    \item report to \texttt{*_s.csv}.
\end{itemize}

The old in-\texttt{GeoDynamics.py} shadow diagnostics are historical
scaffolding and have been removed from the live path.  The parabolic
velocity-progress experiment now lives in \texttt{ShadowDynamics.py} and is
reported through \texttt{contacts\_s.csv}.

\subsection{Velocity-Progress Shadow Response}

The current shadow implementation removes geometric p-zero from compression
control.  Each directed contact stores a contact-owned velocity reference:

\begin{verbatim}
particle.contact_velocity_reference[TargetID]
\end{verbatim}

The reference is set from the first closing normal speed:

\begin{equation}
v_{0,k} = \max(0,-v_{n,k}).
\end{equation}

The parabolic response then uses velocity progress rather than overlap
progress.  During compression, the progress uses a predicted post-impulse
closing speed so a perfectly head-on collision can start moving even before
the measured relative velocity has changed:

\begin{equation}
s_k = \max\left(0,\; v_{close,k} - J^{seed}_k/\mu_k\right).
\end{equation}

The progress and target internal momentum are:

\begin{equation}
f_k = 1-\operatorname{clamp}(s_k/v_{0,k},0,1),
\qquad
I^{target}_{k}
=
\mu_k v_{0,k}\left(1-\left(1-f_k\right)^2\right).
\end{equation}

The planned impulse remains the target-ledger delta:

\begin{equation}
\Delta I_k = I^{target}_{k}-I^{current}_{k}.
\end{equation}

Positive \(\Delta I_k\) stores internal momentum during compression.  Rebound
is not controlled by the compression target.  Once a contact is returning, the
release impulse is driven directly by the stored internal momentum and a
rebound profile.  The first rebound profile is:

\begin{equation}
J^{release}_k = \min(I^{current}_k,J^{raw}_k).
\end{equation}

If \(J^{raw}_k\) has fallen to zero while stored internal momentum remains, the
shadow drains the remaining stored momentum.  This prevents a contact from
keeping internal momentum after the contact geometry has ended.  The
compression-to-rebound flip is based on velocity magnitude:

\begin{equation}
v_{close,k} \leq 0.01v_{0,k}.
\end{equation}

This rule fixed the all-zero shadow dynamics seen in
\texttt{TwoParticleHorizontal}.  Direct stored-ledger release also removed the
long stored-momentum tail.  The current remaining issue is shape mismatch:
compression stores internal momentum through a velocity-progress parabola, but
the first direct rebound profile drains it through current raw overlap impulse.
The next rebound experiment should store the internal momentum at flip and use
a parabolic remaining-momentum target for rebound.

\subsection{Normal Velocity Target Shadow Direction}

Energy remains diagnostic-only.  The next shadow experiment must not use
kinetic energy as a dynamics input.  The latest angled two-particle result
shows that a parabolic scalar internal-momentum ledger can drain completely
while the final kinetic energy remains low.  The next target should therefore
be contact normal velocity or normal momentum, not total scalar internal
momentum.

For compression, the shadow should compute a target closing normal velocity
and apply only the normal impulse needed to move from the current closing
normal momentum toward that target.  For rebound, the shadow should compute a
target separating normal velocity and release only enough stored internal
momentum to move toward that normal-momentum target:

\begin{equation}
J_k =
p^{target}_{n,k} - p^{current}_{n,k}.
\end{equation}

The contact-owned internal momentum ledger still caps rebound.  Kinetic energy
is then used only to judge whether the velocity/momentum rule restored the
expected collision behavior.

\subsection{Dynamics Versus Harness}

\texttt{GeoDynamics.py} should not:

\begin{itemize}
    \item read cfg files;
    \item create particle lists from cfg files;
    \item know about UI flags;
    \item know about reports or capture files;
    \item contain test harness iteration logic.
\end{itemize}

\texttt{GeoBase.py} may do those things because it is not intended to be a
one-to-one GLSL export unit.

\subsection{Project Update Convention}

When the phrase \emph{update project} is used during Geo dynamics work, it
means the project record should be brought current before check-in.  The update
includes:

\begin{itemize}[leftmargin=*]
    \item generate a commit message for the current completed work;
    \item update \texttt{GeoConventions.tex} when the work changes a rule,
    boundary, workflow, or modeling convention;
    \item update \texttt{ParticleGeometry.tex} when the work changes equations,
    data fields, contact geometry, impulse logic, or the active algorithm;
    \item update \texttt{ProblemEvolution.tex} when the work resolves,
    exposes, or reframes a dynamics problem;
    \item update \texttt{ProgrammingStyle.tex} when the work changes the
    accepted control-flow shape, staging style, function-narrowing rule, or
    debugging/programming practice.
\end{itemize}

The documentation update should describe the current implementation and the
reason for the change.  Old material should be removed or clearly superseded
when it no longer describes the active Geo path.

\subsection{Current TODO}

\begin{itemize}
    \item Decide whether \texttt{isParticleContact(...)} belongs in
    \texttt{GeoBase.py} or should be moved to a GLSL-matching dynamics/helper
    module.
    \item Keep \texttt{GeoDynamics.py} one-to-one with future GLSL export.
    \item Keep live \texttt{GeoDynamics.py} clean while the next response rule
    is tested in \texttt{ShadowDynamics.py}.
    \item Continue the parabolic velocity-progress experiment in
    \texttt{ShadowDynamics.py}.  Compression now produces nonzero shadow
    dynamics and direct release drains the ledger; the next test is to impose
    the parabolic target on relative normal velocity/normal momentum rather
    than on scalar internal momentum.
    \item Treat wall/out-of-bounds behavior as a separate problem after the
    particle-particle multi-contact rule is stable.
    \item Add reporting only through \texttt{Reporting.py}, not through dynamics.
\end{itemize}
