% !TeX root = main.tex
\newpage
\section{Lighting Ball}

\subsection{Purpose}

\texttt{LightingBall} is the current lighting bring-up scenario after the
Cathedral and FreeStreamHetero studies were retired.  It keeps the useful
parts of the lighting work in a smaller, more inspectable scene: a stream of
mobile light particles approaches a spherical surface made visible by ordinary
boundary-marker particles.  The sphere gives a known three-dimensional shape
for testing lighting response, surface normals, reflection, and Vulkan export
parity.

\subsection{Configured Sphere}

The scenario uses the configuration item:
\begin{verbatim}
Lighting_ball = {
    x = x_center;
    y = y_center;
    z = z_center;
    radius = radius;
    material_id = material_id;
    wall_flag = wall_flag;
};
\end{verbatim}

The generator samples this sphere and inserts standard boundary particles at
the sampled locations.  These particles remain ordinary boundary markers:
they have boundary \texttt{ptype=2}, occupy cells for broad-phase
locality, and are not mobile physical objects.

The marker particles do not define the collision surface.  When a mobile
source discovers a local sphere marker, the dynamics evaluate the analytic
sphere equation from \texttt{Lighting\_ball}.  The sphere normal is:
\[
    \mathbf n =
    \frac{\mathbf x_{\mathrm{source}}-\mathbf c}
         {\lVert\mathbf x_{\mathrm{source}}-\mathbf c\rVert},
\]
where \(\mathbf c\) is the configured sphere center.  Physical penetration is
computed from the source radius and the signed distance to the analytic sphere
surface.  This keeps the boundary particles in the same role as wall markers:
they make nearby analytic geometry discoverable without becoming the geometry.

\subsection{Python Status}

The Python path is the working reference for the lighting ball.  The
generator \texttt{gbase/GenLightingBall.py} creates the sphere markers and
uses the shared streaming structure for the mobile particle stream.
\texttt{base/ForceDynamics.py} owns the analytic lighting-ball evaluator and
reflects a source from the configured sphere normal when the source is moving
into the surface.

Lighting display is currently reduced to the minimal photon tracer rule.  An
uncontacted photon uses its material debug color only when
\texttt{debug\_visible} is enabled.  A contacted photon uses the contacted
surface material color while \texttt{colFlg} is set.  No eye position, field
of view, diffuse term, specular term, boundary-light buffer, or
boundary-space spread is active in the cleaned \texttt{LightingBall}
configuration.

\subsection{Vulkan Status}

\texttt{LightingBall\_Vk.inc} is aligned with the Python include at the
configuration level.  It selects the same sphere, cell domain, death bounds,
stream geometry, material properties, radius, and timestep, but writes Vulkan
output names such as \texttt{LightingBall\_Vk.tst}.

The current Vulkan simple export now carries the lighting-ball sphere into the
runtime scene.  The generated test file writes \texttt{Lighting\_ball} as a
typed structure, and \texttt{ShaderObj} emits \texttt{sphere.glsl} with the
enabled state, center, radius, material id, and wall flag.  The simple exporter
emits \texttt{ForceDynamicsSimpleLightingBall.glsl}; the common
\texttt{FPM.comp} schedule routes a local marker belonging to the configured
lighting ball through the analytic sphere evaluator before ordinary
particle-particle collision handling.

Vulkan rendering also uses the material-owned color-mode contract.  Material
records contain \texttt{color\_mode} and \texttt{color}.  The generated
\texttt{material.glsl} exposes \texttt{COLOR\_MODE\_COLLISION},
\texttt{COLOR\_MODE\_VELOCITY}, \texttt{COLOR\_MODE\_SOLID}, and
\texttt{COLOR\_MODE\_LUMENS}.  The shared \texttt{color\_map.glsl} no longer
uses the old named solid color schemes.  In \texttt{LUMENS} mode, uncontacted
photons are transparent unless their photon material has
\texttt{debug\_visible=true}; contacted photons use the contacted surface
material color.  The retired eye, diffuse, and specular brightness controls
are not part of the active \texttt{LightingBall} contract.

Photon behavior is carried by runtime \texttt{ptype}, not by the current
\texttt{material\_id}.  During generation a stream whose base material has
\texttt{particle\_type="photon"} emits particles with photon \texttt{ptype}.
At the beginning of each compute frame the photon \texttt{material\_id} is
reset to the base photon material.  When the photon contacts the sphere or a
wall, the collision path sets the photon \texttt{material\_id} to the
contacted surface material so rendering can use that material as the emitted
color source.  No separate \texttt{hit\_material\_id} particle field is part
of the active ABI.

The active \texttt{LightingBall\_Py.inc} and
\texttt{LightingBall\_Vk.inc} files intentionally use only the ordinary
viewer controls:
\begin{verbatim}
view_center = [3.5, 2.5, 2.5];
zoom = 1.25;
gl_point_size = 2.0;
\end{verbatim}
The eye-aligned camera keys were removed from the active scene because the
eye no longer participates in lighting.  If that camera mode is needed later,
it should be reintroduced as view/camera configuration rather than as part of
the lighting model.

\subsection{Next Export Work}

The next export step is no longer porting the old tracer brightness model or
the removed eye-aligned view.  The remaining work is
\texttt{boundarySpaceLighting}: persistent boundary-proxy color state and
render-time spread over the visible surface:
\begin{itemize}
    \item keep photons as transport/debug particles;
    \item map photon contacts to existing boundary proxies;
    \item store persistent proxy color from the contacted surface material;
    \item render boundaries from sampled proxy color instead of drawing light
          only on the photon point;
    \item keep \texttt{color\_map.glsl} under exporter ownership so photon
          \texttt{ptype}, temporary photon \texttt{material\_id}, and material
          display behavior cannot drift from the generated Vulkan contract.
\end{itemize}
