% !TeX root = GeoDynamics.tex
\section{Particle Geometry}
\label{sec:particle-geometry}

This note records the current Geo particle contact model.  It intentionally
describes the code we have now, not the older Neo-style compression/rebound
model.  The current implementation is an instantaneous, frame-local geometric
response:

\begin{itemize}[leftmargin=*]
    \item contact state is rebuilt each frame,
    \item overlap area is computed from current particle geometry,
    \item pair stiffness is owned by the particles,
    \item the collision response applies a normal impulse derived from overlap
    area, stiffness, and source-owned contact internal momentum ledgers,
    \item particles are then moved with double-buffered positions.
\end{itemize}

No first-contact velocity, stored turnaround depth, or global collision
stiffness parameter is part of the current Geo dynamics.  The approved
persistent dynamics exception is contact-owned internal normal momentum stored
under the source particle and keyed by target particle.  Contact slots are
current-frame indexed records; they report current-frame contact state but do
not own persistent dynamics memory.

\section{Particle State}
\label{sec:particle-state}

For particle $i$, the fields used by the current model are:

\begin{equation}
\begin{aligned}
x_i &= (x_i,y_i,z_i), \\
v_i &= (v_{x,i},v_{y,i},v_{z,i}), \\
r_i &= \text{particle radius}, \\
m_i &= \text{particle mass}, \\
q_i &= \text{particle-owned collision stiffness}.
\end{aligned}
\end{equation}

In the GLSL-shaped particle layout these map to:

\begin{itemize}[leftmargin=*]
    \item current position: \texttt{PosLocA.xyz} or \texttt{PosLocB.xyz},
    selected by \texttt{ShaderFlags.positionBuffer},
    \item velocity: \texttt{VelRad.xyz},
    \item radius: \texttt{Data.x},
    \item collision stiffness: \texttt{Data.y},
    \item total active source internal momentum report: \texttt{Data.z},
    \item mass: \texttt{parms.x}.
\end{itemize}

The starting diagnostics shown by the runner are read-only values computed from
the initial particle state.  They are not written back into the dynamics.

\section{Contact Geometry}
\label{sec:contact-geometry}

For source particle $s$ and target particle $t$:

\begin{equation}
\begin{aligned}
\Delta x_{s,t} &= x_t - x_s, \\
d_{s,t} &= \lVert \Delta x_{s,t} \rVert.
\end{aligned}
\end{equation}

The particles are in contact when:

\begin{equation}
d_{s,t} \le r_s + r_t.
\end{equation}

When $d_{s,t} > 0$, the source-to-target unit normal is:

\begin{equation}
\hat{n}_{s,t} = \frac{\Delta x_{s,t}}{d_{s,t}}.
\end{equation}

When the centers are coincident, the current Python implementation uses the
fallback normal:

\begin{equation}
\hat{n}_{s,t} = (1,0,0).
\end{equation}

The penetration depth used for diagnostics and guardrails is:

\begin{equation}
\delta_{s,t} = r_s + r_t - d_{s,t}.
\end{equation}

\section{Circle Overlap Area}
\label{sec:circle-overlap-area}

The current particle runner uses the standard two-circle overlap area for the
contact area $A_{s,t}$.  If the circles do not overlap:

\begin{equation}
A_{s,t} = 0.
\end{equation}

If one circle fully contains the other:

\begin{equation}
A_{s,t} = \pi \min(r_s,r_t)^2.
\end{equation}

Otherwise:

\begin{equation}
\begin{aligned}
a_s &=
\frac{d^2 + r_s^2 - r_t^2}{2 d r_s}, \\
a_t &=
\frac{d^2 + r_t^2 - r_s^2}{2 d r_t}, \\
T &=
\frac{1}{2}
\sqrt{
(-d+r_s+r_t)(d+r_s-r_t)(d-r_s+r_t)(d+r_s+r_t)
}, \\
A_{s,t} &=
r_s^2 \cos^{-1}(a_s)
+ r_t^2 \cos^{-1}(a_t)
- T.
\end{aligned}
\end{equation}

In code, $a_s$ and $a_t$ are clamped into $[-1,1]$ before applying
$\cos^{-1}$ to avoid floating point domain errors.

\section{Relative Normal Velocity}
\label{sec:relative-normal-velocity}

The relative normal velocity is:

\begin{equation}
v_n = (v_t - v_s) \cdot \hat{n}_{s,t}.
\end{equation}

With the current normal convention:

\begin{itemize}[leftmargin=*]
    \item $v_n < 0$ means the particles are closing,
    \item $v_n = 0$ means there is no relative normal motion,
    \item $v_n > 0$ means the particles are separating.
\end{itemize}

The current Geo rule uses $v_n$ to determine closing momentum available for
compression.  Returning contacts may still receive release impulse while they
remain geometrically overlapped, but release is capped by the contact-owned
internal momentum ledger.

\section{Particle-Owned Pair Stiffness}
\label{sec:pair-stiffness}

Collision stiffness is particle owned.  There is no active global
\texttt{NEO\_COLLISION\_STIFFNESS\_Q} or
\texttt{GEO\_COLLISION\_STIFFNESS\_Q} in the current Geo path.

For a pair $(s,t)$, the current stiffness is the average of the two particle
fields:

\begin{equation}
q_{s,t} = \max\left(0,\frac{q_s + q_t}{2}\right).
\end{equation}

\section{Current Source-Owned Contact Response}
\label{sec:current-source-owned-contact-response}

The current implementation handles one or more particle contacts for a source
particle.  If the source has zero contacts, \texttt{GeoProcessCollisions}
returns without applying a collision response for that source.

For each contact $k$ between source $s$ and target $t_k$, the raw geometric
impulse budget is:

\begin{equation}
J_{raw,k} = q_{s,t_k} A_{s,t_k} \Delta t.
\end{equation}

For multi-contact frames, the source's active overlap area is:

\begin{equation}
A_{s,total} = \sum_j A_{s,t_j}.
\end{equation}

Each contact receives a geometric area weight:

\begin{equation}
w_{s,k} =
\frac{A_{s,t_k}}{A_{s,total}}.
\end{equation}

The source particle carries contact-owned internal normal momentum ledgers:

\begin{equation}
I_{s,t_k} \ge 0.
\end{equation}

This is source-owned writable memory keyed by target identity.  It is not
shared pair memory, and it is not owned by the current-frame slot index.  The
source particle's \texttt{Data.z} value is the reported total:

\begin{equation}
I_s = \sum_k I_{s,t_k}.
\end{equation}

When a target is no longer in the source's current contact list, that target's
ledger is pruned.  This prevents stale contact history from being released
into an unrelated later contact.

During compression, the source can store only momentum actually removed from
the frame-start relative closing motion.  The current Geo path uses
frame-local reduced-mass closing momentum:

\begin{equation}
\mu_{s,t_k} =
\frac{m_s m_{t_k}}{m_s + m_{t_k}},
\qquad
P_{close,k} =
\mu_{s,t_k}\max(0,-v_{n,k}).
\end{equation}

This value is not a stored \texttt{p\_zero} episode value.  It is recomputed
from the frame-start velocity snapshot for the current contact.  In a
multi-contact frame, the source's available closing momentum is shared across
contacts by area weight.  The same contact is also capped by the target's
read-only contact context so a one-contact target cannot overspend against a
multi-contact source:

\begin{equation}
P_{s,k} = w_{s,k} P_{s,available},
\qquad
P_{t,k} = w_{t,k} P_{t,available},
\qquad
P_{k,cap} = \min(P_{s,k}, P_{t,k}).
\end{equation}

The compression impulse is then:

\begin{equation}
J_{c,k} = \min\left(J_{raw,k}, P_{k,cap}\right).
\end{equation}

The contact-owned internal momentum is increased by $J_{c,k}$:

\begin{equation}
I_{s,t_k}' = I_{s,t_k} + J_{c,k}.
\end{equation}

When compression reaches the point where the source can no longer give the full
raw impulse, the source enters the returning phase.  During returning, the
source may release only the stored internal momentum for that same contact:

\begin{equation}
J_{r,k} =
\min\left(J_{raw,k}, I_{s,t_k} + J_{c,k}, I_{s,t_k}'\right).
\end{equation}

The contact-owned stored value is then reduced:

\begin{equation}
I_{s,t_k}' = \max(0, I_{s,t_k}' - J_{r,k}).
\end{equation}

The applied impulse for each contact is:

\begin{equation}
J_{applied,k} = J_{c,k} + J_{r,k}.
\end{equation}

Contact impulses are accumulated as one source-owned momentum vector:

\begin{equation}
\Delta \mathbf{p}_s =
-\sum_k J_{applied,k}\hat{n}_{s,t_k}.
\end{equation}

The source particle velocity update is applied once:

\begin{equation}
v_s' = v_s + \frac{\Delta \mathbf{p}_s}{m_s}.
\end{equation}

The target particle is not updated by the source call.  Instead, the target
will later be processed as a source particle and will apply its own source-side
update.  Before source-owned writes are applied, the current implementation
plans one frame-local impulse magnitude per unordered pair from snapshot data.
Each directed source side then applies the planned value through its own
source-owned write.  This preserves momentum symmetry but does not yet
guarantee kinetic-energy conservation.

The Python implementation must keep this source-owned write shape unless the
GLSL architecture changes first.  A Python-only pair solve that writes both
particles is forbidden because it would hide a dynamics rule that the shader
cannot reproduce safely.

\section{Frame Algorithm}
\label{sec:frame-algorithm}

For each simulation frame:

\begin{enumerate}[leftmargin=*]
    \item Clear each particle's current contact list, current-frame contact
    slots, and reporting fields.  Contact-owned internal momentum ledgers are
    preserved under the source particle until their target contact is inactive.
    \item Copy the frame-start positions into \texttt{PosLocFrame}.  Contact
    geometry and contact detection read from this frame-start snapshot.
    \item Copy the frame-start velocities into \texttt{VelRadFrame}.  Relative
    normal velocity and available momentum read from this frame-start snapshot.
    \item For each ordered pair $(s,t)$, test snapshot-position overlap.
    \item If the pair overlaps, add $t$ to $s$'s current contact list using the
    next indexed contact slot.
    \item Plan one impulse magnitude per unordered contact pair from snapshot
    data.  Each directed side contributes a candidate compression/release
    impulse, and the unordered pair uses the lower candidate magnitude.
    Diagnostic-only shadow plans may also be computed at this stage, but they
    must not change the applied source velocity.
    \item For each source particle $s$:
    \begin{enumerate}[leftmargin=*]
        \item If $s$ has active contacts, compute snapshot geometry, stiffness,
        area weights, planned impulses, and contact-owned internal momentum
        updates.
        \item Accumulate contact impulse vectors and apply one source velocity
        update.
    \end{enumerate}
    \item Move all particles into the inactive position buffer using their
    updated velocities and $\Delta t$.
    \item Flip \texttt{ShaderFlags.positionBuffer}.
    \item Sync the convenience aliases \texttt{rx}, \texttt{ry}, and
    \texttt{rz} from the newly active position buffer.
\end{enumerate}

The position update is:

\begin{equation}
x_i^{new} = x_i^{old} + v_i' \Delta t.
\end{equation}

\section{Reporting Diagnostics}
\label{sec:reporting-diagnostics}

The runner now reports frame number, simulation time, total momentum, and
kinetic energy.  These diagnostics are observational only.

Total momentum:

\begin{equation}
P = \sum_i m_i v_i.
\end{equation}

Total kinetic energy:

\begin{equation}
K = \sum_i \frac{1}{2}m_i \lVert v_i \rVert^2.
\end{equation}

Momentum drift is useful for detecting asymmetric errors:

\begin{equation}
\Delta P = P_{current} - P_{start}.
\end{equation}

Kinetic energy drift is the general detector for symmetric velocity drift:

\begin{equation}
\Delta K = K_{current} - K_{start}.
\end{equation}

For example, if two equal-mass particles drift from speeds $0.05$ to
$0.050006$ in opposite directions, total momentum can still be conserved while
$\Delta K$ reveals the speed growth.

\subsection{Contact Energy Diagnostics}

The contact capture file records the source-level momentum and kinetic-energy
effect of the contact response.  For a source particle $s$, the applied
contact impulses produce:

\begin{equation}
\Delta \mathbf{p}_s =
\sum_k \Delta \mathbf{p}_{s,k}.
\end{equation}

The source-level kinetic-energy estimate from the net impulse is:

\begin{equation}
\Delta K_{s,net}
=
\mathbf{v}_s \cdot \Delta \mathbf{p}_s
+
\frac{\lVert \Delta \mathbf{p}_s \rVert^2}{2m_s}.
\end{equation}

The capture also records the sum of contact-by-contact estimates:

\begin{equation}
\Delta K_{s,contacts}
=
\sum_k
\left(
\mathbf{v}_s \cdot \Delta \mathbf{p}_{s,k}
+
\frac{\lVert \Delta \mathbf{p}_{s,k} \rVert^2}{2m_s}
\right).
\end{equation}

The difference is reported as a source cross term:

\begin{equation}
\Delta K_{s,cross}
=
\Delta K_{s,net} - \Delta K_{s,contacts}.
\end{equation}

This is reporting-only.  It is used to see when simultaneous contact impulses
interact as a source-level vector sum instead of independent scalar contact
events.

\subsection{Shadow Dynamics}

Shadow dynamics are alternate response rules evaluated by a separate shadow
simulation object.  The live object remains
\texttt{GeoBase + GeoDynamics}; the shadow object is
\texttt{ShadowBase + ShadowDynamics}.  Both are initialized from the same
configuration data, but after initialization their particles, ledgers, phases,
velocities, positions, and reports are separate.  Shadow reports use
%\texttt{*_s.csv}.

Earlier in-place diagnostic families have served their purpose and are no
longer part of live \texttt{GeoDynamics.py}.  Those historical families
included source-scaled shadows, vector-rebound shadows, target-ratio shadows,
vector-internal shadows, and the first p-zero/parabolic shadow fields.  Future
alternate response rules should be implemented in \texttt{ShadowDynamics.py}
so they can mutate shadow state without touching live state.

For the current batch tests, the vector-rebound diagnostic showed that rebound
direction drift is not the main source of three-particle kinetic-energy loss:

\begin{equation}
\hat{d}_{compression}\cdot\hat{d}_{rebound} \approx 1.
\end{equation}

Therefore the active unresolved problem is the selection of simultaneous
multi-contact impulse magnitudes, not the rebound direction.

\section{Linear Shadow Contact Response}
\label{sec:linear-shadow-contact-response}

\texttt{ShadowLinDynamics} is the current readable experiment for applying the
same geometric model as one source-complete chain.  It is not yet the approved
replacement for the active Geo path, but it makes the intended per-source
equations and ownership boundary explicit.

For source \(s\), each contact \(k\) contributes overlap area \(A_k\), contact
normal \(\mathbf{n}_k\), and relative normal velocity:

\begin{equation}
v_{n,k} = (\mathbf{v}_{t_k}-\mathbf{v}_s)\cdot\mathbf{n}_k.
\end{equation}

Closing momentum is calculated from reduced mass:

\begin{equation}
p_{\mathrm{closing},k} = \mu_k \max(0,-v_{n,k}).
\end{equation}

After all source contacts are known, the source assigns an overlap-area weight:

\begin{equation}
w_k = \frac{A_k}{\sum_j A_j},
\qquad
p_{\mathrm{share},k} = w_k \sum_j p_{\mathrm{closing},j}.
\end{equation}

The current linear planned impulse is:

\begin{equation}
J_{\mathrm{raw},k}=q_k A_k\Delta t,
\end{equation}

\begin{equation}
J_{\mathrm{compression},k}
=
\min(J_{\mathrm{raw},k},p_{\mathrm{share},k})
\quad\text{when }v_{n,k}<-\epsilon,
\end{equation}

\begin{equation}
J_{\mathrm{release},k}
=
\min(J_{\mathrm{raw},k},I_k)
\quad\text{otherwise}.
\end{equation}

The source receives one velocity write from the vector sum:

\begin{equation}
\Delta\mathbf{p}_s
=
\sum_k
-(J_{\mathrm{compression},k}+J_{\mathrm{release},k})\mathbf{n}_k,
\qquad
\mathbf{v}'_s=\mathbf{v}_s+\frac{\Delta\mathbf{p}_s}{m_s}.
\end{equation}

After the velocity write, each source-owned contact ledger is updated:

\begin{equation}
I'_k
=
\max(0,I_k+J_{\mathrm{compression},k}-J_{\mathrm{release},k}).
\end{equation}

The unresolved question is whether this source-only allocation remains correct
for simultaneous multi-contact compression and rebound.  It must be evaluated
against the existing three-particle captures before promotion.

The elastic target compass is intended to remain a long-term diagnostic.  It
answers a reference question:

\begin{quote}
Given the current frame's active closing contact graph, what would an elastic
hard-contact result require?
\end{quote}

It is not the current Geo dynamics rule and should not be directly applied as
an overlap-area response.  The target-ratio and vector-internal shadows are
experiments.  Current captures show that target-ratio allocation is invalid for
rebound because separating contacts have no closing hard-contact target, and a
simple vector internal momentum ledger does not improve the three-particle
kinetic-energy drift.

The parabolic p-zero response is the current strongest shadow lead.  When a
contact first has closing normal momentum, the shadow stores

\begin{equation}
P_{0,k} = \mu_k \max(0,-v_{n,k})
\end{equation}

and computes the corresponding p-zero overlap area \(A_{0,k}\) from the same
force-law relation used by the old geometry model.  During the active contact,
the shadow does not recompute this reference from the growing internal
momentum ledger.  It uses the fixed overlap fraction

\begin{equation}
f_k = \operatorname{clamp}\left(\frac{A_k}{A_{0,k}},0,1\right)
\end{equation}

and the target internal momentum

\begin{equation}
I^{shadow}_{k}
=
P_{0,k}\left(1-\left(1-f_k\right)^2\right).
\end{equation}

The important distinction is that the curve itself is a target ledger, while
the shadow impulse is only the frame delta:

\begin{equation}
\Delta I^{shadow}_{k}=I^{shadow}_{k}-I^{current}_{k}.
\end{equation}

A previous attempt to promote this idea directly added p-zero release and
returning-phase branches to the applied path.  That made the decision tree too
abrupt and was backed out.  The next shadow implementation should instead use
one continuous target-ledger response and change the compression-to-rebound
flip to a velocity-magnitude tolerance:

\begin{equation}
|v_{n,k}| \leq \epsilon_v .
\end{equation}

Once a shadow contact flips to returning, it should remain returning until the
contact-owned shadow internal momentum drains or the contact ends.

The latest shadow result shows that draining the scalar internal momentum
ledger is not sufficient for angled contacts.  In \texttt{TwoParticleAngled},
the ledger reaches zero but the final kinetic energy remains low.  Because
energy is diagnostic-only, the next rule should not target energy directly.
Instead, the parabolic target should be imposed on relative normal velocity or
normal momentum.  The scalar internal momentum ledger then acts as the rebound
budget, while the actual impulse is chosen from the difference between current
and target normal momentum.

The normal-velocity shadow result restored both two-particle horizontal and
two-particle angled cases.  The remaining failure is now isolated to
multi-contact source allocation.  In \texttt{ThreeParticle}, particle 1 has two
simultaneous symmetric contacts.  Particle 2 and particle 3 each have one
dominant contact and their internal momentum profiles look correct.  Particle
1's stored internal momentum, however, peaks near the sum of its two contact
ledgers rather than the weighted source-level capacity:

\begin{equation}
I_{1}^{expected}
\approx
w_{12} I_{12} + w_{13} I_{13},
\qquad
w_{12}+w_{13}=1.
\end{equation}

For the symmetric capture, \(I_2\) and \(I_3\) peak near \(0.03408681\), so a
half-weighted source result should also be near \(0.03408681\).  The observed
particle 1 peak is approximately \(0.06160446\).  This shows that
multi-contact storage must be solved at the source level before distributing
per-contact impulses.

\section{Known Limitations}
\label{sec:known-limitations}

The current Geo rule is intentionally simple and incomplete:

\begin{itemize}[leftmargin=*]
    \item The impulse is based on overlap area, stiffness, and $\Delta t$; it
    uses contact-owned internal momentum to cap rebound/release and area
    weights to distribute multi-contact compression capacity, but it does not
    yet solve multi-contact kinetic-energy restoration.
    \item The source and target are not updated together in one pair solve.
    Pairwise write-back is intentionally forbidden under the current GLSL
    source-owned execution model.
    \item Contact slot indices are not contact identity.  Persistent contact
    ledgers are keyed by target under the source particle, while slots are
    rebuilt from the current frame's contact list.
    \item The response is not guaranteed to conserve kinetic energy.
    The normal-velocity shadow target restores the current two-particle
    horizontal and angled cases, but \texttt{ThreeParticle} still loses kinetic
    energy because particle 1's simultaneous contact storage is too large and
    its release must be allocated as one source-level vector result.
    \item Wall response is not part of this particle-pair geometry note.
\end{itemize}

These limitations are expected at this stage.  The current reporting exists so
we can see momentum conservation, kinetic energy drift, and tunneling behavior
while iterating on the next response rule.
