% !TeX root = main.tex
\newpage
\section{Photon Lighting Analysis}

\subsection{Current LightingBall State}

The current \texttt{LightingBall} implementation is still a photon-tracer and
collision-color bring-up path, not a surface-lighting system.  Photon
particles are generated as mobile runtime particles with photon \texttt{ptype}.
They move through the same source-owned compute schedule as other mobile
particles, but their optical rule is special:
\begin{itemize}
    \item photons ignore other photons;
    \item regular particles ignore photon targets;
    \item photon-source contacts with regular particles, walls, or the
          lighting sphere are optical events rather than mechanical contact
          resolution;
    \item photon speed is preserved by reflecting the velocity direction;
    \item \texttt{colFlg} marks the photon as in collision for the render
          frame;
    \item photon \texttt{material\_id} is temporarily changed to the contacted
          surface material so \texttt{LUMENS} can draw the photon with the
          surface color.
\end{itemize}

This is useful for debugging whether photons hit the correct surface and
whether the material color source is being read correctly.  It does not yet
make the sphere or walls look illuminated.  The visible light is still carried
by the moving photon points themselves.

\subsection{Why The Current Result Is Unsatisfying}

Direct photon rendering makes sparse sampling visible.  A small number of
photons produces isolated points, streaks, or momentary contact flashes.  Even
with denser emission, the render result is tied to where the photon particles
happen to be at the current frame rather than to accumulated light on the
surface.

The current \texttt{LUMENS} path also mixes two concepts that should be
separate:
\begin{itemize}
    \item photon tracer visibility: whether and how a photon particle itself
          should be drawn for debugging;
    \item surface illumination: how much light a boundary receives and later
          displays.
\end{itemize}

The eye-alignment brightness calculation belongs to the optional tracer path.
It is not a surface-lighting equation because moving the camera or eye should
not change how much light was deposited on a wall or sphere.

\subsection{Target Direction From PhotonLighting.tex}

\texttt{PhotonLighting.tex} describes the right target architecture:
photons are transport samples, while renderable boundaries accumulate and
display received light.  The intended pipeline is:
\begin{enumerate}
    \item move photon particles at fixed speed;
    \item detect photon-boundary intersections;
    \item compute hit position and boundary normal;
    \item compute incidence-weighted deposited energy;
    \item accumulate that energy into a boundary-light buffer;
    \item optionally spread the deposit over nearby boundary samples;
    \item temporally filter the boundary-light buffer;
    \item render the boundary from sampled boundary-proxy color;
    \item optionally render a sampled subset of photons as tracers.
\end{enumerate}

The important shift is that photons no longer need to be dense enough to fill
the visible surface.  They only need to sample light transport well enough for
the boundary-light buffer to converge to a usable illuminated image.

\subsection{What We Already Have}

Several pieces of the target system already exist in partial form:
\begin{itemize}
    \item photon generation through \texttt{particle\_streams};
    \item stable photon behavior through runtime \texttt{ptype};
    \item fixed-speed photon reflection;
    \item photon-wall, photon-sphere, and photon-particle optical contact
          detection;
    \item surface material color lookup through \texttt{material\_id};
    \item analytic sphere normals in the lighting-ball evaluator;
    \item rectangle-wall normals from \texttt{rectangle\_wall\_segments};
    \item material debug visibility and debug color for optional photon
          tracers.
\end{itemize}

These pieces are enough to prove transport, contact, and material-source
behavior.  They are not enough to produce smooth lighting because no boundary
stores received illumination.

\subsection{Missing Pieces}

The missing architecture is boundary-owned light state.  The active
\texttt{LightingBall} baseline now intentionally has no boundary-light buffer,
no filtered-light buffer, no surface-cell table, and no deposition pass.
Current rendering reads each particle's own material/color state.  It does
not read a persistent illumination value attached to a surface proxy.

The next target is the boundary-space proxy model.  To reach that model, the
system needs:
\begin{itemize}
    \item persistent boundary-proxy color state;
    \item a mapping from photon hit results to one existing boundary proxy;
    \item render-time spatial spreading across neighboring boundary proxies;
    \item render shaders that color boundaries from sampled proxy color;
    \item debug counters for active photons, hits, and proxy writes.
\end{itemize}
Atomics, additive current-frame accumulation, and temporal filtering are not
part of the first boundary-space implementation.  They remain possible later
extensions after the persistent proxy-color path is proven.

\subsection{Recommended Migration Path}

The safest path is incremental and keeps the current photon tracer behavior as
a diagnostic while surface lighting is built.

\subsubsection{Stage 1: Single-Proxy Boundary Color}

Use the existing photon-boundary contact events.  For each photon hit, compute:
\[
    C_{\mathrm{incidence}} =
    \max(-\mathbf d_p \cdot \mathbf n_b, 0)
\]
and deposit:
\[
    E_{\mathrm{deposit}} =
    E_p C_{\mathrm{incidence}} .
\]

Initially write exactly one boundary proxy: the nearest sphere marker, the hit
wall marker, or the marker associated with the analytic contact.  The proxy
stores the contacted material color until another photon overwrites it.  This
verifies the sign of the normal, the material color, and whether writes occur
on the expected object.

\subsubsection{Stage 2: Render Boundary-Light Values}

Add a boundary render path that reads accumulated light and applies it to the
surface base color:
\[
    \mathbf C_{\mathrm{final}} =
    \mathbf C_{\mathrm{base}}
    \left(A_{\mathrm{ambient}} + G\mathbf L_{\mathrm{filtered}}\right).
\]

At this stage the result may be speckled.  That is acceptable.  The goal is to
prove that the boundary, not the photon point, is emitting the visible light.

\subsubsection{Stage 3: Spatial Boundary-Space Spread}

Spread each rendered boundary point over neighboring valid proxies within an
influence radius.  Use a compact kernel such as:
\[
    w_j =
    \begin{cases}
        (1-r_j/R_L)^2, & r_j < R_L,\\
        0, & r_j \ge R_L.
    \end{cases}
\]

Normalize the weights so increasing the number of nearby proxies does not
increase brightness.  This turns sparse photon hits into a smooth surface
response without introducing a visible surface-cell grid.

\subsubsection{Stage 4: Optional Temporal Or Additive Light}

Only after persistent proxy color and render-time spread are visible should
the model consider current-frame additive accumulation, temporal filtering, or
decay.  Those features require extra state and, in Vulkan, may require
atomics or ping-pong buffers.  They are deliberately outside the first
boundary-space implementation.

\subsubsection{Stage 5: Material Optical Response}

Add optical material controls after deposition is visible:
\begin{itemize}
    \item absorption;
    \item reflection;
    \item roughness;
    \item photon energy reduction or termination.
\end{itemize}

Until this stage, boundaries can be treated as fully absorbing for deposition
while still reflecting photons for transport debugging.

\subsubsection{Deferred Directional Specular}

Photon-direction specular lighting is a useful later extension.  A boundary
sample could accumulate the incoming photon direction in addition to deposited
energy, and a coherent set of incoming directions could produce a highlight
from the photon simulation itself.  This is intentionally deferred until the
first scalar path is proven:
\[
    \textrm{photon hit}
    \rightarrow
    \textrm{boundary sample energy}
    \rightarrow
    \textrm{filtered lit sector}.
\]

\subsection{Boundary Sample Choice}

For \texttt{LightingBall}, the first practical boundary-space proxy is the
existing generated sphere marker.  It already has a particle index, a material
id, a position, and a nearby analytic sphere normal.  This makes it a good
first color-state target even though the marker is not the physical collision
surface.

A separate surface-sample set is deferred.  The current direction is to learn
from the existing boundary particles first, then decide whether a compact
Vulkan-only proxy table is worth adding.

\subsection{Effect On Current Debug Controls}

\texttt{debug\_visible} and \texttt{debug\_color} should remain photon-tracer
controls.  They answer whether uncontacted photons should be drawn so their
travel can be inspected.  They should not affect deposited light.

\texttt{gl\_point\_size} also remains a tracer/render diagnostic.  It changes
the apparent size of photon points but should not change the boundary-light
deposition radius.  The deposition radius belongs to a separate lighting
configuration value such as \texttt{photon\_light\_influence\_radius}.

\subsection{Near-Term Decision}

The current decision is to start with existing boundary particles as
persistent boundary-space proxies.  The cleaned scene removes the old
boundary-light sample/state path and visible surface-cell path so the next
implementation can prove one thing:
\[
    \textrm{photon hit}
    \rightarrow
    \textrm{persistent boundary proxy color}
    \rightarrow
    \textrm{kernel-spread surface color}.
\]
