% !TeX root = main.tex
\section{Python-glsl Export Rules }

Python is the readable canonical source for GLSL dynamics.  Naming,
structures, and the dynamics schedule should match as closely as practical.
The Python harness may replace GPU cell traversal with a direct scan, but it
must preserve the same source-owned processing order.

\subsection{Source Of Truth}

Python \texttt{base/ForceDynamics.py} is the export source for dynamics.
\texttt{ForceDynamicsBase.py} is Python infrastructure and diagnostics.

The active Python wall source contains generic two-dimensional function walls
and finite three-dimensional rectangle walls.  Boundary particles identify
locality with positive \texttt{ptype}; they do not carry evaluator IDs in
\texttt{Data.z}.  The canonical two-dimensional wall geometry is
\texttt{curve\_wall\_segments}.  The canonical three-dimensional wall
geometry is \texttt{rectangle\_wall\_segments}.  When
\texttt{rectangle\_wall\_segments} is present, it is the active simple wall
model and takes precedence over \texttt{curve\_wall\_segments}.  Horizontal,
vertical, analytic CD, finite-line, and manual \texttt{Wall*} paths are not
part of the simple export target.  Reservoir behavior is layered on top of
the simple target by an optional piston include and generator metadata; it
must not reintroduce scheduled birth or separate wall evaluators.

The current Vulkan parity target is:
\begin{verbatim}
vulkan/sim/TestPythonBoundarySpheres
\end{verbatim}
It uses the new \texttt{ForceDynamicsSimple*} GLSL family and intentionally
excludes reservoir behavior.  Reservoir wrappers may include the optional
\texttt{ForceDynamicsSimplePiston.glsl} file after the base simple includes.

\subsection{Stable Shader Wrapper}

The shared compute schedule currently lives in
\texttt{vulkan/sim/common/FPM.comp}.  Scenario compute shaders are generated
as thin wrappers around that common file.  They include generated scenario
metadata, generated simple helper GLSL, optional piston helpers, and then call
\texttt{fpm\_comp\_main()} from the common FPM file.

Generated GLSL may be regenerated, but the wrapper generator must not emit a
second inline copy of the FPM loop.  Keeping one common compute schedule avoids
regenerating stale duplicate-list logic, stale photon filtering, or stale
configured-wall dispatch into individual scenario wrappers.  Existing
contact-query helpers such as \texttt{isParticleContact.glsl} remain
hand-maintained shader infrastructure and are not rewritten by the
ForceDynamics export.

\texttt{ForceDynamicsSimple.glsl} exports reusable simple dynamics
functions.  The generated scenario compute shader owns contact discovery by
scanning source corner cells, deduplicating target IDs discovered through
multiple occupied cells, and calling the generated simple functions in the
approved order.  There is no marker-owned evaluator dispatch in the simple
path: boundary-marker targets always mean configured-wall locality.  The
configured simple wall evaluator is selected from scenario geometry:
\texttt{rectangle\_wall\_segments} for three-dimensional rectangle walls when
present, otherwise \texttt{curve\_wall\_segments} for two-dimensional function
walls.

\subsection{Generated Files}

The new simple generated files are:
\begin{itemize}
    \item \texttt{photons.glsl};
    \item \texttt{FPM.comp};
    \item \texttt{ForceDynamicsSimple.glsl};
    \item \texttt{ForceDynamicsSimpleBoundaryParticle.glsl};
    \item \texttt{ForceDynamicsSimpleFunctionWall.glsl};
    \item \texttt{ForceDynamicsSimpleRectangleWall.glsl};
    \item \texttt{ForceDynamicsSimpleLightingBall.glsl};
    \item \texttt{ForceDynamicsSimplePiston.glsl};
    \item \texttt{TestPythonBoundarySpheresP.comp}.
\end{itemize}

Generated scenario wrappers must include the material metadata before photon
helpers:
\begin{verbatim}
#include "material.glsl"
#include "../common/photons.glsl"
\end{verbatim}
Any scenario wrapper that includes \texttt{photons.glsl} requires a
\texttt{material.glsl} file with \texttt{PARTICLE\_TYPE\_REGULAR},
\texttt{PARTICLE\_TYPE\_PHOTON}, and a \texttt{MaterialProperty.particleType}
field.  Fixing Python export parity means every active generator path must
emit that material contract consistently before wrappers are regenerated.

\texttt{photons.glsl} owns the material lookup and optical photon helpers.  In
the simple path a photon-source/dust-target event is not an ordinary mechanical
particle collision.  The generated helper performs swept particle crossing,
fixed-speed reflection, and reflected end-of-frame position calculation.
\texttt{FPM.comp} consumes that event before the ordinary particle collision
processor and skips the mechanical force/contact path for that pair.  This
keeps photons from allocating contact slots or accumulating internal contact
momentum while preserving ordinary dust-dust and wall response.

\texttt{ForceDynamicsSimpleLightingBall.glsl} is the next required parity
helper for \texttt{LightingBall}.  It must export the Python analytic sphere
behavior from \texttt{EvaluateLightingBallContact} and
\texttt{ProcessLightingBallCollision}: boundary particles remain locality
markers, while the physical collision normal and penetration come from the
configured \texttt{Lighting\_ball} sphere equation.

They are produced by:
\begin{verbatim}
tools/ExportForceDynamicsSimpleGLSL.py
\end{verbatim}

The older generated dynamics files remain compatibility artifacts for older
scenario wrappers.  They are not regenerated by the active simple exporter:
\begin{itemize}
    \item \texttt{ForceDynamics.glsl} contains reusable core dynamics;
    \item \texttt{ForceDynamicsBoundaryParticle.glsl} contains straight
    boundary-marker wall helpers and the generic wall-segment declaration;
    \item \texttt{ForceDynamicsCDNozzle.glsl} retains the legacy analytic CD
    nozzle evaluator;
    \item \texttt{ForceDynamicsLinearReservoir.glsl} contains general
    finite-line wall helpers;
    \item \texttt{ForceDynamicsPiston.glsl} contains the analytic moving
    piston helpers;
    \item \texttt{ForceDynamicsParametricCurve.glsl} contains cubic
    parametric evaluation and bounded closest-point search for the older
    wrapper family.
\end{itemize}

The older exporter \texttt{tools/ExportForceDynamicsGLSL.py} is retained only
for historical inspection of that compatibility family.  Active porting work
must use \texttt{tools/ExportForceDynamicsSimpleGLSL.py}.

Ordinary wall and straight boundary-particle scenarios must not require CD
nozzle symbols.  A CD nozzle wrapper includes the core file, the generic
boundary-particle file, and then the CD nozzle file.  Other boundary-particle
wrappers include only the first two files.

\texttt{StartingContactState.glsl} is shared ABI/state layout and should be
edited carefully because C++ must match it.

\subsection{Particle ABI}

Only fields actually present in \texttt{particle.glsl} are legal shader
particle fields: \texttt{PosLocA}, \texttt{PosLocB}, \texttt{VelRadA},
\texttt{VelRadB},
\texttt{Data}, \texttt{parms}, \texttt{CornerList}, \texttt{contactCount},
\texttt{colFlg}, etc.  Python aliases like \texttt{rx}, \texttt{vx},
\texttt{mass}, \texttt{radius}, and reporting fields are not portable.

\subsection{Push-Constant Frame Clocks}

The shared particle push constants expose both render/application time and
simulation time.  They must not be treated as aliases:
\begin{itemize}
    \item \texttt{actualFrame} is the render/application frame and may advance
          while the application is stopped;
    \item \texttt{frameNum} is the simulation frame visible to dynamics,
          release gates, piston timing, and lifecycle checks;
    \item \texttt{positionBuffer} selects the current particle state buffer and
          must advance with \texttt{frameNum}, not with every rendered frame;
    \item \texttt{StopFlg} tells compute dispatches to skip particle dynamics
          while the scene continues to draw.
\end{itemize}

When \texttt{StopFlg} is set, \texttt{frameNum} and
\texttt{positionBuffer} remain frozen.  This prevents timed-release streams,
piston motion, lifecycle checks, and double-buffer selection from advancing
while \texttt{FPM.comp} is intentionally not processing particles.  Graphics
may continue to draw from the frozen simulation state.

\subsection{Compute Atomics}

Compute-shader atomic counters are defined by \texttt{atomic.glsl}.  Graphics
pipeline atomic/status input is defined separately by \texttt{atomicg.glsl}.
Generated ForceDynamics GLSL is included by the compute shader, so it must use
the compute atomic buffer from \texttt{atomic.glsl} and must not reference the
graphics pipeline \texttt{collIn} buffer from \texttt{atomicg.glsl}.

In the active compute shader, error/status state is reported through
\texttt{collOut.ErrorNumber}.  Generated ForceDynamics GLSL must not redefine
the compute atomic buffer or create a competing counter layout.

When C++ or any generator promotes numeric \texttt{.tst} values into GLSL
preprocessor macros, the macro body must be a bare expression token and must
not include a trailing semicolon.  For example:
\begin{verbatim}
#define FORCE_DYNAMICS_SIMPLE_COMPRESSION_STIFFNESS_GAIN 8.000000000
\end{verbatim}
is valid inside later expressions, while:
\begin{verbatim}
#define FORCE_DYNAMICS_SIMPLE_COMPRESSION_STIFFNESS_GAIN 8.000000000;
\end{verbatim}
expands a semicolon into the call site and can produce GLSL parser errors in
ordinary expressions such as \texttt{max(...)}.

The compute atomic structure is defined below:
\begin{verbatim}
layout (binding = 5) coherent buffer CollisionOut
{
	uint CollisionCount; 	// Number of colisions.
	uint numParticles;		// number of particle into compute.
	uint vnumParticles; 	// Number of particles emitted from compute to verterx.
	uint ErrorNumber;
	uint holdPidx;			// Reserved
	uint FrameNumber;
	uint ParticleNumber;

} collOut;
\end{verbatim}

\texttt{ParticleNumber} is the failing source-particle identifier for
source-owned generated errors.  It must not be overloaded onto
\texttt{numParticles}, because \texttt{numParticles} is also used for particle
count/debug accounting.

\subsection{Generated Boundary File}

Boundary parameters are generated from the scenario cfg into
\texttt{boundary.glsl}.  The simple path requires:
\begin{itemize}
    \item \texttt{wall\_contact\_offset};
    \item \texttt{death\_x\_min/max},
          \texttt{death\_y\_min/max}, and
          \texttt{death\_z\_min/max};
    \item \texttt{FunctionWallSegment} when
          \texttt{curve\_wall\_segments} is active;
    \item \texttt{CURVE\_WALL\_SEGMENT\_COUNT};
    \item \texttt{CURVE\_WALL\_SEGMENTS};
    \item \texttt{RectangleWallSegment} when
          \texttt{rectangle\_wall\_segments} is active;
    \item \texttt{RECTANGLE\_WALL\_SEGMENT\_COUNT};
    \item \texttt{RECTANGLE\_WALL\_SEGMENTS}.
\end{itemize}

Lighting-ball scenarios additionally require generated sphere metadata when
\texttt{Lighting\_ball} is present:
\begin{itemize}
    \item an enable/count constant for the configured lighting ball;
    \item the sphere center;
    \item the sphere radius;
    \item the sphere material identifier;
    \item the wall flag used for diagnostics and contact identity.
\end{itemize}

This metadata is consumed by
\texttt{ForceDynamicsSimpleLightingBall.glsl}.  The sampled sphere particles
remain positive-\texttt{ptype} boundary markers.  Their target IDs are
broad-phase locality records only; the shader must evaluate the analytic
sphere from the generated metadata rather than treating the marker particle
as the collision surface.

When \texttt{RECTANGLE\_WALL\_SEGMENT\_COUNT} is nonzero, the simple wall
path must evaluate rectangle walls and must not also evaluate
\texttt{CURVE\_WALL\_SEGMENTS} for the same source.  This mirrors Python's
configured-wall precedence and prevents two wall models with overlapping
\texttt{wall\_flag} values from accumulating duplicate physical wall forces.

Death bounds are required by the simple generated dynamics and are evaluated
unconditionally.  Generated simple GLSL must not wrap the death-plane check in
\texttt{DEATH\_BOUNDS\_AVAILABLE}.  A particle that crosses a death plane is
marked dead through \texttt{Data.w=-1} before it can drift onward to the
cell-array out-of-bounds error.

The simple path does not use \texttt{BOUNDARY\_XMIN},
\texttt{BOUNDARY\_XMAX}, \texttt{BOUNDARY\_YMIN},
\texttt{BOUNDARY\_YMAX}, or any derived \texttt{boundary\_*} values from the
\texttt{.tst} file.  Those fields are legacy simple-box metadata and must not
be regenerated by \texttt{GenericGenData}.

Boundary-particle wall participation is no longer selected by a marker-owned
dispatcher.  If a target is a boundary particle, it is a locality marker for
the active configured wall set.  For two-dimensional scenarios this means
\texttt{curve\_wall\_segments}; for three-dimensional rectangle scenarios this
means \texttt{rectangle\_wall\_segments}.  The evaluator is implied by the
simple model and scenario geometry, not stored in
\texttt{P[BoundaryID].Data.z}.

Reservoir piston wrappers additionally require:
\begin{itemize}
    \item \texttt{piston\_start\_frame};
    \item \texttt{piston\_x\_start};
    \item \texttt{piston\_x\_stop};
    \item \texttt{piston\_velocity\_x/y/z}.
\end{itemize}
Those constants are not required by ordinary simple boundary-sphere wrappers.
The simple piston include does not require \texttt{CHAMBER\_MIN/MAX} or
derived chamber-bound constants.

FreeStream wrappers are ordinary simple streaming wrappers, not reservoir
piston wrappers.  They must include the base simple dynamics, boundary
particle helper, and function-wall helper, but must not include
\texttt{ForceDynamicsSimplePiston.glsl} or define
\texttt{FORCE\_DYNAMICS\_SIMPLE\_PISTON\_AVAILABLE}.  If a freestream run
needs inlet timing, that timing belongs to the generated particle
\texttt{Data.w} release records rather than to a moving piston.

Linear-chamber wrappers include
\texttt{ForceDynamicsLinearReservoir.glsl} and define
\texttt{FORCE\_DYNAMICS\_LINEAR\_RESERVOIR\_AVAILABLE}.  Wrappers requiring a
piston include \texttt{ForceDynamicsPiston.glsl} independently.  Moving-wall contact
input carries \texttt{targetVelocity}; piston relative motion and
maximum-depth containment must use that velocity rather than zero.

Reservoir periodic direction is selected separately by:
\begin{verbatim}
#define PERIODIC_DIRECTION horizontal
#define PERIODIC_DIRECTION vertical
\end{verbatim}

The contact-force measure is selected by:
\begin{verbatim}
#define CONTACT_FORCE_MEASURE depth
#define CONTACT_FORCE_MEASURE area
\end{verbatim}
The current default cfg-generated value is \texttt{depth}.  Python still
computes overlap area for diagnostics and for the optional \texttt{area}
force mode.

\texttt{ForceDynamicsSimple.glsl} defines numeric selector tokens for
\texttt{horizontal\_wall}, \texttt{vertical\_wall},
\texttt{cd\_nozzle\_wall}, \texttt{horizontal}, \texttt{vertical},
\texttt{depth}, and \texttt{area} so preprocessor comparisons do not
accidentally compare undefined identifiers.

\subsection{Python-Only Infrastructure}

Config loading, reports, inline tests, scenario hooks, CSV diagnostics, and UI
plotting stay in Python.

\subsection{Rendering Diagnostics}

Rendering diagnostics are not part of the force law.  Scenario vertex shaders
may color particles from \texttt{colFlg} or from velocity angle, but those
colors must not affect generated dynamics.

Shader \texttt{debugPrintfEXT} output is also diagnostic state.  In the
Vulkan application it arrives through the debug-utils callback with message
ID \texttt{VVL-DEBUG-PRINTF}.  The callback should intercept that ID, strip
the wrapper text down to the message payload after \texttt{DebugPrintf:}, and
print the payload without setting \texttt{Extflg}.  Otherwise an intentional
shader print can behave like a stop condition because the generic
informational-message branch marks \texttt{Extflg}.

When GLSL uses \texttt{colFlg} as a collision-color diagnostic, the active
compute shader must clear \texttt{P[SourceID].colFlg} once per source
invocation before contact processing.  The generated force path may set
\texttt{P[SourceID].colFlg=1u} when \texttt{AccumulateContactForce(...)}
receives a valid particle or wall contact.  This mirrors the Python diagnostic
flag without reintroducing contact-slot storage as a force input.

Reservoir deferred-birth processing is the exception to diagnostic-only use of
\texttt{colFlg}.  Reserved values \texttt{2} and \texttt{3} represent waiting
and birth-candidate lifecycle states.  They remain outside the force law and
must not be cleared by the ordinary collision-color initialization before the
birth admission decision is complete.

For sphere rendering, the mesh resource should provide a unit-radius sphere
and the vertex shader should scale each instance by \texttt{P[index].Data.x}.
This keeps the rendered radius tied to the particle ABI rather than to a
separate scenario configuration radius or OBJ-file diameter.

Velocity-angle rendering is selected with material
\texttt{color\_mode = "VELOCITY"}.  The shared HSV utility
\texttt{colorizeVelocity(angle, sat, val)} expects \texttt{angle} in radians
from the velocity buffer selected with \texttt{ShaderFlags.positionBuffer};
\texttt{hsv\_sat} and \texttt{hsv\_val} remain scalar tuning values for that
mode.

\subsection{Known Translation Rules}

\texttt{create\_vec4(...)} maps to \texttt{vec4(...)}; Python dict
starting-contact state maps to the \texttt{StartingContactState} SSBO; Python
\texttt{None} return maps to GLSL boolean/result structs.

\subsection{Porting Issues Found}

\subsection{permanent host form}

The current ForceDynamics compute flow is:
\begin{verbatim}
main()
{
    vec3 totalForce = vec3(0.0);
    uint SourceID = gl_GlobalInvocationID.x;

    for (uint ii = 0; ii < 8; ii++)
    {
        uint loc = P[SourceID].CornerList[ii].ploc;

        for (uint jj = 0; jj < MAX_CELL_OCCUPANCY; jj++)
        {
            uint TargetID = clink[loc].idx[jj];
            if (TargetID == 0)
                break;

            if (isParticleContact(
                    ii, SourceID, TargetID,
                    uint(ShaderFlags.positionBuffer)) > 0
                && isDuplicate(SourceID) == False)
            {
                ProcessParticleCollision(
                    TargetID, SourceID, totalForce);
            }
        }
    }

    for (uint wall = 1; wall <= 4; wall++)
    {
        ProcessWallCollision(SourceID, wall, totalForce);
    }

    CalcVelocity(
        SourceID, totalForce,
        uint(ShaderFlags.positionBuffer));
    CalcPosition(SourceID, uint(ShaderFlags.positionBuffer));
}
\end{verbatim}

\subsection{Required Force Variable}

\texttt{totalForce} is a source-local variable passed as an \texttt{inout}
parameter to particle and wall collision processors.  Each processor adds its
force contribution immediately.  Contact slots may be recorded for
diagnostics, but ForceDynamics does not loop over them later to calculate
force.

\subsection{no source-ID validity check}

The GPU launches one invocation for each valid source particle, so the
dynamics invocation does not need a source-ID validity check.  The current
wall implementation directly scans four configured walls.  A later boundary
representation may place uniquely identified boundary segments in cells
without changing the source-owned force and integration schedule.

\section{Python-glsl Export Process}

\subsection{Step 1: Report The Export Surface}

Before generating GLSL, run the active simple exporter in report-only mode:
\begin{verbatim}
& "c:\_DJ\gPCD\python\.venv\Scripts\python.exe" tools\ExportForceDynamicsSimpleGLSL.py
\end{verbatim}

This tool parses \texttt{base/ForceDynamics.py}, verifies that the methods
used by the \texttt{ForceDynamicsSimple*} family are present, and reports
whether any expected generated surface has drifted.  Without \texttt{--write}
it performs validation only and leaves all GLSL files untouched.

This step must pass before any GLSL file is generated.  Any unexpected
dependency means the Python dynamics source still depends on infrastructure
that either needs to move into \texttt{ForceDynamics.py} or needs an explicit
translation rule.  Python direct-scan helpers remain wrapper/harness parity
methods and are not part of the generated simple GLSL force path.

\subsection{Step 2: Generate The Simple GLSL Family}

After the report-only gate passes, generate the active simple GLSL family:
\begin{verbatim}
& "c:\_DJ\gPCD\python\.venv\Scripts\python.exe" tools\ExportForceDynamicsSimpleGLSL.py --write
\end{verbatim}

This writes:
\begin{itemize}
    \item \texttt{vulkan/sim/common/ForceDynamicsSimple.glsl};
    \item \texttt{vulkan/sim/common/ForceDynamicsSimpleBoundaryParticle.glsl};
    \item \texttt{vulkan/sim/common/ForceDynamicsSimpleFunctionWall.glsl};
    \item \texttt{vulkan/sim/common/ForceDynamicsSimpleRectangleWall.glsl};
    \item \texttt{vulkan/sim/common/ForceDynamicsSimplePiston.glsl};
    \item \texttt{vulkan/sim/TestPythonBoundarySpheres/TestPythonBoundarySpheresP.comp}.
\end{itemize}

Generated files should not be hand-edited.  Fixes belong in
\texttt{tools/ExportForceDynamicsSimpleGLSL.py}, the stable wrapper shader
template, or generated scenario configuration such as \texttt{boundary.glsl}.

\subsection{Historical Exporter}

The retired compatibility exporter:
\begin{verbatim}
tools/ExportForceDynamicsGLSL.py
\end{verbatim}
generated the older \texttt{ForceDynamics.glsl},
\texttt{ForceDynamicsBoundaryParticle.glsl}, CD-nozzle, linear-reservoir,
piston, and parametric-curve include family.  That exporter and file family
are retained only for historical comparison and old backup wrappers.  They
must not be used for new rectangle-wall, function-wall, reservoir, or simple
Vulkan work.

\subsection{Historical Step 3: Classify Translation Status}

Before translating function bodies, classify every generated method by
translation status.  The exporter report groups methods as:
\begin{itemize}
    \item \texttt{direct}: can be translated nearly 1:1 now;
    \item \texttt{needs\_struct}: needs a GLSL result struct or temporary state
    object;
    \item \texttt{needs\_buffer}: depends on
    \texttt{StartingContactState.glsl} or particle/storage layout;
    \item \texttt{defer}: depends on later interface decisions.
\end{itemize}

Every method in the generated GLSL list must appear in exactly one
translation-status group.  This gate prevents body generation from starting
with unresolved interface questions hidden inside the method list.  Direct
methods may be translated first.  Methods in the other groups require their
GLSL structs, buffers, or call contracts to be designed before code generation
is enabled for them.

\subsection{Step 4: Define Geometry Result Structs}

Python geometry helpers often return either a tuple of values or
\texttt{None}.  GLSL does not use that convention.  These returns are
translated into explicit result structs with a \texttt{valid} flag, or a
named suppression flag where that is the actual Python meaning.

The generated \texttt{ForceDynamics.glsl} stub declares:
\begin{itemize}
    \item \texttt{ParticleEffectiveContactGeometry};
    \item \texttt{ParticleGeometry};
    \item \texttt{ParticlePotentialGeometry};
    \item \texttt{WallPhysicalGhostGeometry};
    \item \texttt{WallGhostGeometry}.
\end{itemize}

The corresponding method signatures return those structs instead of generic
\texttt{bool} values.  This keeps the Python tuple layout visible while giving
GLSL a single value to return from each geometry helper.

\subsection{Step 5A: Generate Direct Read-Only And Math Bodies}

The first real body-generation pass is limited to direct methods that only do
math, read particle data, or write the shared error code.  These bodies are
generated from explicit templates rather than inferred from the Python AST.
This keeps the first generated code small and reviewable.

The Step 5A template set includes:
\begin{itemize}
    \item \texttt{VelocityAngle(...)};
    \item \texttt{particle\_position(...)};
    \item \texttt{particle\_overlap\_area(...)};
    \item \texttt{GetParticlePosition(...)};
    \item \texttt{ParticleCenterDistance(...)};
    \item \texttt{GetStartFrameVelocity(...)};
    \item \texttt{GetPairStiffness(...)};
    \item \texttt{GetContactStiffness(...)};
    \item \texttt{WallContactOffsetDistance(...)};
    \item \texttt{GetParticleMass(...)};
    \item \texttt{SetError(...)}.
\end{itemize}

State-writing integration methods such as \texttt{CalcVelocity(...)} and
\texttt{CalcPosition(...)} remain stubbed until the shader buffer-write
contract is verified.  Geometry and starting-contact methods remain governed
by their struct and buffer design steps.

\subsection{Step 5B: Generate Integration Write Bodies}

After the read-only templates are generated, the exporter may generate the
state-writing integration methods:
\begin{itemize}
    \item \texttt{CalcVelocity(...)};
    \item \texttt{CalcPosition(...)}.
\end{itemize}

\texttt{CalcVelocity(...)} reads the current velocity through
\texttt{GetParticleVelocity(...)} and writes only the alternate velocity
through \texttt{SetNextParticleVelocity(...)}.  It reads
\texttt{ShaderFlags.dt}, checks for invalid time step, reads the source
particle mass, applies the source-local \texttt{totalForce}, and stores the
velocity angle in the selected output buffer's \texttt{.w} component.

\texttt{CalcPosition(...)} implements the position double buffer.  When
\texttt{ShaderFlags.positionBuffer == 0}, the method reads \texttt{PosLocA}
and writes the next position into \texttt{PosLocB}.  Otherwise it reads
\texttt{PosLocB} and writes into \texttt{PosLocA}.  It then flips the
\texttt{.w} active flags using the same convention as Python.  The generated
method must never update the same position buffer that contact detection is
reading during that frame.

Velocity uses the same selector and direction as position:
\texttt{positionBuffer=0} reads \texttt{VelRadA} and writes
\texttt{VelRadB}; selector one reads \texttt{VelRadB} and writes
\texttt{VelRadA}.  \texttt{ApplySourceMaximumDepth(...)} modifies the alternate
candidate velocity, and \texttt{CalcPosition(...)} integrates with that same
accepted alternate velocity.

The lower-bound out-of-bounds check remains matched to Python:
\texttt{ERROR\_PARTICLE\_OUT\_OF\_BOUNDS} is set when boundary mode is off and
the output x or y coordinate is below 1.0.

\subsection{Step 6: Generate Ordinary Particle Geometry}

After the direct math and integration methods are generated, the exporter may
generate the ordinary particle geometry path:
\begin{itemize}
    \item \texttt{GetParticleEffectiveContactGeometry(...)};
    \item \texttt{GetParticlePotentialGeometry(...)};
    \item \texttt{GetParticleGeometry(...)}.
\end{itemize}

At this stage \texttt{GetParticleEffectiveContactGeometry(...)} uses the
physical particle radii from \texttt{P[SourceID].Data.x} and
\texttt{P[TargetID].Data.x}.  It returns \texttt{startingContact = 0.0},
\texttt{startingResolved = 0.0}, and \texttt{suppressContact = false}.  This
creates a valid ordinary-contact geometry path while keeping starting-contact
effective-radius behavior disabled until the
\texttt{StartingContactState.glsl} buffer logic is implemented.

\texttt{GetParticleGeometry(...)} returns a \texttt{ParticleGeometry} struct
with \texttt{valid = false} when there is no active overlap.  Coincident
centers use \texttt{vec3(1.0, 0.0, 0.0)} as the deterministic fallback normal,
matching the Python implementation.

\subsection{Step 7: Generate Ordinary 2D Wall Geometry}

The exporter may generate the ordinary wall geometry path after particle
geometry is available:
\begin{itemize}
    \item \texttt{GetPhysicalWallGhostGeometry(...)};
    \item \texttt{GetWallGhostGeometry(...)}.
\end{itemize}

For now ForceDynamics uses the 2D wall flags from the Python model:
\texttt{1} for \texttt{BOUNDARY\_XMIN}, \texttt{2} for
\texttt{BOUNDARY\_XMAX}, \texttt{3} for \texttt{BOUNDARY\_YMIN}, and
\texttt{4} for \texttt{BOUNDARY\_YMAX}.  The wall bounds are read from
\texttt{boundary.glsl}; they are not derived from
\texttt{ShaderFlags.SideLength} and are not hardcoded in generated dynamics.

\texttt{GetPhysicalWallGhostGeometry(...)} returns a
\texttt{WallPhysicalGhostGeometry} struct with \texttt{valid = false} when the
wall flag is invalid or the source particle does not overlap the wall ghost.
\texttt{GetWallGhostGeometry(...)} wraps that physical geometry with ordinary
physical radius behavior.  Starting-wall effective-radius behavior remains
disabled until the \texttt{StartingContactState.glsl} buffer logic is
implemented.

The cfg may later provide z-wall constants, but Step 7 does not enable z-wall
processing.  The current generated wall path remains 2D to match the live
Python wall loop.

\subsection{Step 8: Generate Boundary-Particle Wall Helpers}

After ordinary wall geometry is available, the exporter may generate the
boundary-particle wall helper path.  Boundary particles are occupancy markers,
not mobile dynamics targets.  When the wrapper selects
\texttt{WALL\_MODEL\_BOUNDARY\_PARTICLE}, marker targets discovered during the
ordinary target scan are evaluated as wall segments and accumulated as wall
forces.

The required generated methods are:
\begin{itemize}
    \item \texttt{IsBoundaryParticle(...)};
    \item \texttt{BoundaryParticleWallFlag(...)};
    \item \texttt{BoundaryParticleVerticalWallFlag(...)};
    \item \texttt{BoundaryParticleCDNozzleWallFlag(...)};
    \item \texttt{CDNozzleRadius(...)};
    \item \texttt{CDNozzleRadiusSlope(...)};
    \item \texttt{EvaluateWallSegment(...)};
    \item \texttt{EvaluateHorizontalWallSegment(...)};
    \item \texttt{EvaluateVerticalWallSegment(...)};
    \item \texttt{EvaluateCDNozzleWallSegment(...)};
    \item \texttt{EvaluateParametricWallSegment(...)};
    \item \texttt{ProcessBoundaryParticleWallCollision(...)}.
\end{itemize}

\texttt{BoundaryParticleAppliesToSource(...)} is Python-only and must not be
exported.  The Python implementation uses it to imitate cell locality during
direct scans.  GLSL receives that locality from the cell occupancy structure,
so the wrapper only sees boundary marker targets that already share a relevant
cell with the source.

\texttt{EvaluateWallSegment(...)} returns the transient wall geometry for the
marker target.  The boundary-particle processor should build a local
\texttt{ContactForceInput} with \texttt{CONTACT\_WALL} and accumulate the force
immediately, just as \texttt{ProcessWallCollision(...)} does for manual wall
flags.

The simple path has no marker-owned evaluator dispatcher.  Each boundary
marker has \texttt{ptype=1} and exists only to make wall evaluation local.
When such a target is discovered, the generated simple shader evaluates the
active configured wall set, chooses the valid wall contact by physical
\texttt{wallFlag}, and accumulates that force immediately.  If
\texttt{rectangle\_wall\_segments} is present, the active wall set is
\texttt{RECTANGLE\_WALL\_SEGMENTS}; otherwise it is
\texttt{CURVE\_WALL\_SEGMENTS}.  \texttt{Data.z} is not a boundary evaluator
channel, and \texttt{pdata.temp\_vel} remains reserved.

The older horizontal, vertical, analytic CD-nozzle, finite-line, and
marker-owned evaluator-ID dispatch paths are historical compatibility work.
They are not part of the \texttt{ForceDynamicsSimple*} export.

Marker target IDs are broad-phase records, not physical wall identities.  A
source may discover two adjacent markers for one wall when its extent spans
their cells.  The periodic and CD wrappers therefore evaluate the candidate
segment and deduplicate valid boundary contacts with a source-local bit mask
keyed by \texttt{BoundaryWallSegment.wallFlag}.  The bit is set only after a
valid segment is successfully accumulated.  At most one force contribution
per physical wall flag is applied to a source in one frame.  Ordinary mobile
particle contacts retain their target-ID duplicate list.

\texttt{WALL\_MODEL\_BOUNDARY\_PARTICLE} remains a wrapper-owned compile-time
selection.  It controls whether marker targets are processed as walls; it does
not choose which evaluator a particular marker represents.  The simple
configured-wall evaluator is selected from the generated scenario geometry,
with rectangle walls taking precedence when both wall sets are present.

The Python diagnostic initializer:
\begin{itemize}
    \item \texttt{InitializeBoundaryParticleWallContactState(...)}
\end{itemize}
may be exported later for Python diagnostic parity, but it is not required for
the generated GLSL force path.  Generated GLSL should not export the Python
contact-slot initializer when the boundary-particle wall collision can be
computed directly from \texttt{EvaluateWallSegment(...)} and transient force
input.

\subsection{Step 9: Use Transient Contact Force Input}

Generated ForceDynamics GLSL does not store contacts in particle-owned contact
slots.  The active \texttt{particle.glsl} ABI does not contain
\texttt{P[SourceID].contacts[...]}, and generated dynamics must not
reintroduce that old layout.

Instead, generated GLSL uses a transient \texttt{ContactForceInput} struct:
\begin{verbatim}
struct ContactForceInput {
    uint targetID;
    uint contactType;
    vec3 normal;
    float overlapArea;
    float penetrationDepth;
    bool valid;
};
\end{verbatim}

The shader computes current contact geometry, builds a local
\texttt{ContactForceInput}, immediately accumulates that contact into the
source-local \texttt{totalForce}, and then discards the temporary contact data.
This keeps the generated force path independent of Python diagnostic contact
slots.

\texttt{AccumulateContactForce(...)} is therefore generated against
\texttt{ContactForceInput}.  The generated accumulator selects the force
measure with \texttt{CONTACT\_FORCE\_MEASURE}.  The current default generated
configuration uses \texttt{depth}, which multiplies stiffness by
\texttt{penetrationDepth}.  The alternate \texttt{area} mode multiplies
stiffness by \texttt{overlapArea}, preserving the earlier overlap-area force
law.  The accumulator also updates source-owned diagnostic internal momentum
in \texttt{P[SourceID].parms.yzw} and mirrors its magnitude to
\texttt{P[SourceID].Data.z}.  This value is not used by
\texttt{CalcVelocity(...)}.  Python methods that allocate or populate
diagnostic contact slots, such as \texttt{AppendContactSlot(...)},
\texttt{GetContactSlots(...)}, and \texttt{GetContactState(...)}, remain
deferred for Python parity and are not part of the generated force-storage
model.

\subsection{Step 10: Generate Collision Processors}

Once transient force input exists, generated GLSL may implement the collision
processor methods:
\begin{itemize}
    \item \texttt{ProcessParticleCollision(...)};
    \item \texttt{ProcessWallCollision(...)};
    \item \texttt{ProcessBoundaryParticleWallCollision(...)}.
\end{itemize}

These generated processors do not initialize, append, or store contact slots.
\texttt{ProcessParticleCollision(...)} computes a \texttt{ParticleGeometry},
builds a local \texttt{ContactForceInput} with
\texttt{CONTACT\_PARTICLE}, and calls \texttt{AccumulateContactForce(...)}.
\texttt{ProcessWallCollision(...)} does the same with
\texttt{WallGhostGeometry} and \texttt{CONTACT\_WALL}.
\texttt{ProcessBoundaryParticleWallCollision(...)} does the same with the wall
segment geometry returned by \texttt{EvaluateWallSegment(...)}.

The Python diagnostic setup methods \texttt{InitializeContactState(...)},
\texttt{InitializeWallContactState(...)}, \texttt{GetContactState(...)},
\texttt{AppendContactSlot(...)}, \texttt{GetContactSlots(...)}, and
\texttt{AppendWallContactSlot(...)} remain deferred.  They describe Python
reporting/contact-slot behavior, not the generated GLSL force path.

\subsection{Step 11: Generate ForceDynamicsSimple Constants}

Before compiling the generated include, \texttt{ForceDynamicsSimple.glsl}
must define the constants used by generated simple ForceDynamics code.  These
constants belong at the top of the generated simple include because they may
change with the ForceDynamics model.  The current required constants are:
\begin{itemize}
    \item \texttt{CONTACT\_PARTICLE};
    \item \texttt{CONTACT\_WALL};
    \item \texttt{ERROR\_INVALID\_DT};
    \item \texttt{ERROR\_PARTICLE\_OUT\_OF\_BOUNDS}.
\end{itemize}

\texttt{CONTACT\_PARTICLE} and \texttt{CONTACT\_WALL} may have matching
values elsewhere, but generated ForceDynamicsSimple owns the constants it
needs.  Do not modify \texttt{sim/common/constants.glsl} for
ForceDynamics-only constants.  Wall dimensions and wall-contact configuration
are the exception: those are generated from scenario configuration into
\texttt{boundary.glsl}.

\subsection{Step 12: Export Wall Contact Configuration}

\texttt{wall\_contact\_offset} is generated by the application with the other
wall parameters in \texttt{boundary.glsl}.  Generated ForceDynamics wall
geometry reads that symbol directly, for example:
\begin{verbatim}
const float wall_contact_offset = 0.20;
\end{verbatim}

Generated ForceDynamicsSimple wall code should read the
application-generated symbol rather than hardcoding a cfg value.  If the
symbol is absent, wall geometry export is not complete even if the Python
dynamics are correct.

\subsection{Step 13: Wrapper-Owned Wall Model Selection}

\texttt{ForceDynamicsSimple.glsl} and its helper includes export reusable
wall and contact processors, but the scenario wrapper owns contact discovery.
For the active simple path, boundary marker targets discovered during the
target scan are treated as configured-wall locality and evaluated by the
generated configured-wall helper.

\begin{itemize}
    \item \texttt{WALL\_MODEL\_SIMPLE}: scan manual wall flags
    \texttt{1..4}.
    \item \texttt{WALL\_MODEL\_PIPE}: scan manual wall flags
    \texttt{3..4}; the \(x\) planes are reservoir lifecycle planes rather
    than solid walls.
    \item \texttt{WALL\_MODEL\_BOUNDARY\_PARTICLE}: do not run a manual wall
    scan.  Boundary marker targets discovered during the target scan are
    evaluated as wall segments.
\end{itemize}

The wall numbering convention for manual wall models is \texttt{1} for xmin,
\texttt{2} for xmax, \texttt{3} for ymin, and \texttt{4} for ymax.

\subsection{Step 14: Match Error Reporting To The Active Atomic Include}

\texttt{BoundarySpheresReservoirP.comp} includes \texttt{atomic.glsl}, whose
error field is \texttt{collOut.ErrorNumber}.  Generated
\texttt{SetError(...)} must therefore write:
\begin{verbatim}
collOut.ErrorNumber = error_code;
\end{verbatim}

It must not write \texttt{collIn.ErrorReturn}.  That field belongs to the
graphics pipeline \texttt{atomicg.glsl} layout, not the compute shader
\texttt{atomic.glsl} layout.

\subsection{Step 15: Compile Trial}

Only after Steps 11 through 14 are satisfied should the generated include be
compiled through permanent compute shaders for each wall model.  The compile
trial is now a three-model check:
\begin{verbatim}
WALL_MODEL_SIMPLE
WALL_MODEL_PIPE
WALL_MODEL_BOUNDARY_PARTICLE
\end{verbatim}

The expected compute targets include:
\begin{verbatim}
sim/BoundarySpheresReservoir/
    BoundarySpheresReservoirP.comp
sim/BoundaryParticleReservoirHorizontal/
    BoundaryParticleReservoirHorizontalP.comp
sim/BoundaryCDNozzleReservoir/
    BoundaryCDNozzleReservoirP.comp
\end{verbatim}

A compile failure at this stage should be classified as either a missing
wrapper/helper symbol, a missing generated boundary/config symbol, an
incorrect wall-model define, or an incorrect particle/atomic ABI name.  It
should not be fixed by hand-editing the generated
\texttt{ForceDynamicsSimple*.glsl} files; fixes belong in the simple exporter,
the stable wrapper shader template, or the generated config includes.

\subsection{Historical Step 15A: Retired Reservoir Lifecycle Helpers}

This step describes the previous scheduled-birth Vulkan path.  The live
Python Force model and the new horizontal generator no longer use these
helpers.  Do not port or regenerate them for the replacement prefilled
reservoir.  They remain documented only to identify wrapper code that must be
removed when Vulkan migration resumes.

The reservoir compute path uses the newer \texttt{Data.w} lifecycle rule:
\begin{itemize}
    \item \(\texttt{Data.w} > 0\): stored birth frame;
    \item \(\texttt{Data.w} = 0\): active particle;
    \item \(\texttt{Data.w} < 0\): dead particle.
\end{itemize}

The previous exporter emitted two custom reservoir utility functions:
\begin{verbatim}
StartReservoir(SourceID)
RetireParticlePastXMax(SourceID)
\end{verbatim}

These helpers are no longer emitted.  Their names remain here only to identify
and remove stale wrapper calls.

\texttt{StartReservoir(SourceID)} is called at the beginning of the source
invocation.  It returns \texttt{false} for dead particles and for unborn
particles whose birth frame has not arrived.  When the current frame reaches
the stored birth frame, it places the particle at the inlet birth position,
updates both position buffers, sets \texttt{Data.w=0}, and returns
\texttt{true}.

\texttt{RetireParticlePastXMax(SourceID)} is called after
\texttt{CalcPosition(SourceID)}.  It reads the next position buffer, the buffer
written by \texttt{CalcPosition}, and sets \texttt{Data.w=-1} when the source
center passes the selected outlet boundary.  The historical function name is
retained for wrapper compatibility.  The generated body uses
\texttt{PERIODIC\_DIRECTION}: horizontal flow births and retires along \(x\),
while vertical flow births and retires along \(y\).

The compute shader sequence is:
\begin{verbatim}
if (!StartReservoir(SourceID))
    return;

// contact scan and force accumulation
CalcVelocity(SourceID, totalForce);
CalcPosition(SourceID);
RetireParticlePastXMax(SourceID);
\end{verbatim}

These helpers are lifecycle gates.  They do not alter the ForceDynamics force
law, and they do not write target particles.

The current helper activates a particle immediately when its birth frame
arrives.  Capture analysis confirmed that this can admit a new wave into a
slowed live particle near the inlet, creating artificial overlap compression
and kinetic energy.  Deferred admission was specified but not implemented.
Python now replaces scheduled birth with a prefilled moving-piston reservoir.
The remaining porting work is to export its analytic position and velocity,
one-per-source piston contact, moving-wall maximum-depth constraint, and
piston-impulse diagnostics to GLSL and the Vulkan wrapper.

\subsection{Step 16: Vulkan Runtime Validation}

After the compile trial succeeds, run the ForceDynamics scenarios through the
Vulkan compute path for all selected wall models.  The current generated
\texttt{ForceDynamicsSimple*.glsl} family has been tested successfully for the compiled
particle and wall scenarios exercised so far.  That validation covers:
\begin{itemize}
    \item ordinary particle contacts;
    \item \texttt{WALL\_MODEL\_SIMPLE};
    \item \texttt{WALL\_MODEL\_PIPE};
    \item \texttt{WALL\_MODEL\_BOUNDARY\_PARTICLE}, including the replacement
    horizontal reservoir and parametric CD-nozzle wrappers;
    \item source-local \texttt{totalForce};
    \item \texttt{CalcVelocity(...)};
    \item \texttt{CalcPosition(...)};
    \item double-buffered \texttt{VelRadA}/\texttt{VelRadB} state;
    \item source-local maximum-depth projection and blocked-momentum storage;
    \item six-plane mobile-particle death bounds.
\end{itemize}

\subsection{Step 16A: Permanent Death State}

The replacement prefilled reservoir uses \texttt{Data.w} only as permanent
mobile-particle life state: zero is alive and a negative value is dead.
\texttt{ApplyParticleDeathBounds(...)} reads the alternate position written by
\texttt{CalcPosition(...)} and tests \texttt{death\_x\_min/max},
\texttt{death\_y\_min/max}, and \texttt{death\_z\_min/max}.  Crossing any plane
sets \texttt{Data.w=-1}.  Dead particles are skipped as compute sources,
contact targets, occupancy entries, and rendered particles.  Positive
boundary-marker \texttt{ptype} records are never killed.

\texttt{ResolveStartingCollision} is not included in this Vulkan validation.
Starting-contact behavior depends on persistent starting-contact state, and
that state is still deferred to the \texttt{StartingContactState.glsl}
buffer/export path.  Until that buffer path is implemented, the generated GLSL
ordinary force path should be considered validated separately from starting
overlap resolution.

\subsection{Step 17: Boundary OBJ Visualization}

Boundary OBJ meshes are visualization resources, not dynamics resources.  For
boundary-particle models, the dynamics still come from marker particles and
the selected \texttt{EvaluateWallSegment} path.  A generated OBJ such as:
\begin{verbatim}
sim/BoundaryCDNozzleReservoir/CDNozzleGenerated.obj
\end{verbatim}
is used only to draw the intended wall shape.  The CD generator now rewrites
this OBJ from the same \texttt{curve\_wall\_segments} sampling used to place
locality markers.  The visual mesh and collision profile therefore share one
geometry source and coordinate system.

The current Vulkan OBJ loader accepts only triangle faces in
\texttt{v/vt/vn} form.  Generated OBJ files must therefore include dummy
texture coordinates and normals even when the renderer does not use them:
\begin{verbatim}
v  ...
vt ...
vn ...
f v/vt/vn v/vt/vn v/vt/vn
\end{verbatim}

The graphics command path must select its draw call from
\texttt{application.boundary\_as\_obj}.  When \texttt{boundary\_as\_obj} is
true, the boundary resource is a non-indexed OBJ vertex list and must be drawn
with:
\begin{verbatim}
vkCmdDraw(..., dvo->m_NumElements, 1, 0, 0);
\end{verbatim}
When \texttt{boundary\_as\_obj} is false, the old cube/cell-boundary path
retains its indexed cube draw:
\begin{verbatim}
vkCmdDrawIndexed(..., 36, 1, 0, 0, 0);
\end{verbatim}
This preserves old cube and sphere rendering while allowing generated nozzle
OBJ visualization.
