% !TeX root = main.tex
\section{Problem Evolution}

\subsection{Source-Local Area Weighting Breaks Conservation}

\subsubsection{<previous form>}

The unweighted geometric pair impulse was:

\begin{equation}
J_{ij}=q_{ij}A_{ij}\Delta t.
\end{equation}

Because pair overlap area and pair stiffness are symmetric,
\(A_{ij}=A_{ji}\) and \(q_{ij}=q_{ji}\), independently evaluated source and
target impulses have equal scalar magnitude.

\subsubsection{<current form>}

The failed weighted form applied a source-local overlap-area weight to the raw
impulse:

\begin{equation}
J_{ij}
=
q_{ij}A_{ij}
\left(
\frac{A_{ij}}{\sum_k A_{ik}}
\right)
\Delta t.
\end{equation}

The denominator is the total overlap area for the current source.  When the two
particles have different surrounding contact sets:

\begin{equation}
\sum_k A_{ik}\ne\sum_k A_{jk},
\end{equation}

and therefore:

\begin{equation}
J_{ij}\ne J_{ji}.
\end{equation}

\subsubsection{difference outcome}

No solution.

\subsubsection{<Analysis>}

The source-owned execution structure remains valid: each source reads the same
frame-start snapshot, vector-sums its contact contributions, and writes only
its own velocity.  The conservation failure is in the source-local impulse
magnitude.  Different source contexts calculate different magnitudes for the
same pair contact, so the independently applied impulses are not
equal-and-opposite.

The current \texttt{ThreeParticle} weighted capture ends with:

\begin{verbatim}
vector momentum drift = approximately 0.04899135
kinetic-energy drift  = +0.00090284
internal momentum     = 0
\end{verbatim}

The internal-momentum ledgers drain completely, and accumulated compression
and release impulse balance, but whole-system vector momentum and kinetic
energy do not return to their starting values.  Complete scalar ledger return
therefore does not establish conservation.

Source-local area weights may describe a source allocation, but multiplying
them into the physical pair impulse breaks pair compatibility.  Removing that
factor restored equal-and-opposite pair evaluation and exact vector momentum
in the \texttt{ThreeParticle} test.  Kinetic-energy drift remained, motivating
the separate central-force model.

\subsection{Overlap Area as a Central Force Field}

\subsubsection{<previous form>}

\texttt{WeightedDynamics} stores compression impulse in source-owned contact
ledgers and returns that scalar impulse during separation.

\subsubsection{<current form>}

\texttt{ForceDynamics} applies the central pair force
\(F_{ij}=q_{ij}A_{ij}\) during every overlapping timestep.  It stores no
contact momentum and uses
\(U_{ij}(d)=q_{ij}\int_d^{r_i+r_j}A_{ij}(s)\,ds\) for energy diagnostics.

\subsubsection{difference outcome}

Particle solution.

\subsubsection{<Analysis>}

The force is symmetric for both independently evaluated sources, so pair
impulses are equal-and-opposite.  The changing contact-normal direction is
handled directly by the central force rather than by returning a stored scalar
impulse along a later normal.

At \(\Delta t=0.05\), the current \texttt{ThreeParticle} capture ends with:

\begin{verbatim}
vector momentum drift = 0
total-energy drift    = approximately -0.00000017
potential energy      = 0
\end{verbatim}

Equal-time timestep refinement drives the completed-collision energy error
toward zero.  The current \texttt{ForceDynamics} implementation immediately
accumulates contact forces into a source-local \texttt{totalForce}, updates
velocity once, and then updates position with semi-implicit Euler.  Vector
momentum remains conserved, while integration produces a small bounded energy
excursion.

\subsection{Statistical Energy Validation Requires Walls}

\subsubsection{<previous form>}

Two- and three-particle tests measured bounded energy excursions and small
completed-collision residuals.  They established timestep convergence and
showed that completed-contact force-work residual equals final total-energy
residual, but they do not establish the long-term statistical behavior of many
collisions.

\subsubsection{<current form>}

A long-running particles-in-a-box test now exercises many elastic particle and
wall collisions.  It measures whether the numerical energy excursion remains
bounded or accumulates into systematic drift.

\subsubsection{difference outcome}

Particle solution.

\subsubsection{<Analysis>}

Walls are now stationary equal-particle ghosts processed by a direct
four-wall scan.  Their conservative overlap-derived force and potential
participate in the same source-local dynamics and energy diagnostics as
particle contacts.

In the current 10,000-step \texttt{ParticlesInBox} capture, total-energy error
rises and falls between approximately \(-1.85\times10^{-5}\) and
\(+1.90\times10^{-5}\), with a final error of approximately
\(-7.34\times10^{-7}\).  The result is bounded over this run rather than
monotonic, enabling continued statistical validation over longer runs and
multiple random seeds.

\subsection{Return ForceDynamics to One Source Invocation}

\subsubsection{<previous form>}

\texttt{ForceDynamics} used velocity Verlet, requiring a frame-start force
scan, predicted positions, a second force scan, and stored current and
predicted accelerations.

\subsubsection{<current form>}

Each source invocation now detects particle and wall contacts once,
immediately accumulates their forces into source-local \texttt{totalForce},
calls \texttt{CalcVelocity}, and then calls \texttt{CalcPosition}.  Contact
slots remain available for diagnostics but are not force inputs.

\subsubsection{difference outcome}

Full solution.

\subsubsection{<Analysis>}

The current structure maps directly to a GLSL compute invocation without a
global synchronization point or second force scan.  It preserves the central
overlap-area force law, source ownership, and equal-and-opposite independent
particle-pair evaluation.  Semi-implicit Euler has a larger instantaneous
excursion than velocity Verlet in small collision tests, but the current
\texttt{ParticlesInBox} result is bounded and performs better than the Verlet
implementation in that long-running test.

\subsection{ForceDynamics GLSL Export Reaches Vulkan Runtime}

\subsubsection{<previous form>}

Python \texttt{ForceDynamics} was shaped for GLSL, but the generated
\texttt{ForceDynamics.glsl} path was still a compile and integration
experiment.  The export path had to settle forward declarations, compute
atomic error reporting, ForceDynamics-owned constants, the wall-contact
wrapper contract, and application-generated boundary configuration.

\subsubsection{<current form>}

Generated \texttt{ForceDynamics.glsl} now runs through the Vulkan compute path
for the ordinary scenarios tested so far.  The permanent compute shader owns
the particle cell scan and directly scans the four configured walls.  The
generated dynamics include owns ForceDynamics constants, source-local force
accumulation, ordinary particle and wall geometry, velocity update, and
position update.  Wall dimensions and \texttt{wall\_contact\_offset} remain
application-generated in \texttt{boundary.glsl}.

\subsubsection{difference outcome}

Particle solution.

\subsubsection{<Analysis>}

The ordinary ForceDynamics path has crossed from Python-only validation into
Vulkan runtime validation.  The working path includes ordinary particle
contacts, ordinary wall contacts, direct four-wall processing, source-local
\texttt{totalForce}, \texttt{CalcVelocity}, and \texttt{CalcPosition}.

\texttt{ResolveStartingCollision} is intentionally not part of this completed
validation.  Starting collisions require persistent starting-contact state,
and that state still needs a GLSL buffer/export implementation through
\texttt{StartingContactState.glsl}.  Until that path is implemented, the
export result is a particle solution for ordinary particle and wall dynamics,
with starting-overlap resolution deferred.

\subsection{Retire MomentumDynamics}

\subsubsection{<previous form>}

\texttt{MomentumDynamics} remained registered as a selectable runtime model
without an active model source document.

\subsubsection{<current form>}

\texttt{MomentumDynamics} is removed from the active runtime registry.  Its
model-specific Python files are retained under
\texttt{python/base/old/MomentumDynamics}.

\subsubsection{difference outcome}

Full solution.

\subsubsection{<Analysis>}

The active selectable models are now \texttt{ForceDynamics} and
\texttt{VerAForceDynamics}, and both have corresponding source documents
included by \texttt{main.tex}.  \texttt{VerAForceDynamics} is an independently
wired baseline copy used as the shadow model before changes are made to the
live implementation.  Retiring rather than deleting old models preserves
their implementation history without presenting them as maintained.

\subsection{Resolve Starting Overlap Without Extra Rebound}

\subsubsection{<previous form>}

\texttt{ForceDynamics} treated a starting overlap exactly like any later
physical overlap.  The physical radii defined both the force magnitude and the
diagnostic potential energy from the first frame.  In
\texttt{ResolveStartingCollisions}, this meant the initial configuration
stored a large physical-overlap potential and produced a stronger rebound than
the same collision would have produced if it had evolved into that overlap
from first contact.

\subsubsection{<current form>}

Live \texttt{ForceDynamics} records contacts that are already overlapping at
initialization.  For each starting particle or wall contact, the starting
center distance becomes an effective separation limit.  The force and
potential-energy diagnostics use the effective geometry while the starting
contact is active, so the initial configured overlap contributes no force
until the contact compresses deeper than the starting separation.

When the contact rebounds back to its starting separation, the starting contact
is marked resolved.  Physical-radius force is then suppressed until the actual
physical overlap clears, preventing a second rebound from the configured
overlap.  \texttt{VerAForceDynamics} remains the shadow baseline that uses the
previous physical-radius behavior.

\subsubsection{difference outcome}

Particle solution.

\subsubsection{<Analysis>}

The fix keeps normal contacts unchanged: non-starting particle and wall
scenarios remain physically identical to the VerA baseline.  Starting-overlap
scenarios intentionally diverge from VerA because live ForceDynamics removes
the initial artificial stored potential.  Current targeted runs show
\texttt{ResolveStartingCollisions}, \texttt{ResolveStartingWallCollision}, and
\texttt{ResolveStartingWPCollision} completing without runtime errors and
draining their starting-contact state.

The result is local to starting contacts and avoids an external evaluator.  It
does introduce persistent contact state, so future GLSL work must carry that
state outside the transient per-frame contact slots.  The result is still
marked a particle solution because broader multi-contact starting-overlap
validation remains future work.

\subsection{Reservoir Flow Uses Data.w Lifecycle}

\subsubsection{<previous form>}

Reservoir flow used older ad hoc state names and shader-side release logic.
The Python analysis path and Vulkan path did not share the same simple
birth/live/death rule, and repeated reservoir columns could create artificial
starting collisions if unborn particles were included in contact processing.

\subsubsection{<current form>}

\texttt{SimpleReservoir} now uses \texttt{Particle.Data.w} as the lifecycle
field:
\begin{itemize}
    \item \(\texttt{Data.w} > 0\): unborn particle; value is the birth frame;
    \item \(\texttt{Data.w} = 0\): live particle;
    \item \(\texttt{Data.w} < 0\): dead particle.
\end{itemize}

Python \texttt{ForceDynamics} gates contact scans, wall scans, velocity
updates, position updates, and diagnostics by this live-state rule.  When a
particle reaches its birth frame, it is moved to the inlet birth position with
a \(1.1\) diameter offset and both position buffers are synchronized.  After
position update, a particle that crosses the outlet is marked dead.

The Vulkan reservoir compute path now calls
\texttt{StartReservoir(SourceID)} at the beginning of the source invocation and
\texttt{RetireParticlePastXMax(SourceID)} after \texttt{CalcPosition}.  This
matches the Python lifecycle rule while leaving \texttt{CalcPosition}
unchanged.

\subsubsection{difference outcome}

Particle solution.

\subsubsection{<Analysis>}

The reservoir state is a source-local lifecycle gate, not a change to the
ForceDynamics force law.  Unborn and dead particles return before collision
processing, so repeated stored columns can exist in the input data without
generating artificial overlap, tunneling, wall contact, or diagnostic energy.
Live particles continue to use the ordinary source-owned overlap-area force
model.

The GLSL call order remains compatible with one source invocation:
\begin{verbatim}
if (!StartReservoir(SourceID))
    return;

// source-owned contact scans and force accumulation
CalcVelocity(SourceID, totalForce);
CalcPosition(SourceID);
RetireParticlePastXMax(SourceID);
\end{verbatim}

The current result is a particle solution because the reservoir lifecycle is
working in the active Python and Vulkan reservoir paths.  Broader throughput,
long-run statistical energy, and inlet packing studies remain future
validation work.

\subsection{SimpleReservoirPeriodic Vertex Lifecycle Output}

\subsubsection{<previous form>}

\texttt{SimpleReservoirPeriodicP.vert} used the reservoir lifecycle gate to
return early for unborn and dead particles:

\begin{verbatim}
if (!ParticleLifeActive(index, uint(ShaderFlags.frameNum)))
    return;
\end{verbatim}

This prevented inactive particles from reaching the cell-occupancy insertion
code, but it also returned from the vertex shader before writing a defined
\texttt{gl\_Position}, point size, or output color.  In practice, inactive or
retired particles could still appear as stale visual artifacts, including a
red particle drawn near the center of the screen.  Increasing the cell-array
side length could hide other pressure symptoms, but it did not address the
undefined vertex output.

\subsubsection{<current form>}

Inactive reservoir particles now write a defined offscreen vertex before the
early return:

\begin{verbatim}
if (!ParticleLifeActive(index, uint(ShaderFlags.frameNum))) {
    gl_Position = vec4(2.0, 2.0, 2.0, 1.0);
    gl_PointSize = 0.0;
    fragColor = vec3(0.0);
    return;
}
\end{verbatim}

The offscreen clip-space position keeps the vertex outside the visible region,
while the zero point size and neutral color remove stale display state.

\subsubsection{difference outcome}

Full solution.

\subsubsection{<Analysis>}

The fix does not change reservoir birth, death, collision detection, or
ForceDynamics.  It only makes the vertex-stage lifecycle gate produce defined
graphics output when a particle is not renderable or processable.  Because
\texttt{discard} is not available in vertex shaders, explicitly moving the
vertex outside clip space is the appropriate stage-local way to suppress
inactive particles.

This also clarifies the debugging interpretation: a red particle drawn near
the center was a vertex-output artifact, not proof that a dead particle was
being included in contact processing.  Compute-side lifecycle filtering still
uses \texttt{StartReservoir(SourceID)} for sources and skips targets whose
\texttt{Data.w} is not live.

\subsection{Velocity-Angle HSV Python Visualization}

\subsubsection{<previous form>}

The Python viewer used fixed particle colors with a collision-color override.
This made collision state easy to see, but it did not expose flow direction or
changes in particle velocity angle during reservoir and box tests.

\subsubsection{<current form>}

The Python viewer now supports \texttt{RUN\_CONFIGURATION.hsv\_color}.  When
enabled, \texttt{SimulationRunner\_HSVPanel.py} colors each active particle from
\texttt{VelRad.w} using \texttt{gbase/utilities.py:hsv\_angle}.  The helper
maps a radians angle to hue using \(1/(2\pi)\), clamps
\texttt{hsv\_sat} and \texttt{hsv\_val} to \([0,1]\), and returns a Python RGB
tuple.

\subsubsection{difference outcome}

Full solution.

\subsubsection{<Analysis>}

The color change is diagnostic only.  It reads the velocity angle already
maintained by the dynamics model, but it does not feed color back into contact
detection, force accumulation, integration, reservoir lifecycle, or energy and
momentum reports.

Keeping the HSV helper in \texttt{gbase/utilities.py} and applying it in
\texttt{SimulationRunner\_HSVPanel.py} preserves the boundary between portable dynamics
logic and Python display logic.  A later GLSL utility can mirror the same
\texttt{hsv\_angle(angle, value, sat)} interface for Vulkan rendering without
changing the ForceDynamics collision model.

\subsection{BoundarySpheresAndCube Vulkan Visualization Parity}

\subsubsection{<previous form>}

\texttt{BoundarySpheresAndCube} rendered particles with fixed collision
colors and a sphere mesh whose size was baked from the scenario radius during
vertex-buffer creation.  The generated ForceDynamics GLSL force path applied
contacts, but it bypassed the Python diagnostic contact-slot code that set
\texttt{colFlg}, so shader collision colors could fail to reflect current
contacts.

\subsubsection{<current form>}

The rendered sphere resource now stores a unit-radius mesh, and the
\texttt{BoundarySpheresAndCube} sphere vertex shader scales each instance by
the particle radius in \texttt{P[index].Data.x}.  The compute shader clears
\texttt{P[SourceID].colFlg} once per source invocation, and
\texttt{AccumulateContactForce(...)} sets it when a valid particle or wall
contact is accumulated.

\texttt{BoundarySpheresAndCube} also supports scenario-controlled HSV
velocity-angle rendering through \texttt{HSV\_ON}, \texttt{HSV\_SAT}, and
\texttt{HSV\_VAL} in \texttt{params.glsl}.  The shared GLSL utility
\texttt{colorizeVelocity(angle, sat, val)} reads \texttt{VelRad.w} in radians
and maps it to hue with \(1/(2\pi)\).

\subsubsection{difference outcome}

Full solution.

\subsubsection{<Analysis>}

The result separates display diagnostics from dynamics.  Particle radius,
velocity angle, and collision state are read from the particle ABI, while the
renderer owns mesh scaling and color conversion.  Collision coloring and HSV
coloring are now selectable visual interpretations of the same dynamics state.
Neither feeds back into contact detection, force accumulation, integration, or
energy and momentum diagnostics.

Because \texttt{ForceDynamics.glsl} is generated content, the
\texttt{colFlg} assignment in \texttt{AccumulateContactForce(...)} should be
moved into the exporter template when the generated file is regenerated.

\subsection{Crowding Requires Internal Momentum, Not Energy Repair}

\subsubsection{<previous form>}

The reservoir boundary-particle wall test converted solid pipe walls from
manual configured planes into cell-occupancy boundary markers.  A mobile
particle that encounters a \texttt{ptype=1.0} boundary marker evaluates a wall
segment and applies the ordinary source-owned wall force.

Under crowding, particles near the wall can compress into the particles behind
them.  The observed failure mode is tunneling in the crowded particle column,
not a failure of the boundary marker itself.

\subsubsection{<current form>}

The project rule is now momentum-first: energy is diagnostic bookkeeping, not
a governing correction.  No global step should reach backward to force a
kinetic-energy or total-energy value.  A particle responds only to the current
geometry, current force, current momentum, and stored local contact state.

When overlap compression cannot be resolved as translational motion during the
current frame, the unresolved part should be treated as internal normal
momentum or contact stress.  The stored quantity is local and causal.  It must
later release only through current contact geometry and current contact
normals.  This is the pressure channel required for crowded contacts.

\subsubsection{difference outcome}

Analysis direction.

\subsubsection{<Analysis>}

The crowding case exposes a missing physical channel.  In a real material,
pressure information propagates through contacts at a finite wave speed,
related to the material speed of sound.  The entire particle column does not
know the future rebound path; pressure builds and propagates contact by
contact.

Therefore the next model step should not be a maximum-penetration clamp or an
energy correction.  A clamp can be retained as a diagnostic failure threshold,
but it does not explain where the impulse went.  The correct bookkeeping
question is whether the current impulse became translational particle
momentum, wall/constraint momentum, or stored internal contact momentum.

For a particle-particle crowding contact, the internal state should be scalar
normal momentum owned by the contact or contact pair.  Direction must come
from the current contact normal at release time, not from a stored arbitrary
vector.  For a wall contact, the same idea represents local pressure against
the wall segment or constraint.  Energy diagnostics may later report the
consequence of this bookkeeping, but they must not govern the update.

\subsection{GLSL Wall Models And Boundary-Particle Export}

\subsubsection{<previous form>}

The generated \texttt{ForceDynamics.glsl} path was validated for ordinary
particle contacts and manual wall scans, but the scenario compute wrappers
owned different hardcoded wall loops.  Simple walls scanned flags \(1..4\),
pipe/reservoir walls scanned flags \(3..4\), and boundary-particle walls were
still a Python-side model rule.  Regenerating the common include could also
drop hand-maintained reservoir helpers such as \texttt{StartReservoir} and
\texttt{RetireParticlePastXMax}.

\subsubsection{<current form>}

\texttt{ForceDynamics.glsl} now exports reusable boundary-particle wall
helpers:

\begin{verbatim}
IsBoundaryParticle
BoundaryParticleWallFlag
EvaluateWallSegment
ProcessBoundaryParticleWallCollision
\end{verbatim}

The scenario \texttt{*P.comp} wrapper selects exactly one wall model:

\begin{verbatim}
WALL_MODEL_SIMPLE
WALL_MODEL_PIPE
WALL_MODEL_BOUNDARY_PARTICLE
\end{verbatim}

Simple wrappers scan wall flags \(1..4\).  Pipe wrappers scan wall flags
\(3..4\).  Boundary-particle wrappers do not run a manual wall scan; they
route \texttt{ptype=1.0} target records from cell occupancy to
\texttt{ProcessBoundaryParticleWallCollision}.  The exporter also emits
\texttt{StartReservoir} and \texttt{RetireParticlePastXMax} as custom GLSL
helpers so reservoir wrappers survive regeneration of the common include.

\subsubsection{difference outcome}

Full solution for wall-model wrapper selection and boundary-particle export.
Tunneling detection remains an error-detection extension.

\subsubsection{<Analysis>}

The important architectural separation is now explicit.  The generated
dynamics include owns reusable force and geometry functions.  The stable
scenario wrapper owns contact discovery, lifecycle gates, and wall-model
selection.  Boundary-particle locality is supplied by the GPU cell occupancy
structure; the Python-only \texttt{BoundaryParticleAppliesToSource} helper is
not exported because it only imitates that locality during direct scans.

The current contact queries remain sampled overlap tests.  They can report
contact when the current distance is inside the radius tolerance, but they do
not yet perform swept crossing tests.  Tunneling should therefore be detected
as an error in the hand-maintained contact-query or wall-segment shader path,
not hidden as a dynamics correction.

\subsection{Vertical Boundary-Particle Reservoir Flow In Vulkan}

\subsubsection{<previous form>}

The boundary-particle reservoir Vulkan path initially reused the horizontal
reservoir assumptions.  Boundary marker contacts could be routed through the
boundary-particle wall path, but the reservoir lifecycle helpers still placed
newly born particles and retired escaped particles along the \(x\) direction.
The active wall-segment function and the reservoir flow direction were also
easy to confuse because both appeared in generated shader configuration.

\subsubsection{<current form at that stage>}

The generated boundary configuration now separates the two selectors:

\begin{verbatim}
#define WALL_FUNC vertical_wall
#define PERIODIC_DIRECTION vertical
\end{verbatim}

At that stage, \texttt{WALL\_FUNC} chose the boundary-marker wall-segment evaluator:
\texttt{EvaluateHorizontalWallSegment} or
\texttt{EvaluateVerticalWallSegment}.  \texttt{PERIODIC\_DIRECTION} chooses
the reservoir lifecycle direction.  When it is \texttt{vertical},
\texttt{StartReservoir(SourceID)} preserves the generated \(x\) coordinate
and moves the particle to \(y_{\min}+2.2r\).  The existing
\texttt{RetireParticlePastXMax(SourceID)} wrapper-compatible helper then
retires the particle when its next position passes \(y_{\max}\).  When the
selector is not vertical, the historical \(x\)-direction behavior remains.

The later Python self-describing-marker migration supersedes the
\texttt{WALL\_FUNC} evaluator decision: marker \texttt{Data.z} now selects the
wall evaluator.  \texttt{PERIODIC\_DIRECTION} remains a separate lifecycle
direction.  Vulkan synchronization with marker-owned evaluator IDs is the next
porting step.

During validation, apparent downward screen motion was traced to the viewing
transform rather than to the reservoir force or lifecycle rule.  In
\texttt{Resource::GeneralViewing}, the configured zero-angle view still
receives hard-coded \(180^\circ\) rotations about \(z\) and \(y\), and the
Vulkan projection applies the normal clip-space y correction.  Therefore a
positive simulation \(y\) velocity can appear as downward motion depending on
the current view.

\subsubsection{difference outcome}

Full solution for generated GLSL wall-function selection and vertical
reservoir lifecycle selection.  View-orientation cleanup remains a display
task.

\subsubsection{<Analysis>}

The key correction was not to infer reservoir flow from wall geometry inside
the dynamics law.  Wall response and lifecycle direction are separate
configuration facts:

\begin{itemize}
    \item \texttt{WALL\_FUNC} describes the wall segment represented by a
    boundary marker.
    \item \texttt{PERIODIC\_DIRECTION} describes the axis used for reservoir
    birth and retirement.
\end{itemize}

This keeps the generated \texttt{ForceDynamics.glsl} reusable while letting
the scenario wrapper and generated config includes own scenario-specific
selection.  The remaining view issue should be diagnosed as rendering
orientation, not particle dynamics, unless direct position/velocity reports
show negative \(y\) motion in simulation coordinates.

\subsection{Depth Default For Contact Force Measure}

\subsubsection{<previous form>}

\texttt{ForceDynamics} originally used overlap area directly as the contact
force measure:
\begin{equation}
F_{ij}=q_{ij}A_{ij}.
\end{equation}
This made overlap-area geometry both the diagnostic quantity and the default
force magnitude input.

\subsubsection{<current form>}

\texttt{ForceDynamics} now keeps overlap area in Python diagnostics and keeps
area mode available, but the default configuration uses penetration depth:
\begin{verbatim}
contact_force_measure = "depth";
\end{verbatim}
The generated GLSL path receives the same selection through:
\begin{verbatim}
#define CONTACT_FORCE_MEASURE depth
\end{verbatim}
In depth mode the force measure is \(r_i+r_j-d_{ij}\) for particle contacts
and the analogous wall-ghost penetration depth for wall contacts.  In area
mode the earlier \(A_{ij}\) measure is used.

\subsubsection{difference outcome}

The default force law is simpler and cheaper to reason about while preserving
overlap-area reporting and comparison.  Existing stiffness values now act as
depth stiffness values, so the project reset the Python generation cfg
\texttt{collision\_stiffness\_q} values to a smaller baseline.

\subsubsection{<Analysis>}

Area remains useful as a conservative geometric diagnostic, but penetration
depth is the more direct instantaneous compression measure for the active
force calculation.  The selectable mode keeps the older area model available
without forcing diagnostics, plots, or historical reports to change names in
the same step.

\subsection{Boundary-Particle Reservoir Cell-Guard Margin}

\subsubsection{<previous form>}

The boundary-particle reservoir generator derived the cell-array side length
from the number of particles per release row and the configured particles per
cell row.  The physical pipe walls were then placed very near the cell-array
minimum and maximum bounds.  This worked with \texttt{wall\_guard}, because
particles were retired only by the selected wall/outlet rule, but it created a
failure mode for \texttt{cell\_guard}.

With \texttt{cell\_guard}, a particle approaching a boundary-particle pipe
wall could also enter the cell-array guard band.  The left/min side was
protected by the inlet exemption for vertical reservoir flow, but the right/max
side still used the general upper cell guard:
\begin{verbatim}
next_position.x >= float(WIDTH - 2u)
\end{verbatim}
for vertical flow.  Therefore particles near the right pipe wall could be
retired as escaped particles even though they were still inside the intended
physical pipe.

\subsubsection{<current form>}

The generator has been changed so boundary-particle reservoirs reserve an
internal guard margin between the physical pipe walls and the cell-array
guard.  For vertical boundary-particle flow, the physical pipe walls are moved
inward on \(x\); for horizontal boundary-particle flow, they are moved inward
on \(y\).  Mobile release positions are shifted by the same margin so new
particles are born inside the physical pipe, not inside the cell guard band.

During Vulkan validation, the generated shader files still showed the earlier
cell-array size and boundary values:
\begin{verbatim}
WIDTH = 27
BOUNDARY_XMAX = 26.00
\end{verbatim}
With these values, the max-side cell guard remains close to the old pipe wall.
The immediate next step is therefore to stop relying on a fully inferred side
length and manually set the cell-array side length for this reservoir family.
The generator should use that configured side length to place the physical
pipe walls with an explicit guard margin inside the cell-array bounds.

\subsubsection{difference outcome}

Partial solution.  The reason for the max-side retirement was identified, and
the generator now has the first margin-aware geometry rule.  The next
implementation step is to make cell-array side length manually selectable so
the physical pipe, guard band, and generated Vulkan constants agree.

\subsubsection{<Analysis>}

The important design distinction is that boundary particles define physical
wall response, while \texttt{cell\_guard} defines cell-array safety.  These are
not the same boundary.  When the physical wall is placed too close to the
cell-array guard, the guard becomes an accidental wall-death rule and prevents
the boundary-particle wall model from doing its work.

For boundary-particle reservoirs, the generated layout must satisfy:
\begin{verbatim}
cell-array min/max
    guard margin
        physical boundary-particle wall
            active pipe region
        physical boundary-particle wall
    guard margin
cell-array min/max
\end{verbatim}

The max-side failure is more visible than the min-side failure because the
vertical-flow inlet exemption intentionally skips the lower flow-axis guard so
particles can be born from the reservoir.  There is no equivalent exemption on
the non-flow max wall; instead the physical wall must be placed far enough
inside the cell array.  Manual side-length selection will make this explicit
and prevent generated \texttt{params.glsl}, \texttt{boundary.glsl}, and the
binary particle distribution from drifting out of agreement.

\subsection{CD Nozzle Reservoir Generator}

\subsubsection{<previous form>}

The project had straight boundary-particle reservoir generators for horizontal
and vertical pipe-wall tests.  The older MATLAB CD nozzle scripts generated a
boundary-particle nozzle by sweeping a piecewise radius profile along an axial
direction, but that geometry was not yet represented as a Python generator.
The MATLAB setup also encoded fixed section lengths directly in
\texttt{GetDevalleRadius.m}, so entry length, convergence length, throat
length, divergence length, and exit length were not separately configurable.

\subsubsection{<current form>}

\texttt{BoundaryCDNozzleReservoir.py} began as a boundary-only generator.  In
that first form, it did not generate mobile reservoir particles and did not
run birth/death lifecycle setup.  The generated particle records were boundary
markers:
\begin{verbatim}
ptype = 1.0
Data.w = 0.0
\end{verbatim}
and \texttt{dispatchx} remained one because there were no mobile source
particles to process.

The nozzle profile is parameterized by section lengths and radii:
\begin{verbatim}
dim
nozzle_inlet_length
nozzle_converge_length
nozzle_throat_length
nozzle_diverge_length
nozzle_exit_length
nozzle_inlet_radius
nozzle_throat_radius
nozzle_exit_radius
\end{verbatim}
The total axial length is calculated from the section lengths:
\[
L =
L_{\mathrm{inlet}}+
L_{\mathrm{converge}}+
L_{\mathrm{throat}}+
L_{\mathrm{diverge}}+
L_{\mathrm{exit}}.
\]

When \texttt{dim=2}, the generator writes the upper and lower nozzle profile
in a single \(z\) plane.  The 2D profile plane now defaults to \(z=2.0\), so
the nozzle lies in the same shallow layer used by the other two-dimensional
particle studies.  When \texttt{dim=3}, the generator writes circular boundary
marker rings along the axial direction.

The generator now also creates mobile reservoir particles.  Mobile particles
are written first with \texttt{ptype=0.0}; boundary marker particles are
written after the active particles with \texttt{ptype=1.0}.  Therefore
\texttt{dispatchx} is written as the number of active particles plus the null
particle, while \texttt{num\_particles} includes both active and boundary
particles.  The mobile spacing is:
\[
2r(1+\texttt{fraction\_of\_diameter\_separation})
\]
and the frame spacing is calculated from that distance, the timestep, and the
configured forward reservoir speed.

The nozzle entrance is anchored by a start-plane convention.  For horizontal
flow, \texttt{nozzle\_start\_x} defaults to \(1.0\); for vertical flow,
\texttt{nozzle\_start\_y} defaults to \(1.0\).  The process inlet/outlet
values written as \texttt{wallXMIN}/\texttt{wallXMAX} or
\texttt{wallYMIN}/\texttt{wallYMAX} now describe the reservoir lifecycle
planes, not the radius envelope around the nozzle centerline.  This prevents
Vulkan \texttt{StartReservoir} from moving born particles to the middle of the
cell array.

For visual validation, the generator writes:
\begin{verbatim}
vulkan/sim/BoundaryCDNozzleReservoir/CDNozzleGenerated.obj
\end{verbatim}
The OBJ is a triangle mesh generated from the same radius profile as the
boundary particles.  It is visualization-only; contact dynamics still use
boundary particle markers and \texttt{EvaluateWallSegment}.  The OBJ emits
\texttt{v}, \texttt{vt}, and \texttt{vn} records and faces in
\texttt{v/vt/vn} form so it is compatible with the current simple Vulkan OBJ
loader.

The CD nozzle wall response is now represented by:
\begin{verbatim}
boundary_particle_function = "cd_nozzle_wall";
\end{verbatim}
In Python, \texttt{EvaluateWallSegment(SourceID, BoundaryID)} dispatches to
\texttt{EvaluateCDNozzleWallSegment}.  Boundary markers provide broad-phase
locality and upper/lower wall identity, while the evaluator calculates the
actual wall point and wall normal from the analytic CD nozzle profile at the
source particle's current axial coordinate.  The Python direct-scan loop now
lets the selected evaluator decide whether a marker produces a valid segment,
instead of using the old flat \texttt{BoundaryParticleWallFlag} pre-filter.

The GLSL exporter now emits the corresponding CD nozzle functions:
\begin{verbatim}
BoundaryParticleCDNozzleWallFlag
CDNozzleRadius
CDNozzleRadiusSlope
EvaluateCDNozzleWallSegment
\end{verbatim}
and the generated \texttt{EvaluateWallSegment} dispatcher includes the
\texttt{cd\_nozzle\_wall} branch.  The CD nozzle constants are generated into
\texttt{boundary.glsl}.

\subsubsection{difference outcome}

Particle solution for geometry generation, reservoir staging, Python curved
wall response, and generated GLSL CD nozzle wall-function export.  Vulkan
runtime validation of the generated CD nozzle wall response remains the next
test step.

\subsubsection{<Analysis>}

The CD nozzle should not be treated as another horizontal or vertical pipe
case.  It is a boundary-particle geometry family with a radius function.  The
generator's first responsibility is to produce a trustworthy marker cloud; the
dynamics law should then receive those markers through the same
boundary-particle wall path used by straight walls.

The generated OBJ exposed one Vulkan rendering distinction: OBJ boundaries
are loaded as non-indexed vertex lists, while the old cube boundary uses an
indexed cube draw.  The command buffer now selects the draw method from
\texttt{boundary\_as\_obj}: OBJ boundaries use \texttt{vkCmdDraw}; cube
boundaries retain \texttt{vkCmdDrawIndexed}.  This preserves old cube and
sphere rendering while allowing generated nozzle OBJ visualization.

The current curve-aware wall function is selected by configuration:
\begin{verbatim}
boundary_particle_function = "cd_nozzle_wall";
\end{verbatim}
In 2D this function evaluates the local nozzle profile curve and returns the
wall normal, penetration/overlap geometry, wall flag, and validity.  In 3D the
corresponding radial wall evaluator from the nozzle centerline remains future
work.  The source-owned force accumulation rule does not change; only the
wall-segment evaluator changes.

\subsection{Reservoir Birth Admission Must Reject Starting Overlap}

\subsubsection{<previous form>}

Reservoir particles carry a scheduled frame in \texttt{Data.w}.  When the
frame arrives, Python and GLSL immediately place the particle at the inlet,
set \texttt{Data.w=0}, and process it as live.  If an earlier particle has
been pushed back into the inlet region, the new particle can therefore be born
inside that live particle.  Ordinary collision processing sees the overlap
only after admission and may report a starting collision or tunneling error.

\subsubsection{<current form>}

The approved next rule separates a birth attempt from successful admission.
\texttt{Data.w} remains the lifecycle authority, while reserved
\texttt{colFlg} values describe the attempt:

\begin{verbatim}
colFlg = 2  waiting
colFlg = 3  wants to become live
colFlg = 0  admitted live
\end{verbatim}

At its scheduled frame, a candidate is tested only against frame-start live
mobile targets.  Boundary markers, dead particles, unborn particles, and other
birth candidates do not block the test.  The contact query must return a
distinct \texttt{STARTING\_COLLISION} result so the wrapper can terminate the
remaining neighbor search.  A blocked candidate returns to \texttt{colFlg=2},
sets \texttt{Data.w=frameNum+1}, and receives no force or motion update.  An
unblocked candidate sets \texttt{Data.w=0} and \texttt{colFlg=0} before
entering ordinary dynamics.

Each blocked particle is delayed independently.  Later particles retain their
original scheduled frames, preserving the generated release schedule rather
than shifting an entire reservoir wave train.

\subsubsection{difference outcome}

No solution yet.  The lifecycle and contact-return contract is documented, but
the current Python and GLSL paths still activate particles immediately.  The
next work is to implement and test deferred birth admission.

\subsubsection{<Analysis>}

A boolean contact result can report overlap but cannot tell the wrapper whether
it found an ordinary live collision or a birth-blocking starting collision.
The distinct result is needed because the latter must end the source search and
skip all force and integration work for that frame.

The GPU implementation also has an ordering constraint.  If one candidate
writes itself live while another candidate is still scanning targets, results
can depend on invocation scheduling.  Admission must therefore read immutable
frame-start lifecycle state and commit successful births in a separate phase or
through an equivalent synchronized design.  This preserves source ownership:
each invocation decides and writes only its own lifecycle state.

\subsection{Boundary Marker Configuration Becomes Self-Checking}

\subsubsection{<previous form>}

Boundary-particle reservoir geometry, lifecycle planes, camera bounds, and
cell-domain size were partly inferred from overloaded \texttt{Wall*} values.
Boundary wall evaluation also depended on a scenario-wide wall-function
selection.  A configuration could place its retirement plane on or before the
terminal marker wall, or place valid marker geometry outside the visible plot,
without a focused generator error.

\subsubsection{<current form>}

The Python boundary-particle families now separate their responsibilities:

\begin{itemize}
    \item \texttt{reservoir\_inlet\_*} and
    \texttt{reservoir\_outlet\_*} own lifecycle planes;
    \item \texttt{x\_axis\_lims}, \texttt{y\_axis\_lims}, and
    \texttt{z\_axis\_lims} own display bounds;
    \item \texttt{wall\_boundary\_length} and nozzle profile fields own
    physical marker geometry;
    \item \texttt{side\_len} or \texttt{over\_ride\_side\_length} owns the
    cell domain;
    \item each marker carries its positive evaluator ID in
    \texttt{pdata.ptype}, which is copied into runtime \texttt{ptype} and
    \texttt{Data.z}; \texttt{pdata.temp\_vel} remains reserved.
\end{itemize}

The periodic reservoir generator now creates closed test walls without
duplicate corner markers and places the lifecycle outlet beyond the terminal
marker wall.  Both periodic-reservoir and CD-nozzle generators run a simulation
configuration check before particle generation.  Impossible geometry,
packing, lifecycle, or nozzle-profile combinations stop generation with an
aggregated error; display clipping is reported as a warning.

\subsubsection{difference outcome}

Particle solution.  The Python generation and dynamics paths have been
verified with self-describing boundary markers and dedicated configuration
roles.  Vulkan now transports and dispatches the marker-owned evaluator ID;
scenario runtime validation remains model-specific.

\subsubsection{<Analysis>}

Calculated geometry should be authoritative once generation begins, but manual
configuration remains useful for selecting the domain, lifecycle planes, and
population.  A preflight comparison is the correct boundary between those two
roles: it rejects contradictory inputs before they become difficult runtime
failures.  Marker-owned evaluator IDs also allow horizontal, vertical, and
curved wall segments to coexist without a global wall-function interpretation.

\subsection{Boundary Marker Dispatch And Closed Nozzle Inlet}

\subsubsection{<previous form>}

The Python marker path had become self-describing, but active documentation and
generated GLSL still described evaluator transport through
\texttt{pdata.temp\_vel} or a global \texttt{WALL\_FUNC}.  The vertical
periodic generator retained approximate marker offsets rather than exact wall
bounds.  The CD nozzle emitted only upper and lower profile markers, leaving
the inlet open.  Its generated test file also wrote
\texttt{num\_particle\_colliding} twice; Python accepted the duplicate while
the C++ configuration parser rejected it.

\subsubsection{<current form>}

Boundary evaluator identity is now carried by \texttt{pdata.ptype} and copied
to runtime \texttt{ptype} and \texttt{Data.z}.  Generated GLSL dispatches
\texttt{EvaluateWallSegment} from \texttt{P[BoundaryID].Data.z}: 1 is
horizontal, 2 is vertical, and 3 is CD nozzle.  The global
\texttt{WALL\_FUNC} dispatcher has been removed; the CD wrapper uses only an
optional-module availability define.

Horizontal and vertical periodic reservoirs now generate exact closed marker
walls without duplicate corners.  Their mobile rows were tested at maximum
configured position jitter and produced no starting wall contact.  The 2D CD
nozzle now adds a vertical-evaluator inlet cap, shares its two corners with the
CD profile markers, leaves the outlet open, and preserves the 1.1-diameter
runtime birth offset.  The generated CD test writer emits each setting once.

The periodic vertex shader no longer suppresses all live vertices on frame
zero.  Unborn and dead particles are still rejected by the lifecycle test,
while boundary markers render and populate occupancy immediately.

\subsubsection{difference outcome}

Particle solution.  Python worst-case-jitter checks pass for horizontal
periodic, vertical periodic, and CD-nozzle births.  The periodic Vulkan shader
set compiles with marker-owned dispatch.  CD runtime wall response remains the
next focused Vulkan validation.

\subsubsection{<Analysis>}

Marker type and marker evaluator are the same immutable classification, so one
positive \texttt{ptype} value should own both roles.  Reserving
\texttt{temp\_vel} avoids consuming a field needed by later dynamics work.
Closed inlet geometry must be represented by its own straight-wall evaluator;
the curved CD evaluator remains responsible only for the upper and lower
analytic profile.  Exact shared corner coordinates prevent duplicate occupancy
without weakening either wall rule.

Python and generated GLSL now use the same quadratic converging and diverging
radius functions and analytic slopes.  The regenerated CD shader set compiles
with the quadratic profile.

\subsection{Adjacent Markers Can Represent One Physical Wall}

\subsubsection{<previous form>}

Boundary-marker duplicate suppression used marker \texttt{TargetID}.  This
prevents processing one marker repeatedly when several source corners discover
the same occupancy record, but it does not recognize two different markers as
representations of one physical wall.

At a rounded CD-nozzle profile step, the ending marker of one horizontal row
and the starting marker of the adjacent row can occupy neighboring cells.  A
mobile particle whose extent spans both cells can discover both target IDs.
Both markers then evaluate the same analytic upper or lower wall at the source
axial coordinate, so target-ID-only suppression can accumulate the same wall
force twice.

\subsubsection{<current form>}

Python evaluates the candidate segment and records its physical wall flag in
\texttt{processed\_boundary\_walls}.  The periodic and CD Vulkan wrappers now
use the equivalent source-local wall bit mask.  Only the first valid segment
for each wall flag contributes force during a frame; marker target IDs are not
inserted into the ordinary particle duplicate list.

\subsubsection{difference outcome}

Full solution for the current Python, periodic Vulkan, and CD Vulkan wrappers.

\subsubsection{<Analysis>}

Marker identity belongs to the broad phase, while wall flag identifies the
physical response.  The GLSL wrappers maintain a source-local processed-wall
mask keyed by \texttt{BoundaryWallSegment.wallFlag}.  They evaluate a candidate
first, accumulate only the first valid segment for that flag, and then set the
flag bit.  Ordinary mobile-particle duplicate suppression remains target-ID
based.

\subsection{Capture Diagnostics Expose Artificial Inlet Compression}

\subsubsection{<previous form>}

The Vulkan particle structure could be captured, but reservoir behavior was
judged mainly from rendered motion.  Scheduled waves were admitted when their
stored frame arrived.  Release spacing assumed that the preceding wave
continued at its configured inlet speed and cleared the birth position.

\subsubsection{<current form>}

The Python capture reader now pairs an original particle binary with adjacent
\texttt{<name>\_<frame>.cap} files.  It writes \texttt{<name>.csv} beside the
original binary and reports lifecycle counts, momentum, kinetic energy,
same-cohort initial baselines, and reverse-flow momentum and kinetic energy.
Reverse flow is selected by a negative velocity projection onto the initial
flow direction.

Frame-by-frame captures isolated one inlet lane.  At frame 5722 an older
particle remained at $x=1.16888058$ when a new particle was admitted at
$x=1.22000003$.  For radius $0.1$, their center spacing was only
$0.05111945$, producing artificial starting penetration of $0.14888055$.
Across frames 5722--5725, pair momentum remained approximately $0.02449093$,
while pair kinetic energy rose from approximately $0.00021008$ to
$0.00118541$.  The older particle reversed and the new particle accelerated
downstream, producing the observed counterflow line.

The selected next model removes scheduled runtime birth.  A finite reservoir
will be prefilled with nonoverlapping particles.  At release, the particles
and rear wall begin moving together; the rear wall continues at inlet velocity
as a piston.  The long reservoir will use a rectangular cell domain to avoid
the memory cost of a cubic array.

\subsubsection{difference outcome}

Full diagnostic explanation of the counterflow source.  The staged-birth
model is not repaired; it is superseded as the next development direction by
the prefilled moving-piston reservoir.

\subsubsection{<Analysis>}

Scheduled frame separation is only a prediction based on unconstrained inlet
speed.  It cannot guarantee clearance once downstream contacts slow or return
a particle to the inlet.  Materializing a new particle inside that live
particle creates compression without a causal mechanical source.  Momentum
can remain pair-consistent while kinetic energy grows because the initial
overlap supplies artificial contact potential.  A prefilled reservoir removes
that discontinuity.  Its piston is an explicit external momentum source, so
future diagnostics must include piston impulse rather than treating particle
momentum as a closed-system quantity.

\subsection{Horizontal Prefilled Reservoir Reset}

\subsubsection{<previous form>}

\texttt{BoundaryParticleReservoirPeriodic.py} combined horizontal and vertical
generation, scheduled birth, retirement, wall selection, packing, and test-file
generation.  Repeated release columns occupied the same stored positions and
depended on \texttt{Data.w} birth frames to avoid simultaneous overlap.  The
configuration exposed lifecycle and implementation selectors that obscured
the physical geometry.

\subsubsection{<current form>}

The periodic generator and its cfg have been retired.  Development restarted
with \texttt{BoundaryParticleReservoirHorizontal.py} and
\texttt{BoundaryParticleReservoirHorizontal.cfg}.  The generator owns its
horizontal orientation.  Physical geometry is declared by six-value
\texttt{reservoir\_bounds} and \texttt{pipe\_bounds} arrays inside an explicit
rectangular cell domain.

The current two-dimensional convention uses $z_{front}=0$, $z_{back}=3$,
and particle plane $z=1$.  Validation checks bound ordering, cell-domain
containment, the shared reservoir/pipe interface, and containment of the pipe
opening within the reservoir face.

Packing is currently diagnostic only.  Radius $0.1$, absolute surface gap
$0.1$, and wall-contact clearance produce a centered $30\times13$ capacity
report for the reduced Python geometry, or 390 possible mobile particles.
No mobile records are emitted yet.

The generator creates a closed reservoir inlet, reservoir and pipe side walls,
optional interface shoulders, and an open pipe outlet.  Shared marker
coordinates are emitted once.  The current reduced geometry contains 43
boundary markers plus the null record.  The milestone writes only the binary
particle file and deliberately does not write test metadata.

\subsubsection{difference outcome}

Particle solution for validated horizontal geometry, packing capacity, and
boundary-only binary generation.  Mobile packing and piston motion remain
unimplemented.

\subsubsection{<Analysis>}

Separating physical bounds from particle packing removes lifecycle timing from
geometry generation.  The boundary-only binary provides a narrow test surface:
wall topology and evaluator IDs can be inspected before mobile particles,
piston state, GLSL generation, or Vulkan configuration are reintroduced.

\subsection{Vulkan Immutable Velocity And Permanent Death Bounds}

\subsubsection{<previous form>}

Vulkan positions were double buffered, but all invocations read and wrote one
\texttt{VelRad} field.  A source could therefore observe a target velocity
already updated by another invocation in the same dispatch.  The replacement
horizontal reservoir also had configured death bounds but no permanent
post-integration death path.

\subsubsection{<current form>}

The particle ABI now contains \texttt{VelRadA} and \texttt{VelRadB}.  They use
the same selector and direction as \texttt{PosLocA} and \texttt{PosLocB}.
Generated dynamics read all relative motion from the current velocity buffer,
write the candidate to the alternate velocity buffer, apply maximum-depth
projection there, and integrate position with that accepted alternate
velocity.

\texttt{BoundaryParticleReservoirHorizontalP.comp} runs the generated
source-owned dynamics and marker-wall path successfully.  After
\texttt{CalcPosition}, generated \texttt{ApplyParticleDeathBounds} tests the
new position against six cfg-generated planes.  A mobile particle crossing any
plane receives \texttt{Data.w=-1}; dead particles are excluded from dynamics,
contacts, occupancy, and rendering.  Boundary markers remain permanent.
Capture parsing now matches the expanded ABI and selects velocity from the
buffer paired with the active position.

\subsubsection{difference outcome}

full solution

\subsubsection{<Analysis>}

Velocity double buffering restores the immutable frame-start rule required by
source-owned execution without cross-workgroup barriers.  Permanent death is
separate from the retired scheduled-birth lifecycle: particles begin alive,
may die once after leaving the configured domain, and never re-enter the
simulation.

\subsection{Analytic Moving Piston In Python}

\subsubsection{<previous form>}

The new horizontal generator initially closed the reservoir inlet with a row
of stationary vertical boundary markers.  Wall contacts therefore had zero
target velocity, and a moving piston would have required marker motion,
cell-occupancy updates, and duplicate-contact control.

\subsubsection{<current form>}

The fixed inlet marker row has been removed.  Python evaluates one analytic
vertical piston plane per active mobile source.  The plane remains at the
reservoir start until \texttt{piston\_start\_frame}, moves with
\texttt{reservoir\_velocity}, and clamps at the reservoir end.

Piston velocity is carried by its wall contact.  Relative-normal diagnostics,
the penetration-step guard, internal-momentum compression and release, and
maximum-depth containment all use that target velocity.  At maximum depth the
source is constrained to the piston normal velocity, not zero.  Python reports
current piston state, contact count, frame and cumulative impulse, and
cumulative vector momentum transfer.  The game UI renders the analytic plane
as a vertical line.

\subsubsection{difference outcome}

particle solution

\subsubsection{<Analysis>}

A flat moving piston is exactly described by one position and velocity and
does not need particle occupancy to define its geometry.  Analytic evaluation
preserves source ownership, avoids duplicate marker forces, and makes the
external momentum source explicit.  GLSL/Vulkan piston integration and a
heterogeneous linear converging outlet section remain future work.

\subsection{General Finite-Line Chamber And GLSL Export}

\subsubsection{<previous form>}

The replacement reservoir still separated \texttt{reservoir\_bounds} from
\texttt{pipe\_bounds}, used horizontal and vertical evaluator IDs, and stopped
at the first valid marker for a wall flag.  The piston existed only in Python.

\subsubsection{<current form>}

The horizontal generator now packs one \texttt{chamber\_bounds} domain and
describes stationary straight geometry with
\texttt{linear\_chamber\_segments} and \texttt{linear\_wall\_segments}.  Each
finite segment stores \(A,B,C\), endpoints, and a wall flag and uses evaluator
ID 4.  Python generation and simulation have been verified.  Contact selection
retains the deepest valid candidate for each physical wall flag.

The exporter now generates
\texttt{ForceDynamicsLinearReservoir.glsl}, including general finite-line
evaluation, piston position and velocity, piston collision processing, and
moving-target maximum-depth constraints.  Scenario include wiring and shader
compilation remain to be validated.  CD nozzle work is the next model step.

\subsubsection{difference outcome}

particle solution

\subsubsection{<Analysis>}

One finite-line equation handles horizontal, vertical, and sloped straight
walls without deriving response geometry from marker placement.  Separate
segment collections retain configuration clarity while sharing one collision
implementation.  Deepest-contact selection removes scan-order dependence at
segment transitions, and explicit target velocity preserves the moving-piston
maximum-depth rule in generated GLSL.

\subsection{Piecewise Parametric CD Nozzle And Vulkan Integration}

\subsubsection{<previous form>}

The CD nozzle used section-specific inlet, converging, throat, diverging, and
exit parameters together with a dedicated radius-and-slope evaluator.  Its
visual OBJ had been generated from older coordinates, so the displayed profile
could differ in position and aspect ratio from the collision profile.  Piston,
linear-wall, and CD helpers were also coupled in generated includes.

\subsubsection{<current form>}

The CD generator now validates an ordered collection of finite cubic
\texttt{curve\_wall\_segments}.  Shared parametric math evaluates points,
tangents, extrema, bounds, minimum wall gap, and closest points.  Boundary
markers use evaluator ID 5 and provide one locality marker per cell per
physical wall.  The wall response derives the closest curve point and normal
from segment coefficients rather than marker orientation.

The GLSL exporter emits independent
\texttt{ForceDynamicsPiston.glsl} and
\texttt{ForceDynamicsParametricCurve.glsl} modules.  The Vulkan CD wrapper
uses double-buffered velocity, skips dead sources and targets, applies the
analytic piston, and checks six-plane death bounds after position integration.
The vertex pass excludes dead particles from occupancy and rendering.

The generator now rebuilds \texttt{CDNozzleGenerated.obj} from the same
segment sample parameters used for boundary-marker generation.  The OBJ is a
triangle ribbon with the texture-coordinate and normal records required by the
current loader; it is visualization only.

\subsubsection{difference outcome}

full solution

\subsubsection{<Analysis>}

Piecewise parametric geometry removes nozzle-section decision trees and makes
straight, quadratic, and cubic walls instances of one representation.  Sharing
segment coefficients across validation, Python contact response, GLSL contact
response, marker placement, and visualization removes duplicated profile
state.  Keeping piston and curve support in separate generated modules lets a
scenario select only the geometry and moving-boundary behavior it requires.

\subsection{Starting-Contact Effective-Radius Tunneling Defect}

\subsubsection{<previous form>}

Ordinary particle contact originally used the physical source and target
radii directly.  Contact existed whenever the physical center distance
satisfied
\begin{equation}
d_{ij}<r_i+r_j,
\end{equation}
and the physical penetration depth was
\begin{equation}
\delta_{ij}=r_i+r_j-d_{ij}.
\end{equation}
There was no interval in which physically overlapping particles were hidden
from contact force or tunneling detection.

Commit \texttt{20b4881}, dated June 18, 2026 at 11:23 AM, introduced special
starting-contact geometry.  Its commit title was \emph{update git ignore}, but
the commit also added approximately 300 lines to
\texttt{ForceDynamicsBase.py}.  The added model scaled both radii by
\begin{equation}
s=\frac{d_{ij}(0)}{r_i+r_j}
\end{equation}
and used
\begin{equation}
r_i^{\mathrm{eff}}=s r_i,
\qquad
r_j^{\mathrm{eff}}=s r_j
\end{equation}
so an initially overlapping pair appeared to have zero effective
penetration.  Commit \texttt{c41ded4}, later on June 18, moved this logic into
\texttt{ForceDynamics.py}; it did not originate the model.

The intent was to avoid an immediate force discontinuity when a configured
initial state already contained overlap.  The contact could compress relative
to its initial separation and rebound to that separation.  Once it returned,
the state became \texttt{resolved}.  Force geometry was then suppressed until
the physical center distance reached the full physical separation
$(r_i+r_j)$.

\subsubsection{<current form>}

\texttt{TwoParticleHorizontalBug001.cfg} provides a narrow Python
reproduction.  Two radius-(0.25) particles begin near the right wall with a
small vertical offset.  The second particle is compressed between the first
particle and the wall.  After the initial pair response and later wall
rebounds, the pair remains physically overlapped while its starting-contact
state is resolved and suppressed.

The frame reports capture every particle and momentum frame from 0 through
400.  The first reported pair phase ends at frame 25 when the effective
overlap returns to zero.  The physical particles do not separate to their
diameter of (0.5); their center distance remains near (0.42) for much of
the subsequent motion.  Because the state remains suppressed, the later
approach toward the left wall produces no valid force geometry.

The center ordering reverses between frames 381 and 382:
\begin{align}
x_2-x_1 &= 0.01206345 && \text{at frame 381},\\
x_2-x_1 &=-0.00229142 && \text{at frame 382}.
\end{align}
The particles therefore pass through one another while still listed as a
physically overlapping pair.  Before the reporting correction,
\texttt{contacts.csv} represented these suppressed detections as active
contact rows whose normal, distance, depth, and force were all zero.  It also
wrote empty rows for every boundary marker, obscuring the two mobile-particle
history.

Python tunneling diagnostics now test the physical proximity
\begin{equation}
p_{ij}^{\mathrm{phys}}=d_{ij}-r_j
\end{equation}
before effective-radius or starting-contact logic.  A negative physical
proximity reports \texttt{ERROR\_PARTICLE\_TUNNELING} with source and target
IDs.  This diagnostic change does not alter force, velocity, position, or
internal momentum.

Reporting now distinguishes valid force contacts from physical detections
whose geometry is suppressed.  Valid rows use
\texttt{contact\_status=active} and \texttt{geometry\_valid=1}; suppressed
rows use \texttt{contact\_status=suppressed} and
\texttt{geometry\_valid=0}.  Null, dead, and boundary-marker particles are
excluded from mobile contact reports.  Momentum reports include the runtime
error code, description, source, target, and wall.  Every new simulation run
clears all files directly inside its configured report directory before
writing new output.

\subsubsection{difference outcome}

particle solution

\subsubsection{<Analysis>}

The defect is not primarily a tolerance, force-stiffness, or timestep problem.
It is a violation of the contact-geometry invariant.  The effective-radius
model creates the interval
\begin{equation}
d_{ij}(0)<d_{ij}<r_i+r_j
\end{equation}
in which the real particles overlap but the force model treats the contact as
open.  If relative motion reverses before complete physical separation, the
particles can recompress and exchange ordering without force, maximum-depth
containment, or the former effective-radius tunneling check.  This is a
history-dependent transparent-overlap region.

The model replacement must preserve physical radii for all geometry.  Physical
overlap must always create physical contact, and contact may end only at
$(d_{ij}\ge r_i+r_j)$.  Starting overlap must instead initialize local
compression state and its corresponding potential-energy baseline.  Contact
history may control stored momentum, phase, and constitutive response, but it
must never resize particles or suppress physically valid geometry.  The same
rule applies to starting wall penetration.

The reporting and physical tunneling checks provide a verified diagnostic
surface, but the dynamics correction remains intentionally pending.  The next
implementation step is to remove effective-radius scaling and resolved-contact
suppression, initialize starting compression from physical penetration, and
rerun \texttt{TwoParticleHorizontalBug001} as the regression case.

\subsection{Generic Parametric Boundary Migration}

\subsubsection{previous form}

Python wall processing retained several overlapping representations: manual
\texttt{WallXMIN/WallXMAX/WallYMIN/WallYMAX} planes, horizontal and vertical
marker evaluators, an analytic CD-nozzle evaluator, finite linear segments,
and cubic parametric segments.  Runtime selection depended on
\texttt{walls\_on}, marker evaluator IDs 1--5, and a fallback manual wall scan.
At shared marker cells, evaluator ID 5 did not identify which physical segment
the marker represented, so nearest-segment ties could select the wrong wall.

The first generic cfg migration also exposed a data-source distinction.
\texttt{pdata\_from\_file=false} intentionally loads mobile particles from
\texttt{PARTICLE\_DATA}; it does not load generated binary boundary records.
Requiring binary input would have repaired missing markers by changing the cfg
contract rather than repairing wall construction.

\subsubsection{current form}

All cases in \texttt{batch\_test\_001.cfg} now define stationary walls with
\texttt{curve\_wall\_segments}.  Straight walls are cubic records whose
quadratic and cubic coefficients are zero.  Every boundary marker uses
\texttt{ptype=1} and serves only as a locality record.  Wall evaluator IDs are
not stored in \texttt{Data.z} in the generic path.  Once a local marker is
found, Python evaluates every configured segment directly, retains the deepest
valid contact for each positive physical \texttt{wall\_flag}, and accumulates
that wall force once.

The manual wall scan, \texttt{walls\_on}, \texttt{Wall*} geometry, evaluator
IDs 1--4, and their Python dispatcher functions have been removed from the
active Force model.  \texttt{pdata\_from\_file} keeps its original meaning:
true loads the complete generated binary, while false loads mobile cfg records
and constructs generic boundary markers in memory.  Generator and runtime marker
construction share the same sampling functions in
\texttt{ParametricCurve.py}.

The Python viewer draws boundary markers at their configured radius as unfilled
green circles.  They receive no sphere shading, collision lens, or particle
number label, while point mode remains unchanged.

\subsubsection{difference outcome}

full solution

\subsubsection{Analysis}

Wall geometry now has one source of truth.  Boundary particles answer only the
broad-phase question ``is wall evaluation local?''; the parametric segment
answers the geometric questions of closest point, tangent, outward normal,
penetration, and physical identity.  This separation avoids both marker-order
dependence and duplicated wall equations.  Reservoir generators can extend
\texttt{GenericGenData} for population and piston behavior without creating a
second stationary-wall model.

\subsection{Simple GLSL Restart}

\subsubsection{previous form}

The GLSL port accumulated several generations of wall and reservoir behavior
inside one family of includes and scenario wrappers.  The active files mixed
manual simple walls, pipe walls, boundary-particle walls, analytic CD-nozzle
walls, finite-line reservoir walls, piston helpers, death rules, and old
scheduled-birth lifecycle helpers.  That made it difficult to verify the
small particle cases from \texttt{batch\_test\_001.cfg} because every compile
or runtime failure had to be separated from reservoir and CD-nozzle history.

\subsubsection{current form}

The simple Vulkan target is now
\texttt{vulkan/sim/TestPythonBoundarySpheres}.  A new exporter,
\texttt{tools/ExportForceDynamicsSimpleGLSL.py}, generates a separate simple
GLSL family:
\begin{itemize}
    \item \texttt{ForceDynamicsSimple.glsl};
    \item \texttt{ForceDynamicsSimpleBoundaryParticle.glsl};
    \item \texttt{ForceDynamicsSimpleFunctionWall.glsl};
    \item \texttt{TestPythonBoundarySpheresP.comp}.
\end{itemize}

The generated compute schedule preserves the old cell/corner broad phase
rather than scanning every particle.  Each source scans its eight occupied
corner cells, deduplicates target IDs found through multiple cells, processes
mobile particle contacts, and treats boundary-marker targets as locality for
\texttt{curve\_wall\_segments}.  It skips null, boundary, and dead particles
as sources and applies death bounds after position integration.  The simple
\texttt{.tst} contract no longer emits derived
\texttt{boundary\_x/y/z\_min/max} fields; wall geometry comes only from
\texttt{curve\_wall\_segments}.

\subsubsection{difference outcome}

particle solution

\subsubsection{Analysis}

The restart removes the decision tree from the low-particle validation path.
The simple path can now be compared against Python without reservoir birth,
piston motion, CD geometry, finite-line compatibility, or legacy manual-wall
fallbacks.  Reservoir and piston work can resume later by building on the
same generic wall contract after the simple Python and GLSL paths agree.

\subsection{Reservoir Restart On The Simple Contract}

\subsubsection{previous form}

The earlier reservoir path mixed several concerns: scheduled particle birth,
outlet retirement, finite-line wall arrays, boundary-marker evaluator IDs,
and piston behavior.  This made it difficult to tell whether a failure came
from reservoir lifecycle, wall geometry, broad-phase locality, or the core
particle dynamics.

During the first simple-GLSL restart, reservoir and piston support were
intentionally deferred.  Death bounds were also guarded in the generated
simple GLSL by \texttt{DEATH\_BOUNDS\_AVAILABLE}.  That guard made death
planes optional even though the new generic \texttt{.tst} contract always
emits six death-plane constants.

\subsubsection{current form}

\texttt{GenReservoir.py} now inherits \texttt{GenericGenData.py}.  It keeps
the generic wall contract and adds only reservoir-specific generation:
packed mobile particles, piston/chamber metadata, and optional reuse of
\texttt{piston\_velocity} as the initial reservoir particle velocity.  Mobile
particles are active at frame zero.  Boundary particles remain
\texttt{ptype=1} locality markers for \texttt{curve\_wall\_segments}.  The
reservoir does not reintroduce scheduled birth, outlet retirement, manual
walls, or evaluator IDs in \texttt{Data.z}.

The simple GLSL exporter now emits
\texttt{ForceDynamicsSimplePiston.glsl} as an optional reservoir include.
The base simple particle/wall path remains unchanged, while reservoir wrappers
can include piston contact processing without using legacy finite-line or
birth-lifecycle helpers.

Death bounds are now unconditional in
\texttt{ForceDynamicsSimple.glsl}.  A Vulkan capture showed particle 72
reaching the cell-array boundary at frame 544.  The frame matched the
kinematic calculation from the generated state:
\[
(33.0-19.4)/(0.05\cdot0.5)=544.
\]
Because \texttt{death\_x\_max=30} should have killed the particle earlier,
the captured \texttt{ERROR\_PARTICLE\_OUT\_OF\_BOUNDS} exposed the optional
death-bound guard as an incorrect contract.

\subsubsection{difference outcome}

particle solution

\subsubsection{Analysis}

The reservoir is now an extension of the simple model rather than a competing
model.  The base dynamics remain source-owned particle and parametric-wall
processing.  Piston contact is optional wrapper behavior with explicit
metadata, and death planes are mandatory lifecycle rules for every simple
generated scenario.  This keeps the next validation target narrow: verify the
simple GLSL particle/wall path, then verify the reservoir wrapper by adding
only piston contact and packed initial conditions.

\subsection{Dead Particles In Vertex Broad Phase}

\subsubsection{previous form}

The simple reservoir compute path marked mobile particles dead by writing
\texttt{Data.w < 0} after they crossed a configured death plane.  Compute
source processing and contact-target processing skipped dead particles, but
the \texttt{TestReservoir} vertex shader still advanced far enough to color
and insert those records into the cell-occupancy corner lists.

This created visual noise after particles crossed the death zone and could
also leave dead records visible to the next broad-phase build.  The problem
was not a new particle type: \texttt{ptype} still distinguished null,
mobile, and boundary records.  The missing rule was lifecycle filtering in
the vertex path.

\subsubsection{current form}

\texttt{TestReservoirP.vert} now clears the particle's corner-list entries and
then returns before rendering or occupancy insertion when
\texttt{Data.w < 0}.  Dead particles receive zero point size and neutral
output color, and they do not contribute cell locations for later compute
passes.

The active particle categories remain:
\begin{itemize}
    \item \texttt{ptype=-1}: null particle;
    \item \texttt{ptype=0}: mobile particle;
    \item \texttt{ptype=1}: boundary marker.
\end{itemize}
Dead state is orthogonal to \texttt{ptype}; it is carried by negative
\texttt{Data.w}.

\subsubsection{difference outcome}

full solution

\subsubsection{Analysis}

The fix keeps lifecycle ownership consistent across compute and graphics.
Compute death bounds decide when a mobile particle becomes dead, while the
vertex stage must honor that state before doing either visual work or
broad-phase occupancy work.  This prevents dead records from being mistaken
for active contact participants and removes stale color artifacts from the
reservoir visualization.

Error diagnostics were also separated from particle-count accounting by using
a dedicated \texttt{ParticleNumber} field for the failing source ID.  That
keeps \texttt{numParticles} available for its count/debug role and makes
source-owned runtime errors readable in Vulkan logs.

\subsection{Function-Wall Boundary Model Direction}

\subsubsection{previous form}

The generic wall path moved toward fully parametric curve segments using
\texttt{x(t)} and \texttt{y(t)}.  Runtime contact used a bounded closest-point
search: sample candidate \texttt{t} values, choose the nearest sample, refine
with Newton iteration, clamp \texttt{t} to \([0,1]\), and then calculate the
wall tangent and normal at that closest point.

That representation is powerful, but it is broader than the current model
needs.  The active two-dimensional conduit walls are intended to be smooth,
single-valued functions of position, not looping or self-overlapping curves.
The boundary particle is also not meant to be the wall geometry itself.  Its
job is to mark that a cell contains a wall so the mobile particle can be sent
to the wall evaluator.  The evaluator should then use the mobile particle's
\(x,y\) position and the wall definition to calculate the physical response.

The leakage seen near the start of the converging section exposed this
confusion.  Small particles could enter cells near the wall transition without
getting a reliable wall response.  Treating marker centers as geometry, or
making marker rasterization increasingly dense, does not solve the underlying
model issue.  It only changes which cells happen to trigger the evaluator.

\subsubsection{current form}

The next boundary model should use function walls for the active Python and
simple GLSL work.  A wall segment is evaluated directly from the mobile
particle position:
\[
    y_{wall} = f(x_{particle}).
\]
For cubic function segments this can be represented as
\[
    y = y_0 + a_1 dx + a_2 dx^2 + a_3 dx^3,
    \qquad dx = x - x_0.
\]
The local slope is
\[
    \frac{dy}{dx} = a_1 + 2a_2 dx + 3a_3 dx^2.
\]
The wall definition, or the generator, must provide the outward normal
orientation for the segment.  Runtime code should not need separate lower-wall
and upper-wall branches.  Once the outward normal is known, every wall uses the
same signed-distance calculation:
\[
    d_{out} = (p - p_{wall}) \cdot n_{out},
\]
where \(p\) is the particle center, \(p_{wall}\) is the point on the wall at
that particle position, and \(n_{out}\) is the outward wall normal.  The
penetration depth is then
\[
    \delta = r + d_{out}.
\]
A contact exists when \(\delta > 0\).  The response force acts inward, opposite
the outward normal.

Vertical walls are the only required special case.  They can be represented as
\(x = c\) with a configured outward normal.  They still use the same signed
normal penetration formula after selecting the wall point
\((c, y_{particle})\).

Boundary markers remain broad-phase sentinels.  A marker says that the cell is
near a wall; it does not define the wall's position, tangent, normal, or
penetration depth.  The physical geometry comes from the function-wall segment
and its normal orientation.

\subsubsection{difference outcome}

particle solution

\subsubsection{Analysis}

The cleanup should happen in this order.

First, clean up the active model before adding more behavior:
\begin{itemize}
    \item keep the leakage-response experiments backed out of the active path;
    \item retire the general closest-point/Newton parametric evaluator from the
          simple and reservoir direction once the function-wall evaluator is in
          place;
    \item settle on one wall configuration shape for the active path instead of
          maintaining both generic parametric curves and function walls;
    \item keep Vulkan unchanged until the Python behavior is validated.
\end{itemize}

Next, make the Python changes:
\begin{itemize}
    \item update the configuration contract to describe function-wall segments
          and vertical wall segments explicitly;
    \item update the generator so boundary particles mark wall-containing cells
          as locality sentinels, not as sampled wall geometry;
    \item implement the Python wall evaluator from particle \(x,y\), wall
          function, slope, and outward normal;
    \item validate starting geometry, small-particle wall contact, and the
          converging-section transition in Python before exporting anything to
          Vulkan.
\end{itemize}

Only after Python passes should the Vulkan path be changed:
\begin{itemize}
    \item update the simple GLSL exporter/templates for function-wall metadata
          and evaluation;
    \item regenerate the simple common GLSL and the active test wrapper;
    \item run Vulkan captures around the converging transition and compare the
          wall-contact behavior with Python;
    \item then resume reservoir and piston work on top of the verified simple
          boundary model.
\end{itemize}

This keeps tomorrow's work focused: clean the model representation first,
prove the Python wall response, then port the verified behavior to GLSL.

\subsection{Compressor Piston Bring-Up}

\subsubsection{previous form}

The reservoir piston was present in generated metadata, but Python runtime
configuration did not consistently see the same piston model.  Source cfg
files provided \texttt{piston\_velocity} and
\texttt{piston\_start\_frame}; generated test files provided
\texttt{piston\_velocity\_x/y/z}, \texttt{piston\_x\_start}, and
\texttt{piston\_x\_stop}.  The runtime dynamics required
\texttt{piston\_enabled}, \texttt{piston\_velocity},
\texttt{piston\_x\_start}, \texttt{piston\_x\_stop}, and
\texttt{piston\_start\_frame}.  Running directly from the source cfg could
therefore report the piston as disabled or place it at zero with zero
velocity.

The first piston contact geometry also used an offset ghost-particle
construction.  That made the visual piston line and the physical contact
plane disagree.  A particle could visually reach the piston before the
dynamics treated the piston plane as physically present.

\subsubsection{current form}

Runtime configuration now normalizes both source cfg and generated test-file
forms.  When scalar generated fields are present,
\texttt{piston\_velocity\_x/y/z} are converted into the vector
\texttt{piston\_velocity}.  When a source cfg provides
\texttt{piston\_velocity} but omits \texttt{piston\_enabled}, the piston is
enabled.  When source cfg files omit derived piston bounds, the runtime derives
\texttt{piston\_x\_start} and \texttt{piston\_x\_stop} from the reservoir
boundary curves and \texttt{piston\_start\_offset}, matching the generator.

The Python piston contact is now an analytic plane contact at the displayed
piston position.  Penetration is calculated from the particle left edge and
the piston plane:
\[
    \delta = r - (x_{particle}-x_{piston}).
\]
A contact exists when \(\delta>0\).  This makes the white piston line,
contact geometry, and moving-wall velocity refer to the same physical plane.

The Python viewer now includes piston diagnostics.  With dynamics diagnostics
enabled, the UI prints piston position and velocity near the first mobile
particle.  The viewer also supports a particle-diameter grid so rendered
radius, world scale, and dynamics radius can be checked visually.

\subsubsection{difference outcome}

particle solution

\subsubsection{Analysis}

The closed-end \texttt{TestCompressor} case is now a useful compression test.
With particle initial velocity set to zero, the moving piston drives a packed
particle block into the closed downstream cap.  The first observed
\texttt{ERROR\_MAX\_DEPTH\_CONSTRAINT} with
\texttt{source=1, first\_wall=1} identifies the piston wall, not a stationary
pipe wall.  The particle-particle packing did not fail first; the rigid
piston-particle contact reached the hard compression limit first.

That result is physically meaningful for the current model.  The piston is an
infinitely rigid prescribed-motion boundary.  If it continues to advance after
the packed particle block cannot accept more compression, the piston contact
exceeds the configured hard penetration depth.  In this sense the wall has
``hydraulically fractured'' before the mobile-particle chain exceeded its own
compression limit.

The next diagnostics for this path should report the full source contact set
at failure: target IDs, wall flags, penetration depths, target depth, hard
depth, candidate velocity, contained velocity, and blocked momentum.  That
will separate expected compressor overpressure from mistakes in the
source-owned constraint projection.

\subsection{Compressor Overcompression Feasibility}

\subsubsection{previous form}

The compressor test could report
\texttt{ERROR\_MAX\_DEPTH\_CONSTRAINT} or
\texttt{ERROR\_PENETRATION\_STEP\_TOO\_LARGE} after many particles had
entered simultaneous contact from the piston to the closed pipe exit.  The
first failing particle was not necessarily at the piston, which made the
failure look like a random downstream shader or projection error.  The
generator reported kinematic contact timing, but it did not directly state
how much of the available compression depth a single timestep could consume.

The GLSL error detail also used the same subcode for every failed
maximum-depth projection.  A capture could identify the source particle, but
not whether the failed projection happened immediately after the force-driven
velocity or after the later post-resolution remaining-depth constraints were
registered.

\subsubsection{current form}

The compressor test is now interpreted as a dense-contact feasibility test.
When nearly every particle is in contact, the piston is the load source but
the first impossible local constraint may appear anywhere in the packed
column.  Successful runs obtained by increasing particle radius and reducing
surface separation show that the failure is sensitive to configuration scale:
particle radius controls allowed penetration depth, while piston speed and
\(\Delta t\) control the per-frame closing distance.

The generator reports explicit overcompression ratios:
\[
    \frac{v_c\Delta t}{\delta_{\mathrm{target}}}
    \qquad\mathrm{and}\qquad
    \frac{v_c\Delta t}{\delta_{\mathrm{hard}}}.
\]
For simple particle cases \(v_c\) is the relative normal particle speed.  For
reservoir and compressor cases it is the piston normal speed.  These ratios
make it visible when a configuration is asking one frame to consume too much
of the available compression depth.

The simple GLSL exporter now labels the two max-depth projection phases:
\begin{itemize}
    \item \texttt{Sub:9401}: first projection after force velocity;
    \item \texttt{Sub:9402}: second projection after post-resolution
          remaining-depth constraints.
\end{itemize}
Existing detail codes remain for hard-depth violation, invalid normals,
constraint overflow, and unstable projection.

\subsubsection{difference outcome}

particle solution

\subsubsection{Analysis}

The model is behaving differently in sparse gas-like collision and dense
liquid-like compression.  Sparse contacts can be judged by pair timing and
stiffness response.  Dense compressor loading must also be judged by the
ratio between per-frame displacement and available compression depth.  A run
that fails after the whole chamber is collision-saturated is not necessarily
showing a broken piston or a bad wall; it may be showing that the selected
radius, spacing, piston speed, and timestep have formed an overcompressed
packed column.

The next implementation work is not to hide this condition.  The correct
next step is to use the new ratios and GLSL subcodes to separate three cases:
configuration overcompression, timestep-resolution failure, and true
projection bugs in the source-owned multi-contact solver.

\subsection{Vulkan Frame Pacing and Contact Workload}

\subsubsection{previous form}

The CD-nozzle and reservoir Vulkan runs were interpreted mainly through
dynamics errors such as \texttt{ERROR\_MAX\_DEPTH\_CONSTRAINT},
\texttt{ERROR\_PENETRATION\_STEP\_TOO\_LARGE}, and
\texttt{ERROR\_WALL\_TUNNELING}.  When the visual frame rate hesitated or
jerked, the first suspicion was the Python viewer, the HSV display, or the
particle renderer.  That explanation did not apply to the Vulkan run.

The Vulkan particle-cell occupancy path uses a graphics pass to rebuild the
cell lists.  \texttt{CommandParticleGraphics.cpp} clears both the collision
index buffer and the lock/count buffer with \texttt{vkCmdFillBuffer}, then
the vertex shader repopulates them:
\begin{verbatim}
slot = atomicAdd(L[sltidx], 1);
clink[sltidx].idx[slot] = index;
\end{verbatim}
The compute shader then scans \texttt{clink[loc].idx[slot]} until it reaches
particle ID zero.  Particle ID zero is the null particle, so zero is the
correct empty sentinel for occupancy slots.

\subsubsection{current form}

The zero convention has two separate domains:
\begin{itemize}
    \item particle ID zero means an empty occupancy slot;
    \item cell location zero is a valid cell unless the indexer is changed to
          reserve it.
\end{itemize}
Therefore \texttt{CornerList.ploc} must continue to use \texttt{npos} as its
empty-cell sentinel.  Clearing corner locations to zero would make unused
corners look like valid cell zero and would cause compute to scan cell zero
for non-existent corners.

The major Vulkan frame-pacing contributors are now understood as workload
and synchronization issues:
\begin{itemize}
    \item the main loop calls \texttt{vkDeviceWaitIdle()} every frame, so
          variable GPU work becomes visible as CPU-side frame pacing jitter;
    \item capture/readback paths call \texttt{vkQueueWaitIdle()}, map a
          staging buffer, and write a capture file, so capture frames are
          expected to stall;
    \item dense compression increases the number of occupied cell slots
          visited by each source;
    \item the generated compute wrapper currently performs two occupancy
          scans per source, one for force accumulation and one for
          post-resolution remaining-depth constraints;
    \item the duplicate-target array is sized from
          \texttt{MAX\_CELL\_OCCUPANY * 8} and initialized twice per source,
          so oversized cell occupancy increases fixed per-source work even
          when zero termination is correct.
\end{itemize}

Lowering \texttt{MAX\_CELL\_OCCUPANY} improved frame pacing because it reduced
both the maximum scan depth and the duplicate-list initialization size.  This
does not change the physical model, but it makes clear that the generator
should choose the smallest safe occupancy capacity and report why that value
is required.

\subsubsection{difference outcome}

particle solution

\subsubsection{Analysis}

The current Vulkan hitches are not evidence by themselves that wall geometry
or collision response is wrong.  They are partly a scheduling and capacity
problem.  The first cleanup is to preserve the existing sentinel convention:
zero for empty particle slots, \texttt{npos} for empty cell references.  The
next Vulkan cleanup is to remove unnecessary whole-device waits from ordinary
frame pacing and to replace the second contact scan with source-local contact
records captured during the first scan.

The intended compute schedule is:
\begin{verbatim}
scan local occupancy once
build compact source-owned contact records
accumulate force from those records
calculate candidate velocity
apply maximum-depth projection from those records
check post-resolution remaining depth from those records
move the source
\end{verbatim}
If the source-owned contact record list overflows, that is a real dynamics
error, not a reason to rediscover the same contacts through a second broad
scan.

\subsection{Depth-Dependent Contact Stiffness}

\subsubsection{previous form}

The active contact law used a constant \texttt{collision\_stiffness\_q}.
Contact force in depth mode was proportional to current penetration depth:
\[
    F = q\,\delta .
\]
This works for sparse two-particle collisions because there is usually one
dominant contact and enough open space for ordinary force integration to slow,
stop, and reverse the particles.

In dense compressor and CD-nozzle runs, the same constant stiffness must
resist a collective pressure field.  Many particles can be in simultaneous
contact from the piston to the pipe exit.  The force law remains modest even
as individual contacts approach the configured hard penetration depth.
Maximum-depth containment then becomes the first strong resistance instead of
the last safety limit.

\subsubsection{current form}

The approved direction is to keep the same momentum-governed dynamics and
replace constant stiffness with a reversible depth-dependent stiffness.  No
energy variable is added to the dynamic state.  Energy may remain a diagnostic
or a design guide, but the runtime state remains position, velocity, force,
and existing contact/momentum diagnostics.

The proposed form is:
\[
    x = \mathrm{clamp}
        \left(\frac{\delta}{\delta_{\mathrm{hard}}},0,1\right),
\]
\[
    q_{\mathrm{eff}} =
        q_{\mathrm{base}}
        \left(1 + g x^p\right),
\]
\[
    F = q_{\mathrm{eff}}\delta .
\]
Here \(q_{\mathrm{base}}\) is the existing contact stiffness, \(g\) is the
compression gain, and \(p\) controls how sharply resistance rises near hard
compression.

The same force curve is used while compressing and while releasing.  The
force depends on current physical penetration, not on whether the contact is
moving inward or outward.  The impulse generated by this force over time
naturally slows, stops, and reverses the particles.  There is no separate
stored energy or manual force payback.

\subsubsection{difference outcome}

particle solution

\subsubsection{Analysis}

Depth-dependent stiffness fits the intended particle interpretation: contact
is soft at first, but resistance rises as the remaining compressible depth is
consumed.  This should reduce dependence on the hard maximum-depth projection
as the first meaningful pressure response in dense flow.

Momentum consistency still comes from symmetric force evaluation.  For
particle-particle contacts, both source invocations must evaluate the same
frame-start geometry and the same scalar force law with opposite normals.  For
stationary walls and prescribed pistons, the equal-and-opposite momentum
belongs to the external boundary system.  Piston and wall impulse reporting
remains the correct diagnostic for those external reactions.

The important cleanup is to distinguish ordinary reversible contact force
from blocked maximum-depth motion.  Ordinary contact force changes velocity
through \(F\Delta t\).  Maximum-depth containment should only remove illegal
inward motion that would exceed the configured hard limit; it should not
become the main material stiffness.

\subsubsection{implementation result}

The depth-dependent stiffness law was added first to the Python dynamics and
then exported into the shared simple GLSL family.  Testing
\texttt{TestReservoir} showed a large stability improvement, which supports
the diagnosis that constant \texttt{collision\_stiffness\_q} was too weak for
dense reservoir compression.

This is an important result because the improvement was made in the common
contact-force path rather than in a reservoir-specific special case.  The same
law applies to particle-particle contacts, function-wall contacts, and piston
contacts when they pass through the simple dynamics contact-force path.

\subsubsection{remaining synchronization}

The GLSL implementation currently uses fallback preprocessor defaults for the
new material shape:
\begin{verbatim}
FORCE_DYNAMICS_SIMPLE_COMPRESSION_STIFFNESS_GAIN  = 8.0
FORCE_DYNAMICS_SIMPLE_COMPRESSION_STIFFNESS_POWER = 2.0
\end{verbatim}
These values match the successful test direction, but they should not remain
hardcoded.  The next cleanup is to export
\texttt{compression\_stiffness\_gain} and
\texttt{compression\_stiffness\_power} from the configuration into generated
local GLSL so Python and Vulkan cannot drift.

\subsection{FreeStream Timed Release And Vulkan Debug Print Bring-Up}

\subsubsection{previous form}

\texttt{FreeStream} was introduced as a streaming inlet experiment using the
new simple function-wall GLSL family.  Its generated mobile particles carried
future release frames in \texttt{Data.w}, and the Vulkan wrapper was copied
from nearby simple or reservoir wrapper structure during bring-up.  That made
two separate problems easy to confuse: particles that were not visibly moving,
and shader debug prints that appeared to change how long the application ran.

The copied compute wrapper also enabled the optional piston include even
though freestream flow is not a piston-driven reservoir.  That added an
unwanted moving-wall path before the ordinary source-owned contact scan.

\subsubsection{current form}

\texttt{FreeStream} uses \texttt{GenStreaming.py} to generate timed inlet
release records.  Mobile particles are written at the inlet with initial
velocity from \texttt{initial\_particle\_velocity}.  Their
\texttt{Data.w} value stores the frame at which they become active; negative
values still mean dead particles.  Boundary markers remain ordinary
\texttt{ptype=1} function-wall locality records.

The freestream compute wrapper must include the base simple dynamics,
boundary-particle locality helpers, and function-wall helpers only.  It must
not include \texttt{ForceDynamicsSimplePiston.glsl} or define
\texttt{FORCE\_DYNAMICS\_SIMPLE\_PISTON\_AVAILABLE}.

During debugging, \texttt{debugPrintfEXT} messages were found to be active
runtime events rather than passive log lines.  They arrive through the Vulkan
debug-utils callback as \texttt{VVL-DEBUG-PRINTF} messages.  The callback was
setting \texttt{Extflg} for informational messages, so a debug print at frame
100 made the application stop or short-circuit at frame 100, while moving the
same print to frame 200 let it run to frame 200.

\subsubsection{difference outcome}

particle solution

\subsubsection{Analysis}

The FreeStream generated binary was not the immediate source of the missing
motion: particle records had the expected mobile type, release frame, radius,
mass, stiffness, and inlet velocity.  The first cleanup is therefore wrapper
and diagnostic hygiene.  Freestream must not enable piston behavior, and
shader debug-print messages should be intercepted separately from validation
errors.

The debug-utils callback should recognize \texttt{VVL-DEBUG-PRINTF}, strip
the Vulkan wrapper text down to the shader message after
\texttt{DebugPrintf:}, print that payload, and return without setting
\texttt{Extflg}.  Validation errors and warnings may still set
\texttt{Extflg}; shader printf output should not behave like a stop
condition unless a separate debug-break option is explicitly added.

This keeps shader printf usable as a frame-local diagnostic while preserving
the distinction between a real Vulkan validation problem and an intentional
particle-dynamics trace.

\subsubsection{configuration synchronization}

The CD-nozzle Python and Vulkan include configurations now carry the same
depth-dependent contact controls:
\begin{verbatim}
compression_stiffness_gain
compression_stiffness_power
target_penetration_fraction
hard_penetration_fraction
min_compression_frames
\end{verbatim}
This keeps data generation, Python dynamics, and Vulkan test-file metadata on
the same material convention.

A Vulkan compile failure exposed one important porting rule.  When a
\texttt{.tst} value is promoted into local GLSL as a preprocessor macro, the
macro body must be the raw numeric token and must not include a trailing
semicolon.  Otherwise an expression such as
\begin{verbatim}
max(0.0, FORCE_DYNAMICS_SIMPLE_COMPRESSION_STIFFNESS_GAIN)
\end{verbatim}
expands into invalid GLSL.  The generated local GLSL should therefore emit:
\begin{verbatim}
#define FORCE_DYNAMICS_SIMPLE_COMPRESSION_STIFFNESS_GAIN 8.000000000
#define FORCE_DYNAMICS_SIMPLE_COMPRESSION_STIFFNESS_POWER 2.000000000
\end{verbatim}
rather than semicolon-terminated macro values.

\subsection{Lighting Generator Uses Release Frames}

\subsubsection{previous form}

The project had two competing lifecycle descriptions.  The current simple and
reservoir theory correctly retired runtime scheduled birth and deferred
admission, while FreeStream still used positive \texttt{Data.w} values for
timed inlet release records.  That made a new lighting/visualization
generator ambiguous: it needed scheduled appearance without reviving the old
reservoir birth model.

\subsubsection{current form}

\texttt{GenLightingData.py} inherits \texttt{GenStreaming.py} and uses
explicit \texttt{PARTICLE\_DATA} records.  Each particle may provide
\texttt{releaseFrame}; the generator validates that it is finite and
nonnegative, then writes it to \texttt{pdata.state\_flg}.  In the generated
runtime record this becomes \texttt{Data.w}.  Zero means active from frame
zero, and a positive value means pending until the current frame reaches that
release frame.

\subsubsection{difference outcome}

particle solution

\subsubsection{Analysis}

Positive \texttt{Data.w} is now a generator-scoped release-frame convention
for \texttt{GenStreaming.py} and \texttt{GenLightingData.py}.  It is not a
general reservoir birth/admission rule.  The generator never relocates the
particle at runtime and never writes target particles; it only emits an
initial record whose lifecycle gate delays participation until the configured
frame.

The ordinary simple and reservoir models still generate particles active from
frame zero and use negative \texttt{Data.w} as permanent dead state.  This
keeps the retired deferred-birth investigation separate from controlled
streaming and lighting release schedules.

\subsection{Python Lumens Eye-Aperture Visualization}

\subsubsection{previous form}

Python particle coloring was still partly described by the retired flat
\texttt{hsv\_color} switch, and lighting particles had no material-owned
lumens color scheme.  The first lumens visibility pass also made it easy to
color a photon at wall rebound if its future ray would pass near the eye,
even though the particle had not yet reached the eye aperture in the current
simulation frame.

\subsubsection{current form}

Color selection belongs to \texttt{material\_properties.color\_scheme}.  The
named material schemes include \texttt{COLLISION}, \texttt{HSV},
\texttt{WHITE}, \texttt{RED}, \texttt{GREEN}, \texttt{BLUE}, and
\texttt{LUMENS}.  The \texttt{LUMENS} scheme is Python-only in this pass.
It uses:
\begin{verbatim}
lumens_eye_position
lumens_eye_angle_degrees
lumens_eye_fov_degrees
lumens_eye_radius
lumens_fade_in
lumens_surface_color
lumens_debug_dim
lumens_draw_eye_radius
\end{verbatim}

The lumens renderer evaluates the current particle position, the current
velocity angle stored in \texttt{VelRad.w}, the eye position, the eye look
angle, and the eye field of view.  A photon can become visibly colored only
after its current position crosses the configured eye radius.  If
\texttt{lumens\_fade\_in} is true, brightness ramps from the aperture boundary
toward the eye center; otherwise visibility inside the aperture uses the
alignment/FOV brightness without distance fade.

\subsubsection{difference outcome}

rendering and diagnostics

\subsubsection{Analysis}

The lumens model is now a controlled Python visualization prototype rather
than a dynamics rule.  It does not change particle forces, positions,
velocities, or release-frame lifecycle state.  Non-visible photons may be
hidden or drawn dimly for debugging, but that decision remains a viewer
choice.

\texttt{GenLightingData.py} provides a scenario-specific
\texttt{SpecificDraw} hook used by the Python pygame runner to draw the eye
point, look ray, field-of-view rays, and optional aperture radius.  This keeps
lighting diagnostics beside the lighting generator while leaving the generic
dynamics model independent of renderer-only tools.

The Vulkan side is intentionally unchanged for this check-in.  The later 3D
work should replace static Python eye cfg values with camera-derived eye
position and view direction, then port the material-owned lumens display rule
to the Vulkan rendering path.

\subsection{Three-Dimensional Rectangle Walls Enter The Simple Path}

\subsubsection{previous form}

The active simple wall path was documented and generated around
\texttt{curve\_wall\_segments}.  Those segments represented finite
two-dimensional function walls.  Boundary markers provided locality, but the
wall evaluator was still the function-wall helper.  The Vulkan
\texttt{TestPythonBoundarySpheres} wrapper therefore had no active finite
rectangle-wall model for three-dimensional box walls.

\subsubsection{current form}

\texttt{rectangle\_wall\_segments} is now the active three-dimensional simple
wall model when present.  Each segment carries:
\begin{verbatim}
origin, uAxis, vAxis, uLength, vLength, inwardNormal, wallFlag
\end{verbatim}
and the generated test-file row uses fifteen numeric values.  Boundary marker
particles remain locality sentinels with \texttt{ptype=1}; they are not
physical wall geometry and do not store evaluator IDs.

The Python runtime, generic data generator, simple GLSL exporter, and
\texttt{TestPythonBoundarySpheres} compute wrapper now share the configured
wall precedence rule: rectangle walls take precedence when present; otherwise
the simple path falls back to function-wall segments.  The active generated
GLSL family includes \texttt{ForceDynamicsSimpleRectangleWall.glsl}, and the
scenario wrapper calls \texttt{EvaluateConfiguredWallSegment(...)} for
boundary-marker targets.

\subsubsection{difference outcome}

particle solution

\subsubsection{Analysis}

This milestone moves the simple path from a two-dimensional wall-only Vulkan
port toward the intended three-dimensional boundary model.  The rectangle
wall evaluator performs finite patch checks in the configured \(u\) and \(v\)
directions, tests inward-normal penetration, and returns the inverted normal
needed by the existing source-owned force path.  The generated
\texttt{boundary.glsl} contract now requires both the rectangle segment array
and a zero-count fallback so wrappers can compile in either wall mode.

The data generator must size cell occupancy for boundary-marker locality, not
for physical wall volume.  In \texttt{TwoParticleHorizontal}, the six
rectangle patches require \texttt{cell\_occupancy\_list\_size=16}.  Smaller
values fail the generator's corner-occupancy capacity check before runtime.
The remaining work is broad Vulkan validation across every active scenario
wrapper and cleanup of any generated or hand-maintained shader drift found by
that validation.

\subsection{Runner Selection For 3D Lumens Bring-Up}

\subsubsection{previous form}

The data-generation UI's \texttt{Simulate} action imported only
\texttt{SimulationRunner3D\_HSVPanel.py}.  As a result, every simulation run
entered the Open3D runner even when the cfg declared:
\begin{verbatim}
dimensions = 2;
\end{verbatim}
This made the lumens configuration difficult to reason about because
\texttt{lumens.cfg} is currently a two-dimensional pygame visualization
prototype, while the next planned work is three-dimensional lumens behavior.

\subsubsection{current form}

The UI now treats \texttt{dimensions} as the runner-selection contract.
\texttt{dimensions=2} selects \texttt{SimulationRunner\_HSVPanel.py}, and
\texttt{dimensions=3} selects \texttt{SimulationRunner3D\_HSVPanel.py}.
Missing \texttt{dimensions} defaults to the two-dimensional runner for
compatibility with older cfg files.  Included cfg values are folded into the
selection path before the runner is chosen.

\subsubsection{difference outcome}

rendering and diagnostics

\subsubsection{Analysis}

This change separates two decisions that had been accidentally coupled:
whether a scenario is simulated in the two-dimensional pygame viewer or the
three-dimensional Open3D viewer, and how lumens visibility should eventually
derive its eye state.  The current \texttt{lumens.cfg} remains a
two-dimensional diagnostic that uses static eye position, look angle, field
of view, and aperture radius.  The planned 3D lumens work belongs in the
three-dimensional runner, where the eye can be derived from camera/view state
instead of static two-dimensional cfg fields.

This is a viewer-routing change only.  It does not change particle dynamics,
release-frame lifecycle, material selection, or the lumens color rule.

\subsection{Lighting In 3D Vulkan Boundary Rendering}

\subsubsection{previous form}

The Vulkan lighting bring-up still inherited boundary rendering assumptions
from the older particle-boundary application split.  Scenario shaders carried
large duplicated boundary, sphere, particle, fragment, and compute bodies, so
common changes could drift between \texttt{CDNozzle}, \texttt{FreeStream},
\texttt{FreeStreamHetero}, \texttt{TestPythonBoundarySpheres}, and
\texttt{TestReservoir}.  Rendering switches such as
\texttt{boundary\_as\_obj}, \texttt{particle\_as\_spheres},
\texttt{show\_cell\_boundary\_cube}, and
\texttt{show\_wall\_as\_boundary\_cube} were not consistently represented in
both the generated shader metadata and the Vulkan draw command path.

This created two visible failure modes during 3D lighting work.  First, a
run with boundary drawing disabled could still record an indexed boundary
draw and bind a null cube index buffer.  Second, a run with
\texttt{boundary\_as\_obj=true} could generate \texttt{HAS\_BOUNDARY} in
\texttt{boundary.glsl} but fail to expose that define to the common boundary
vertex shader, collapsing the boundary vertices instead of drawing the OBJ.

\subsubsection{current form}

Boundary and sphere rendering are now controlled by generated
\texttt{boundary.glsl} feature defines:
\begin{verbatim}
HAS_BOUNDARY
HAS_SPHERE
\end{verbatim}
\texttt{HAS\_BOUNDARY} is emitted when a cell cube, wall cube, or boundary
OBJ is configured for drawing.  \texttt{HAS\_SPHERE} is emitted when
particle spheres are configured for drawing.  The common boundary vertex
shader consumes those defines, while each scenario wrapper includes its local
\texttt{boundary.glsl} before including \texttt{../common/Boundary.vert}.

The duplicated scenario shader bodies have been reduced to wrappers around
the common shader bodies:
\begin{verbatim}
../common/Boundary.vert
../common/FPM.vert
../common/FPM.frag
../common/FPM.comp
\end{verbatim}
Scenario wrappers remain responsible for local generated metadata, uniforms,
materials, and optional helper includes.  The shared compute body calls
\texttt{EvaluateConfiguredWallSegment(...)} so rectangle-wall scenarios can
override the wall evaluator while function-wall scenarios use the default
function-wall path.

The Vulkan boundary-and-sphere command path now mirrors the same rendering
contract.  It records boundary cube or OBJ draws only when a boundary is
configured, records sphere draws only when spheres are configured, and no
longer binds the cube index buffer for sphere-only or no-boundary cases.

\subsubsection{difference outcome}

rendering and diagnostics

\subsubsection{Analysis}

This work is part of the 3D lighting visualization path, not a change to the
force law.  It does not change contact detection, contact force, release-frame
lifecycle, wall geometry, or particle integration.  It makes the Vulkan
renderer obey the same feature flags that the generated shader metadata
advertises, which is required before lighting particles can be inspected
reliably in three-dimensional scenes.

The important separation is that shader feature defines hide or show rendered
geometry, while Vulkan command recording controls which buffers may legally be
bound and drawn.  A disabled boundary must be skipped before command
recording reaches the cube vertex or index buffers; a shader cannot repair a
null-buffer validation error after the command has already been recorded.

\subsection{Photon Material Behavior And Three-Dimensional Streams}

\subsubsection{previous form}

The cathedral lighting scene began as a regular streaming setup with one light
stream and one dust stream.  The streams could release particles over time, but
the stream shape was still effectively a line on one depth plane.  Photon and
dust particles were distinguished mainly by material naming and color intent;
the force path had no material-owned rule that prevented photons from
colliding with other photons.

This made the intended lighting behavior ambiguous.  A photon beam could create
ordinary particle-particle contacts inside itself, and a three-dimensional
cathedral beam could still look like a two-dimensional sheet if the generator
only varied one span axis.

\subsubsection{current form}

Particle type is now a material property.  A material may declare:
\begin{verbatim}
particle_type = "regular";
particle_type = "photon";
\end{verbatim}
The generated material contract preserves this value in the \texttt{.tst}
material entries and in \texttt{material.glsl}.  The shader-side material
record carries \texttt{particleType}, and the photon helper identifies photon
particles by looking up the particle's material id.

The common particle-particle force loop skips photon targets.  Photon-photon
pairs are skipped, and regular-source/photon-target pairs are skipped so dust
cannot receive force from photons.  Photon-source/dust-target pairs remain
active so the photon can redirect at fixed speed.  Dust-dust and photon-wall
interactions remain available to their ordinary paths.  The boundary and wall
paths are not disabled by the particle-pair skip.

Photon speed is invariant.  The configured stream
\texttt{initial\_particle\_velocity} sets the speed by vector magnitude, so the
material contract does not add a separate \texttt{particle\_speed} or
\texttt{photon\_speed} field.  Photon collisions with dust or walls may change
direction, but the outgoing photon velocity must retain the incoming speed.

The streaming generator also supports emitter-plane patches.  A stream fixes
one inlet coordinate and spans finite \(u\) and \(v\) axes with independent
counts.  \texttt{FreeStream} uses the same structure with a degenerate
one-cell \(v\) span so its historical line stream remains unchanged.
\texttt{Cathedral} uses non-degenerate rectangular patches for both the photon
beam and the dust source, producing a true three-dimensional release pattern.

\subsubsection{difference outcome}

particle solution

\subsubsection{Analysis}

The photon rule is intentionally material-owned rather than scenario-name
owned.  This keeps the force decision local to the particle pair and lets later
lighting scenarios reuse the same shader helper without hard-coding Cathedral
particle ids.

The photon-dust rule is intentionally non-Newtonian light behavior, not
reciprocal particle momentum exchange.  Photon and dust particles are treated
as separate momentum systems: photon contacts may redirect photons at fixed
speed, while dust momentum remains controlled by dust-dust and dust-wall
interactions.  A dust source therefore skips photon targets instead of
accumulating photon contact force.

Keeping photon speed on the stream velocity avoids duplicating speed in both
material and stream configuration.  The material identifies photon behavior;
the stream supplies the actual velocity vector.  A fixed-speed photon response
therefore reflects direction from the contact normal and then normalizes back
to the incoming speed magnitude.

The emitter-plane fields are generator input, not runtime shader metadata.
Python expands each configured stream patch into concrete particle positions,
velocities, materials, and release frames.  The Vulkan path sees the resulting
particle data and material contract; it does not need to know which stream
span axes created the particles.

Cell-occupancy sizing remains a data-capacity question rather than a
dimension-flag question.  Moving from a line stream to a rectangular patch can
change which particles share a cell or cell corner, so the generator report is
still the authority for \texttt{cell\_occupancy\_list\_size}.

\subsection{Stopped Vulkan Frames And Simulation Time}

\subsubsection{previous form}

The Vulkan stop flag was added as a compute-side pause control, but the frame
clock semantics were still mixed.  \texttt{StopFlg} could make
\texttt{FPM.comp} return before particle dynamics, while the frame value used
by shader lifecycle checks still advanced with the render/application frame.
The selected particle position buffer could also be treated as a render-frame
toggle instead of a simulation-frame toggle.

This produced a confusing partial-pause state.  The scene continued to draw,
but timed-release particles could become active because
\texttt{frameNum} advanced while dynamics were stopped.  In a lighting stream
scenario, that made one stream appear to continue running even though the
compute shader had been told not to process particles.

\subsubsection{current form}

The push-constant contract now separates render time from simulation time:
\begin{itemize}
    \item \texttt{actualFrame} is render/application time and may always
          advance;
    \item \texttt{frameNum} is simulation time and advances only when dynamics
          are allowed to run;
    \item \texttt{positionBuffer} advances only with \texttt{frameNum};
    \item \texttt{StopFlg} pauses compute particle processing while graphics
          continues to draw the current simulation state.
\end{itemize}

Shader code that controls lifecycle, timed release, piston motion, or
dynamics diagnostics uses \texttt{frameNum}.  Render-only reporting or capture
logic may use \texttt{actualFrame} when it needs wall-clock frame progression.

\subsubsection{difference outcome}

rendering and diagnostics

\subsubsection{Analysis}

The stop flag is not a complete pause by itself.  It only prevents the compute
shader from updating particle dynamics.  A correct stopped frame also requires
the simulation clock and simulation buffer selector to remain frozen, because
graphics shaders still evaluate lifecycle state and draw from the selected
particle buffer.

This distinction is especially important for lighting streams, where pending
particles use \texttt{Data.w} release frames.  If render time and simulation
time are conflated, a stopped scene can still release new particles.  Keeping
\texttt{frameNum} in the simulation-time domain preserves the visible paused
state while allowing the application to continue drawing and responding to
input.

\subsection{Vulkan Shader Export Recovery}

\subsubsection{previous form}

During the three-dimensional lighting work, Vulkan dynamics behavior moved
ahead through direct edits to shared shader files.  The common compute path
gained rectangle-wall dispatch, photon pair filtering, fixed-speed photon
reflection, stop-flag handling, and a configured duplicate-list size.  Those
changes proved the intended runtime behavior, but they also allowed the Python
exporter to fall behind.

The immediate risk was that regenerating simple GLSL wrappers could reintroduce
the old inline compute loop, including stale duplicate-list sizing and missing
photon helper usage.  That made the export path unsafe even though the current
Vulkan shader files were still the working reference.

\subsubsection{current form}

The active Vulkan compute schedule is centralized in
\texttt{vulkan/sim/common/FPM.comp}.  Scenario compute wrappers include that
common file and provide only scenario-owned setup: generated params, material
metadata, boundary metadata, workgroup layout, simple dynamics helpers, optional
piston support, and the configured-wall dispatcher that chooses rectangle walls
when present.

\texttt{tools/ExportForceDynamicsSimpleGLSL.py} was adjusted so it no longer
emits a second stale copy of the FPM loop into scenario wrappers.  The exporter
now emits wrappers that include \texttt{material.glsl},
\texttt{photons.glsl}, \texttt{ForceDynamicsSimpleRectangleWall.glsl}, the
configured-wall dispatcher, and \texttt{../common/FPM.comp}.  This prevents a
wrapper regeneration from overwriting the manually proven common compute
schedule with the old inline loop.

\subsubsection{difference outcome}

particle solution

\subsubsection{Analysis}

This is a recovery step, not the final export architecture.  Vulkan is
currently working from the shared common shader path, but the source-of-truth
boundary is still imperfect: \texttt{FPM.comp} is hand-maintained common GLSL
while the helper functions and scenario wrappers are generated from Python.

The next correction is to repair Python export parity.  The Python-generated
data and GLSL metadata must always emit the photon-aware material contract
before wrappers depend on \texttt{photons.glsl}; otherwise a scenario can
include photon helpers while still carrying an older \texttt{MaterialProperty}
layout without \texttt{particleType}.  The exporter should also document and
own the current wrapper contract explicitly so regeneration is a validation
step rather than a chance to regress a working Vulkan shader.

\subsection{Cathedral Optical Photon Export}

\subsubsection{previous form}

The Python Cathedral prototype proved that photon-dust interaction should
redirect photons without pushing dust, but the first Vulkan photon path still
used the ordinary particle collision processor after detecting a photon-source
and dust-target contact.  That preserved fixed photon speed, but it also
allowed the source to allocate contact state, accumulate force, and enter the
ordinary maximum-depth/contact-resolution machinery.

The result was a sticky optical contact.  A photon could appear to lose useful
motion while it remained inside the dust contact region, and diagnostic
internal momentum could grow even though photons are not supposed to own
mechanical contact storage.

\subsubsection{current form}

Photon-dust contact is now an optical event in both Python and the generated
simple Vulkan shader path.  For a photon source targeting a dust particle, the
runtime performs swept source motion against the target's physical radius,
reflects the photon direction at its incoming fixed speed, computes the
reflected end-of-frame position from the impact point and remaining timestep,
and then skips the ordinary particle collision processor.

The active simple exporter now owns this common shader behavior.  It writes
\texttt{vulkan/sim/common/photons.glsl} with material-owned photon lookup,
pair filtering, fixed-speed reflection, and swept optical contact helpers.  It
also writes \texttt{vulkan/sim/common/FPM.comp}, so the shared compute loop
consumes optical photon-dust events before mechanical collision processing and
keeps scenario wrappers as thin includes around the common schedule.

\subsubsection{difference outcome}

particle solution

\subsubsection{Analysis}

The important separation is between optical redirection and mechanical
contact.  Dust-dust and wall interactions remain ordinary force contacts.
Photon-photon pairs are skipped.  Regular-source/photon-target pairs are
skipped so dust is not moved by photons.  Photon-source/dust-target pairs
remain detectable, but successful detection does not allocate a contact slot,
does not add source-local force, and does not accumulate internal contact
momentum.

Export ownership matters as much as the physics rule.  The working Vulkan
behavior is no longer a hand-edited exception: regeneration through
\texttt{tools/ExportForceDynamicsSimpleGLSL.py --write} now rewrites the
photon helper and the shared FPM schedule with the optical rule intact.  The
next Cathedral customization can therefore adjust stream placement, material
display, and scene geometry without re-opening the Python/Vulkan photon drift
that caused the sticky-contact behavior.
