
% !TeX root = main.tex
\newpage
\section{Programming Conventions}

\subsection{Current .venv environment}

Use this .venv to coordinate programming tasks:
\begin{verbatim}
c:\_DJ\gPCD\python\.venv\Scripts\python.exe
\end{verbatim}

\subsection{Python Viewer Color Utilities}

Python viewer colorization belongs in viewer or utility code, not in
\texttt{<model>Dynamics.py}.  The dynamics model may calculate and store the
velocity angle in \texttt{VelRad.w}, but conversion from angle to RGB is a
display operation.

The Python HSV velocity-angle helper is:
\begin{verbatim}
gbase/utilities.py:hsv_angle(angle, value, sat)
\end{verbatim}

\texttt{SimulationRunner\_HSVPanel.py} applies this helper when a particle's
material \texttt{color\_mode} is \texttt{VELOCITY}.  The helper reads the
angle in radians, maps it to hue with \(1/(2\pi)\), clamps saturation and
value to \([0,1]\), and returns an RGB tuple for the Python renderer.  The
former global \texttt{hsv\_color} switch and the old named solid color schemes
are retired; display selection belongs to \texttt{material\_properties} as
\texttt{color\_mode} and \texttt{color}.

\subsection{Flat Runtime Configuration}

Runtime and generator configuration items are stored directly in the cfg root.
The retired \texttt{RUN\_CONFIGURATION} subgroup must not be added to new cfg
files or required by Python code.  Python consumers use
\texttt{get\_run\_configuration(config)} during the transition so old external
cfg files with the subgroup remain readable, but active project cfg files use
the flat form.

Future GLSL color utilities should mirror this interface but remain separate
from generated ForceDynamics collision code.  Color is diagnostic display
state and must not become a dynamics input.

\subsection{Scenario-Specific Python Drawing}

Python simulation-specific diagnostic drawing belongs in the generator module,
not in the dynamics model.  A generator may expose:
\begin{verbatim}
SpecificDraw(screen, run_configuration, dynamics, view_box, draw_helpers)
\end{verbatim}

\texttt{SimulationRunner\_HSVPanel.py} calls this optional hook after drawing
configuration boundaries and before drawing particles.  The hook may draw
scenario aids such as a lumens eye point, look ray, field-of-view rays, or eye
aperture.  It must not mutate particle state, contact state, or dynamics
configuration.

The \texttt{draw\_helpers} object owns viewer conversion helpers such as
world-to-screen mapping and world-radius-to-pixel conversion.  Custom draw
code should use those helpers instead of duplicating screen-coordinate math in
each generator.

\subsection{GLSL Viewer Color Utilities}

GLSL velocity-angle colorization belongs in rendering utilities and scenario
vertex shaders, not in the ForceDynamics force law.  The current shared GLSL
helper is:
\begin{verbatim}
vulkan/sim/common/util.glsl:colorizeVelocity(angle, sat, val)
\end{verbatim}

The helper reads \texttt{angle} in radians, maps it to hue with
\(1/(2\pi)\), clamps saturation and value to \([0,1]\), and returns a
\texttt{vec3} RGB color.  Runtime color selection is driven by material
\texttt{color\_mode}; the velocity path uses the mode constant:
\begin{verbatim}
const uint COLOR_MODE_VELOCITY;
\end{verbatim}

Vertex shaders that output \texttt{vec4} colors must wrap the returned
\texttt{vec3} with an alpha channel.  Vertex shaders that output
\texttt{vec3} colors may assign the return value directly.
