% !TeX root = main.tex
\section{Programming Conventions}

\subsection{File Boundaries}

\texttt{<model>Dynamics.py} is the GLSL-portable dynamics layer.  It should contain
only functions that correspond directly to shader-side dynamics functions.

\texttt{<model>DynamicsBase.py} is the Python harness.  It may load configuration files,
create Python data structures, build particle arrays, set push constants,
iterate particles, and call the model dynamics functions.

\texttt{Reporting.py} is reserved for reporting and diagnostic output.  It
should not own dynamics decisions.


\subsection{Creating New Dynamics}

The command for creating a new model is:
Create <model>

Creating new dynamics classes requires the following definitions.
\begin{enumerate}

\item Create <new model>DynamicsFlowChart.py, <new model>Dynamics.py, and
<new model>DynamicsBase.py by copying the corresponding files for the model
selected by the \texttt{live\_base = <model>} configuration item in
\texttt{ParticleUtil.cfg}.

\item Copy the live model's <live model>DynamicsSource.tex and
<live model>DynamicsFlowTikz.tex files and rename them
<new model>DynamicsSource.tex and <new model>DynamicsFlowTikz.tex.  The initial
copies document the unchanged starting model before development begins.

\item Add <new model>DynamicsSource.tex to \texttt{main.tex}.

\item Register <new model>Dynamics in \texttt{SimulationRunner.py} so that
\texttt{live\_base} and \texttt{shadow\_base} can select the new model.

\item Now \texttt{live\_base = <old model>}. Change \texttt{ParticleUtil.cfg} to make \texttt{live\_base = "<new model>Dynamics";} and 
\texttt{shadow\_base = "<old model>Dynamics";}.

\item Determine the data required for the diagnostics plots.
\item Create a new diagnostics plotting script named <new model>Plots.py
\end{enumerate}
\subsection{Programming Styles}

When discussing software architecture, two common programming styles frequently emerge. Although there is not a universally accepted pair of names for these approaches, they are often described as \emph{control-oriented} and \emph{data-oriented} designs.

\subsection{Control-Oriented (Hierarchical) Design}

In this style, functions determine what happens next by directly invoking other functions within their bodies.

\begin{lstlisting}[language=C++]
void Process()
{
LoadData();

```
if(condition)
    MethodA();
else
    MethodB();
```

}

void MethodA()
{
...
MethodC();
}
\end{lstlisting}

Characteristics include:

\begin{itemize}
\item Deep call trees.
\item Control flow is distributed among functions.
\item Functions decide what operation occurs next.
\item Execution resembles a hierarchy or tree structure.
\end{itemize}

Common names for this style include:

\begin{itemize}
\item Hierarchical Design
\item Top-Down Design
\item Control-Oriented Programming
\item Call-Tree Architecture
\item Procedural Decomposition
\end{itemize}

A potential drawback is that execution flow may become difficult to follow because the program continually branches into other functions.

\subsection*{Call-Tree Representation}

\begin{verbatim}
Function A
|
+-- Function B
|   |
|   +-- Function D
|   |
|   +-- Function E
|
+-- Function C
\end{verbatim}

The primary focus is on \emph{who calls whom}.

\subsection{Data-Oriented (Pipeline) Design}

In this style, each function performs a specific operation and returns data to the caller. The returned data then becomes the input for the next stage.

\begin{lstlisting}[language=C++]
auto data = LoadData();
auto particles = BuildParticles(data);
auto collisions = DetectCollisions(particles);
auto result = ResolveCollisions(collisions);
\end{lstlisting}

Characteristics include:

\begin{itemize}
\item Execution path is explicit.
\item Functions have clear inputs and outputs.
\item Easier testing and verification.
\item Improved modularity and reuse.
\end{itemize}

Common names include:

\begin{itemize}
\item Pipeline Architecture
\item Data-Flow Programming
\item Functional Composition
\item Chain of Transformations
\item Sequential Processing
\end{itemize}

\subsubsection*{Pipeline Representation}

\begin{verbatim}
Data
|
v
Stage A
|
v
Stage B
|
v
Stage C
|
v
Result
\end{verbatim}

The primary focus is on \emph{how data moves through the system}.

\subsection{Application to GPU and HPC Systems}

Many high-performance computing (HPC) and GPU applications naturally adopt a pipeline architecture.

For example, a particle collision detection system may be represented as:

\begin{verbatim}
Particles
|
v
Broad Phase
|
v
Candidate Pairs
|
v
Narrow Phase
|
v
Collision Resolution
|
v
Updated Particles
\end{verbatim}


Each stage consumes data produced by the previous stage and generates data for the next stage.

This architecture is common in:

\begin{itemize}
\item GPU pipelines
\item Vulkan rendering systems
\item Compute shader workflows
\item Scientific computing applications
\item Data processing frameworks
\end{itemize}

\subsection{Summary}

The distinction between these approaches can be summarized as follows:

\begin{center}
\begin{tabular}{|l|l|}
\hline
\textbf{Control-Oriented} & \textbf{Data-Oriented} \\
\hline
Focus on function calls & Focus on data movement \\
Hierarchical call trees & Linear processing stages \\
Control flow distributed & Control flow explicit \\
Often deeper nesting & Often flatter structure \\
Harder to parallelize & Easier to parallelize \\
\hline
\end{tabular}
\end{center}
For large simulation codes, physics engines, and GPU applications, data-oriented pipeline designs are often preferred because they make data dependencies explicit and are generally easier to reason about, profile, parallelize, and maintain.


\subsection{Source-Complete Linear Dynamics}

The dynamics implementation demonstrates the preferred shape for dynamics
that will eventually execute once per source particle in GLSL.  The top-level
function must read like the algorithm:

\begin{verbatim}
initialize source totals
detect or visit each source-owned contact
process each contact into source totals
apply one source velocity write
apply one source position write when required by the model
\end{verbatim}

Loops belong in the narrowest function that owns the sequence.  Per-contact
functions process exactly one contact and do not hide a second contact-list
loop.  \texttt{ForceDynamics} immediately accumulates each detected particle
or wall force into \texttt{totalForce}; its contact slots are diagnostic only.

This shape keeps the main decision order visible, makes stepper debugging
practical, and exposes which functions must eventually map into a per-source
shader invocation.  All contact response must read the same immutable
frame-start positions and velocities.  The Python harness may express velocity
and position writes as separate source loops while preserving the same
per-source GLSL schedule.

\subsection{Interactive Source Documentation}

When the generated flow is included in the project PDF, native TikZ is preferred
for interactive navigation.  Each function node links to a stable labeled
function section in \texttt{<model>DynamicsSource.tex}.  Graphviz remains useful
for standalone DOT, SVG, PDF, and PNG artifacts, but links embedded in a
Graphviz PDF are not expected to survive inclusion through
\texttt{\textbackslash includegraphics}.

\subsection{Naming Convention}

Functions intended to map to GLSL should use GLSL-like names and signatures
where practical.  Current examples are:

\begin{verbatim}
InitializeContactState(SourceID, TargetID)
ProcessParticleCollision(TargetID, SourceID, totalForce)
ProcessWallCollision(SourceID, wall, totalForce)
AccumulateContactForce(SourceID, contactState, totalForce)
CalcVelocity(SourceID, totalForce)
CalcPosition(SourceID)
isParticleContact(Frame, SourceID, TargetID, positionBuffer)
\end{verbatim}

\texttt{<model>Dynamics.py} should avoid Python-only helper names unless the
helper is also expected to exist in GLSL.

Before adding or renaming any dynamics field, buffer, function, or constant, check
the GLSL source files first.  The GLSL names are the naming authority:

\begin{verbatim}
C:\_DJ\gPCD\vulkan\sim\common\*.glsl
C:\_DJ\gPCD\vulkan\sim\BoundarySpheresMotion\BoundarySpheresMotionP.comp
\end{verbatim}

Python names should stay coordinated with those files whenever possible.  If a
Python-only harness name is needed, it belongs in
\texttt{<model>DynamicsBase.py}, \texttt{SimulationRunner.py}, or
\texttt{Reporting.py}, not in \texttt{<model>Dynamics.py}.

\subsection{Data Convention}

Particle records use field access:

\begin{verbatim}
Particle.vx
Particle.radius
\end{verbatim}

rather than dictionary access:

\begin{verbatim}
Particle["vx"]
\end{verbatim}

The Python structures should be shaped to resemble GLSL buffers and push
constants.  Python-only configuration parsing must not leak into
\texttt{<model>Dynamics.py}.

\subsection{Runtime Names}

Runtime functions use generic descriptive names, and runtime error constants
use the \texttt{ERROR\_*} family.

When adding particle position, velocity, radius, collision-list, push-constant,
or error fields, first look for the corresponding GLSL structure or function
name in the common shader files.  For example, position double buffering should
prefer names coordinated with the GLSL particle buffers and
\texttt{ShaderFlags.positionBuffer}, rather than inventing unrelated Python-only
names.
