% !TeX root = main.tex
\section{Python Configuration Variables}

\subsection{Scope}

This section documents the current Python configuration contract for generic
mobile-particle and reservoir scenarios.  Stationary walls are defined only
by \texttt{curve\_wall\_segments}; the active Python dynamics no longer reads
\texttt{WallXMIN}, \texttt{WallXMAX}, \texttt{WallYMIN},
\texttt{WallYMAX}, \texttt{WallZMIN}, \texttt{WallZMAX}, or
\texttt{walls\_on}.

\subsection{Generator Selection}

\begin{description}
    \item[\texttt{module\_dir}] Directory containing the dynamically loaded
    generator.
    \item[\texttt{module\_class}] Generic explicit-particle cases select
    \texttt{GenericGenData.py}.  Reservoir generators inherit or reuse that
    generic particle and parametric-wall contract.  The FreeStream streaming
    inlet experiment selects \texttt{GenStreaming.py}.  Lighting and
    visualization release-frame cases select \texttt{GenLightingData.py}.
    \item[\texttt{include\_file}] Scale profile merged over the parent cfg.
    \texttt{sub\_BPRPPy.cfg} provides the reduced Python profile and
    \texttt{sub\_BPRPVk.cfg} provides the larger Vulkan profile.
    \item[\texttt{data\_dir}] Output directory.
    \item[\texttt{output\_file\_prefix}] Base name of generated output.
\end{description}

\subsection{Streaming Inputs}

\texttt{GenStreaming.py} extends the generic function-wall generator with
timed inlet release records.  The generated mobile particles are placed at
the configured inlet, receive \texttt{initial\_particle\_velocity}, and store
their release frame in \texttt{Data.w}.  This section documents that
streaming generator contract; it does not make scheduled reservoir birth an
active \texttt{ForceDynamics} rule.

The active streaming contract is \texttt{particle\_streams}.  It is a list of
stream records.  Each stream record defines one material, one release schedule,
and one source geometry.  The source geometry may be an axis-aligned inlet
plane patch or a curve emitter with finite width.  This supports single-source
cases such as \texttt{FreeStream} and multi-source cases such as
\texttt{Cathedral}.

\begin{verbatim}
particle_streams = (
  {
    name                       = "photon_beam";
    material_id                = 0;

    // material_properties.particle_type seeds the generated pdata.ptype.
    initial_particle_velocity  = [ 0.048296, -0.012941, 0.0 ];

    release_start_frame        = 10;
    release_columns            = 10;
    particles_per_wave         = 32;
    spacing_factor             = 1.05;

    inlet_axis                 = "x";
    inlet_value                = 2.25;

    span_u_axis                = "y";
    span_u_min                 = 2.0;
    span_u_max                 = 8.0;
    span_u_count               = 8;

    span_v_axis                = "z";
    span_v_min                 = 1.25;
    span_v_max                 = 3.75;
    span_v_count               = 4;
  },
  {
    name                       = "dust_rise";
    material_id                = 1;

    // material_properties.particle_type seeds the generated pdata.ptype.
    initial_particle_velocity  = [ 0.0, 0.05, 0.0 ];

    release_start_frame        = 10;
    release_columns            = 10;
    particles_per_wave         = 32;
    spacing_factor             = 1.05;

    emitter = {
      type = "curve";
      length_count = 8;
      width_axis = [0.0, 0.0, 1.0];
      width = 2.5;
      width_count = 4;

      curve = {
        kind = "cubic_bezier";
        p0 = [2.0, 1.0, 2.5];
        p1 = [3.5, 1.4, 2.5];
        p2 = [6.5, 0.8, 2.5];
        p3 = [8.0, 1.2, 2.5];
      };
    };
  }
);
\end{verbatim}

\begin{description}
    \item[\texttt{name}] Human-readable stream name used for diagnostics.
    Names should be unique within one configuration.
    \item[\texttt{material\_id}] Base material id assigned to every particle
    emitted by this stream.  During generation, the selected material's
    \texttt{particle\_type} seeds the particle's runtime \texttt{ptype}.
    Photon runtime behavior is carried by \texttt{ptype}; photon
    \texttt{material\_id} may later be overwritten during contact to identify
    the contacted surface color.
    \item[\texttt{initial\_particle\_velocity}] Three values giving the
    initial velocity for particles emitted by the stream.
    \item[\texttt{release\_start\_frame}] First frame at which the first
    generated wave becomes active.
    \item[\texttt{release\_columns}] Number of release waves to generate.
    \item[\texttt{particles\_per\_wave}] Number of mobile particles in each
    released wave.  For emitter-plane patch streams this must equal
    \(\texttt{span\_u\_count} \times \texttt{span\_v\_count}\).  For curve
    emitters this must equal
    \(\texttt{emitter.length\_count} \times
    \texttt{emitter.width\_count}\).
    \item[\texttt{spacing\_factor}] Multiplier applied to one particle
    diameter when calculating the frame gap between waves.
    \item[\texttt{inlet\_axis}] Fixed coordinate axis for the inlet plane.
    Valid values are \texttt{"x"}, \texttt{"y"}, and \texttt{"z"}.
    \item[\texttt{inlet\_value}] Fixed coordinate value on
    \texttt{inlet\_axis}.  For example, \texttt{inlet\_axis="x"} and
    \texttt{inlet\_value=2.25} emits particles on the plane \(x=2.25\).
    \item[\texttt{span\_u\_axis}] First varying coordinate axis on the inlet
    plane.  It must differ from \texttt{inlet\_axis}.
    \item[\texttt{span\_u\_min}, \texttt{span\_u\_max}] Inclusive coordinate
    range for the first lattice direction.
    \item[\texttt{span\_u\_count}] Number of particles in the first lattice
    direction.
    \item[\texttt{span\_v\_axis}] Second varying coordinate axis on the inlet
    plane.  It must differ from both \texttt{inlet\_axis} and
    \texttt{span\_u\_axis}.
    \item[\texttt{span\_v\_min}, \texttt{span\_v\_max}] Inclusive coordinate
    range for the second lattice direction.
    \item[\texttt{span\_v\_count}] Number of particles in the second lattice
    direction.  Set this to \texttt{1}, with equal min and max, for a
    one-dimensional line stream expressed through the same structure.
    \item[\texttt{emitter}] Optional generalized source geometry.  When this
    block is present, it replaces the axis-aligned
    \texttt{inlet\_axis}/\texttt{span\_u}/\texttt{span\_v} fields.
    The current supported form is \texttt{type="curve"}.
    \item[\texttt{emitter.length\_count}] Number of samples along the curve.
    Samples include both curve ends when the count is greater than one.
    \item[\texttt{emitter.width\_axis}] Three-value direction for the finite
    width across the curve.  The generator normalizes this vector.
    \item[\texttt{emitter.width}] Total physical width across
    \texttt{emitter.width\_axis}.
    \item[\texttt{emitter.width\_count}] Number of samples across the width.
    Samples are centered around the curve; a count of one uses zero offset.
    \item[\texttt{emitter.curve.kind}] Supported values are
    \texttt{"polyline"} and \texttt{"cubic\_bezier"}.
    \item[\texttt{emitter.curve.points}] For \texttt{"polyline"}, a sequence
    of three-value points.  Particles are sampled by arc length along the
    polyline.
    \item[\texttt{emitter.curve.p0}--\texttt{p3}] For
    \texttt{"cubic\_bezier"}, four three-value control points.
\end{description}

Emitter-plane and curve-emitter fields are consumed by Python during
generation.  They are not written as stream-shape metadata to \texttt{.tst}
or generated GLSL files.  The generated particle records carry the resulting
positions, velocities, materials, and release frames.

The older single-stream fields
\texttt{streaming\_release\_start\_frame},
\texttt{streaming\_release\_columns},
\texttt{streaming\_particles\_per\_wave},
\texttt{streaming\_spacing\_factor},
\texttt{streaming\_inlet\_x}, and \texttt{streaming\_inlet\_y} describe the
retired flat form.  Active streaming cfg files should use
\texttt{particle\_streams}.

Boundary markers are still generated from the configured wall segments and use
\texttt{ptype=1}.  Streaming scenarios do not require piston configuration.

\subsection{Lighting Release Inputs}

\texttt{GenLightingData.py} inherits from \texttt{GenStreaming.py} so it can
reuse the positive-\texttt{Data.w} release gate, but its population comes from
explicit \texttt{PARTICLE\_DATA} entries.  Each particle may specify:
\begin{verbatim}
releaseFrame
\end{verbatim}
The generator stores that value in \texttt{pdata.state\_flg}.  Generated
runtime records therefore use:
\begin{itemize}
    \item \(\texttt{Data.w}=0\): active from frame zero;
    \item \(\texttt{Data.w}>0\): pending until the current frame reaches the
          stored release frame;
    \item \(\texttt{Data.w}<0\): dead state only, not a valid configured
          release frame.
\end{itemize}

Lighting configurations should still provide \texttt{material\_properties}.
Per-particle records select the material with \texttt{material\_id}; shared
mass scaling, thermal velocity, color mode, color, and density metadata belong to
the material record.

\subsection{Lighting/Lumens View Inputs}

The Python lumens viewer is a render diagnostic for photon-style particles.
It is Python-only for now.  The two-dimensional pygame runner uses the static
cfg eye fields below.  The three-dimensional Open3D runner is selected by
\texttt{dimensions=3}; later Vulkan work should derive the eye position and
look angle dynamically from the 3D camera/view state rather than from static
cfg values.

\begin{description}
    \item[\texttt{lumens\_eye\_position}] Three values giving the eye center
    in world coordinates as \([x,y,z]\).  The 3D viewer uses all three values
    for eye-radius and photon path checks.  The 2D pygame prototype uses the
    first two values and ignores \(z\).
    \item[\texttt{lumens\_eye\_angle\_degrees}] Direction the eye is looking,
    in degrees.
    \item[\texttt{lumens\_eye\_fov\_degrees}] Angular field of view centered
    on the eye look direction.
    \item[\texttt{lumens\_eye\_radius}] Aperture radius around the eye center.
    A photon must cross this current-position radius before it can become
    visible.
    \item[\texttt{lumens\_fade\_in}] When true, brightness fades from the
    aperture boundary toward the eye center.  When false, distance inside the
    aperture does not reduce brightness.
    \item[\texttt{lumens\_surface\_color}] Base RGB color used for visible
    photons before brightness scaling.
    \item[\texttt{lumens\_debug\_dim}] When true, non-visible photons are
    drawn very dim for debugging.  When false, non-visible photons are not
    drawn.
    \item[\texttt{lumens\_draw\_eye\_radius}] Optional diagnostic drawing of
    the eye aperture in scenarios that provide the custom draw hook.
\end{description}

First-pass visibility uses the current particle position, the particle's
current velocity angle in \texttt{VelRad.w}, the eye position, the configured
eye angle, the configured field of view, and the eye radius.  A photon is not
allowed to change to its visible color at rebound merely because its future
ray would intersect the eye; the color change is gated by the current
position crossing the eye aperture.

\subsection{Simulation Runner Selection}

The data-generation UI selects the Python simulation runner from the flat
configuration value:
\begin{verbatim}
dimensions
\end{verbatim}

\texttt{dimensions=2} selects the pygame runner
\texttt{SimulationRunner\_HSVPanel.py}.  This runner is a two-dimensional
display path and currently projects particle positions as \(x,y\).  Future
two-dimensional projection work may add an explicit view-plane selector, but
that remains a viewer option rather than a dynamics rule.

\texttt{dimensions=3} selects the Open3D runner
\texttt{SimulationRunner3D\_HSVPanel.py}.  This runner displays the full
three-dimensional particle positions and is the intended place for the
camera-derived 3D lumens eye work.

The Open3D runner uses an orthographic camera by default so Python 3D
inspection matches the Vulkan projection convention.  Camera configuration is:
\begin{description}
    \item[\texttt{camera\_projection}] Projection mode.  The default is
    \texttt{"orthographic"}.  \texttt{"perspective"} may be used for
    exploratory viewing.
    \item[\texttt{camera\_ortho\_scale}] Vertical orthographic view size.
    Missing values default from the largest cell-domain dimension.
    \item[\texttt{camera\_target}] Three values giving the look-at point.
    Missing values default to \texttt{view\_center} when it is present, then
    to the center of the cell domain.
    \item[\texttt{view}] Three rotation values \texttt{[rotx, roty, rotz]} in
    degrees for the 3D camera.  When present, this item overrides the older
    yaw/pitch camera direction.  \texttt{[0,0,0]} is the \(xy\) view looking
    along \(z\); \texttt{[90,0,0]} is an \(xz\)-style view; and
    \texttt{[0,90,0]} is a \(zy\)-style view.  The same rotation is applied
    to the camera up vector so the PyVista and Open3D runners present the same
    orientation.
    \item[\texttt{camera\_yaw\_degrees}] Yaw angle used to derive the camera
    eye from \texttt{camera\_target} when \texttt{view} is absent.
    \item[\texttt{camera\_pitch\_degrees}] Pitch angle used to derive the
    camera eye from \texttt{camera\_target} when \texttt{view} is absent.
    \item[\texttt{camera\_distance}] Distance from target to eye.  In
    orthographic mode this controls view direction placement and clipping
    range, not visual scale.
    \item[\texttt{camera\_near\_plane}, \texttt{camera\_far\_plane}] Optional
    camera clipping planes for the Open3D view.
    \item[\texttt{camera\_fov\_degrees}] Perspective field of view, used only
    when \texttt{camera\_projection="perspective"}.
\end{description}

\subsection{Material Color Modes}

\texttt{material\_properties.color\_mode} owns particle color selection.
The active named modes are:
\begin{verbatim}
COLLISION
VELOCITY
SOLID
LUMENS
\end{verbatim}

The former flat \texttt{hsv\_color} switch and the old named solid color
schemes are retired.  \texttt{VELOCITY} uses \texttt{hsv\_sat} and
\texttt{hsv\_val} as viewer parameters.  \texttt{SOLID} uses the material
\texttt{color} array.  \texttt{LUMENS} is forced for photon \texttt{ptype}
records; uncontacted photons use material \texttt{debug\_visible} and
\texttt{debug\_color}, while contacted photons use the current
\texttt{material\_id} color as the visible surface color.

\subsection{Rectangular Cell Domain}

All active cfg files use explicit dimensions:
\begin{verbatim}
cell_array_width
cell_array_height
cell_array_depth
\end{verbatim}

The retired \texttt{side\_len} and
\texttt{over\_ride\_side\_length} fallbacks are not accepted by
\texttt{get\_cell\_dimensions}.  Valid cell coordinates range from zero
through the corresponding dimension minus one.

\subsection{Horizontal Physical Bounds}

\begin{description}
    \item[\texttt{chamber\_bounds}] Six values in the order
    $x_{start},x_{end},y_{bottom},y_{top},z_{front},z_{back}$.
    \item[\texttt{death\_bounds}] Six permanent mobile-particle death planes
    in the order \(x_{min},x_{max},y_{min},y_{max},z_{min},z_{max}\).
\end{description}

The generator requires positive $x$ and $y$ spans, finite numeric values,
a chamber contained by the cell domain.  For the current 2D model, the array must use
\texttt{z\_front=0} and \texttt{z\_back=3}; all records lie at
\texttt{z=1}.

\subsection{Packing Inputs}

\begin{description}
    \item[\texttt{radius}] Mobile particle radius.
    \item[\texttt{particle\_separation\_distance}] Absolute
    surface-to-surface gap between adjacent packed particles.
    \item[\texttt{wall\_contact\_offset}] Fraction of radius added to the
    initial wall clearance.
\end{description}

The generator derives diameter, center spacing, wall clearance, row and column
counts, first center, last center, and total capacity, then emits that complete
mobile packing.

The retired scheduled-release fields are absent:
\begin{verbatim}
flow_type
reservoir_start_birth
reservoir_birth_jitter_frames
reservoir_birth_spacing_factor
reservoir_outlet_x
reservoir_outlet_y
\end{verbatim}

The old population controls
\texttt{particle\_columns}, \texttt{particles\_per\_row}, and
\texttt{particles\_per\_cell\_row} do not define the new physical packing.
They should be removed from the reduced profile as that profile is completed.

\subsection{Piston Inputs}

\begin{description}
    \item[\texttt{piston\_velocity}] Three values giving piston velocity while
    it is moving.  The
    horizontal piston requires a positive x component.
    \item[\texttt{piston\_start\_frame}] Nonnegative integer frame at which
    piston motion begins.
\end{description}

The piston begins at \texttt{chamber\_bounds.x\_start}, remains stationary
before its start frame, and is clamped at
\texttt{chamber\_bounds.x\_end}.  These derived positions are not separate
cfg fields.

\subsection{Parametric Wall Inputs}

\texttt{curve\_wall\_segments} is a keyed collection of finite function-wall
segments.  Source cfg and include files use only the key-value form:
\begin{verbatim}
curve_wall_segments = {
    wall2 = {
        boundary_kind = "regular";
        function = "x_of_y";
        u_start = 1.5;
        u_end = 6.0;
        f_start = 6.0;
        a1 = 0.0;
        a2 = 0.0;
        a3 = 0.0;
        normal_sign = 1;
        wall_flag = 2;
    };
};
\end{verbatim}

\texttt{function="y\_of\_x"} means \(y=f(x)\).  \texttt{function="x\_of\_y"}
means \(x=f(y)\).  The function is:
\[
f(u)=f_{\mathrm{start}}+a_1(u-u_{\mathrm{start}})
     +a_2(u-u_{\mathrm{start}})^2
     +a_3(u-u_{\mathrm{start}})^3.
\]
\texttt{boundary\_kind} is \texttt{"regular"} or \texttt{"reservoir"}.
\texttt{normal\_sign} selects the outward normal from the local slope.  The
positive \texttt{wall\_flag} identifies one physical wall for contact
deduplication.  Segments must remain inside the cell and death bounds;
connected conduit profiles must remain continuous and preserve their intended
gap.

The parser normalizes the key-value records to the internal numeric wall
representation used by Python dynamics and generated output.  Old source cfg
rows such as \texttt{[0, 1, ...]} are invalid and should fail at load time.
Generated \texttt{.tst} or GLSL-facing output may still use numeric wall rows
when that is the required downstream format.

The generic generator uses these segments for validation, boundary-marker
locality, Python wall response, generated GLSL metadata, and any visual wall
object derived from the same samples.  Parametric boundary markers use
\texttt{ptype=1}; they do not store a wall evaluator ID.

\subsection{Generator-Owned Wall Behavior}

The cfg does not select \texttt{wall\_type}, \texttt{walls\_on},
\texttt{boundary\_particle\_function}, or a conduit axis.  The generator
owns the geometry and emits generic boundary markers with \texttt{ptype=1}.
The current binary contains:

\begin{itemize}
    \item one null particle with \texttt{pnum=0};
    \item regular mobile particles with \texttt{ptype=0};
    \item photon mobile particles with \texttt{ptype=-2};
    \item boundary markers with \texttt{ptype=1}.
\end{itemize}

The generator writes both \texttt{.bin} and \texttt{.tst}, including physical
death planes, \texttt{curve\_wall\_segments}, contact-force measure, dispatch
size, and rectangular cell dimensions.  It does not emit legacy derived
\texttt{boundary\_x/y/z\_min/max} fields.

\texttt{pdata\_from\_file} retains its data-source meaning.  When true,
Python loads mobile and boundary records from the generated binary.  When
false, Python loads mobile records from \texttt{PARTICLE\_DATA} and constructs
generic locality markers in memory from \texttt{curve\_wall\_segments}.  Both
modes use the shared marker sampling in
\texttt{ParametricCurve.py} and therefore produce the same wall locality.

\subsection{Display And Diagnostics}

\texttt{x\_axis\_lims}, \texttt{y\_axis\_lims}, and
\texttt{z\_axis\_lims} control Python plotting and game-camera bounds.
\texttt{material\_properties.color\_mode} and
\texttt{material\_properties.color} control particle coloring and do not
affect dynamics.  \texttt{hsv\_sat} and \texttt{hsv\_val} parameterize the
\texttt{VELOCITY} material color mode only.
The game simulation draws the analytic piston as a vertical line and reports
its position, velocity, contacts, impulse, and cumulative momentum transfer.
In sphere mode, boundary markers retain their configured visual radius but
are drawn as unfilled green circles without particle-number labels.  Point
mode retains its minimal point rendering.
