% !TeX root = main.tex
\newpage
\section{Future Modification and Cleanup}

\subsection{Boundary-Space Lighting}

\subsubsection{Schedule}

The next LightingBall work is boundary-space lighting.  The active scene has
been reduced to the photon tracer, analytic sphere, ordinary view controls,
and material color behavior.  Reintroduce lighting one small step at a time.

\subsubsection{Target Form}

The first implementation should prove:
\[
    \textrm{photon hit}
    \rightarrow
    \textrm{persistent boundary proxy color}
    \rightarrow
    \textrm{kernel-spread surface color}.
\]

Use existing boundary particles as the first persistent proxies.  Do not
restore the retired \texttt{BoundaryLightSample}/\texttt{BoundaryLightState}
arrays, visible surface-cell grid, eye-driven brightness, or per-frame
clear/filter path.

\subsubsection{Deferred Work}

Directional specular, additive energy accumulation, temporal filtering,
decay, and a separate compact Vulkan proxy table are deferred until the
minimal boundary-space proxy-color path is visible and stable.

\subsection{Photon-Direction Specular Lighting}

\subsection{Boundary-Light Washout Cap}

Add an optional render control for LightingBall boundary lighting:
\begin{verbatim}
boundary_light_washout_limit = 0.75;
\end{verbatim}

This would let a lit yellow sector become pale yellow without becoming pure
white.  The center of a dense photon patch could still wash out strongly, but
the final mix toward white would be capped by the configured limit.

\subsubsection{Schedule}

Defer photon-direction specular lighting until the scalar boundary-light path
is proven in both Python and Vulkan:
\[
    \textrm{photon hit}
    \rightarrow
    \textrm{boundary sample energy}
    \rightarrow
    \textrm{filtered lit sector}.
\]

\subsubsection{Target Idea}

Later, boundary light samples may accumulate the incoming photon direction as
well as deposited energy.  A coherent set of photon directions could then
produce a specular highlight from the simulation itself rather than from a
configured analytic light vector.  This should be a second-stage extension,
not part of the first LightingBall memory contract.
