% !TeX root = main_fps.tex
\section{Methodology}

The Field Particle System (FPS) is a particle methodology
in which particle interactions are governed by locally
accessible quantities evaluated at the present instant.
Particle interactions are represented through geometric
contact states, and forces arise directly from the
evolving interaction geometry. State updates are
performed through a source-owned formulation in which
particles evaluate interactions independently and modify
only their own state variables. Consequently, force
evaluation, acceleration accumulation, and state
evolution remain locally owned throughout the simulation.

FPS is source-owned, meaning that particle interactions are evaluated from the 
perspective of the source particle and state updates are restricted to quantities 
owned by the evaluating particle. Shared-state modification does not occur 
during pair interactions.

The interaction model is purely geometric, with
forces arising from geometric relationships between
particles. Particle separation, overlap depth, and
contact orientation define the interaction state
from which force and particle motion are
subsequently determined.

The governing dynamics are instantaneous, meaning that
particle interactions are evaluated from quantities
available at the present instant. Previous states and
future states do not participate directly in the
governing dynamics.

The contact model permits particles to occupy a common
geometric region during interaction. The resulting
penetration depth is treated as a physical interaction
state rather than a numerical error requiring immediate
correction.

The methodology does not employ energy as a governing
state variable. Energy may be computed as a diagnostic
quantity but does not directly participate in the
governing dynamics.

The present implementation employs normalized state variables 
rather than dimensional physical quantities. Variables such as 
velocity, force, acceleration, momentum, and particle size are 
represented using relative values that typically lie within the 
interval $[0,1]$. Particle behavior is therefore governed by the 
relative relationships between quantities rather than their absolute 
physical magnitudes. The present work focuses on the geometric 
and dynamical behavior of the methodology independent of a 
particular physical scale.

The methodology of FPS is founded upon a set of
fundamental properties that define the governing
dynamics and interaction model. The following sections 
describe the geometric quantities and 
dynamic relationships used to evolve the system 
through time.


\subsection{FPS Dynamics}

Particles are represented by circular boundaries
defined by a center location $\mathbf{x}_i$ and an
effective radius $r_i$. Particle interactions occur
when the boundaries of two particles interpenetrate,
establishing an interaction state. The resulting penetration 
depth serves as the fundamental interaction quantity 
from which force, acceleration, and particle motion 
are subsequently determined.

Consider two particles having center locations
$\mathbf{x}_i$ and $\mathbf{x}_j$ and radii
$r_i$ and $r_j$. The center separation distance is

\begin{equation}
d_{ij}
=
\left\|
\mathbf{x}_j-\mathbf{x}_i
\right\|.
\end{equation}

Particle interaction occurs whenever

\begin{equation}
d_{ij}<r_i+r_j.
\end{equation}

Under these conditions the particle boundaries
occupy a common geometric region. The penetration
depth is

\begin{equation}
\delta_{ij}
=
r_i+r_j-d_{ij}.
\end{equation}

The penetration depth defines the interaction state
of the particle pair. As the particle boundaries
separate, the penetration depth continuously
decreases to zero. As interpenetration increases,
the penetration depth increases continuously.

The penetration depth therefore defines the
interaction state from which subsequent force
evaluation and state evolution are derived.


\subsubsection{Interaction Geometry}
The interaction state is described by the geometric
relationship between the source particle and the
target particle. The source orientation vector

\begin{equation}
\mathbf O_i
=
r_i\mathbf n_{ij}
\end{equation}

extends from the source center to the source
boundary along the interaction direction. The
proximity vector

\begin{equation}
\mathbf P_{ij}
=
\mathbf x_j-r_j\mathbf n_{ij}-\mathbf x_i
\end{equation}

extends from the source center to the leading edge
of the target particle.

The penetration depth is therefore

\begin{equation}
\delta_{ij}
=
\|\mathbf O_i\|
-
\|\mathbf P_{ij}\|.
\end{equation}

\subsubsection{Force Generation}

Interaction forces are generated directly from the
geometric interaction state established by the
penetration depth. Increasing penetration depth
corresponds to increasing geometric incompatibility
between the particle boundaries and therefore
produces larger interaction forces.

Each particle stores a particle-owned interaction
strength $q_i$ specified by the particle
configuration. For a particle pair, the pair
interaction strength is defined as

\begin{equation}
q_{ij}=\max\left(0,\frac{q_i+q_j}{2}\right).
\end{equation}

The interaction force magnitude is determined from
the product of the pair interaction strength and an
interaction measure, 

\begin{equation}
F_{ij}=q_{ij} M_{ij}
\end{equation}

where $M_{ij}$ is the interaction measure. In the
present implementation the interaction measure is
taken as the penetration depth,

\begin{equation}
M_{ij}=\delta_{ij},
\end{equation}

yielding

\begin{equation}
F_{ij}=q_{ij}\delta_{ij}.
\end{equation}

The interaction force is assumed proportional 
to the penetration depth, 
providing a linear geometric response in 
the present implementation.

The interaction strength $q_{ij}$ controls the
magnitude of the force generated for a given
penetration depth and therefore acts as an effective
interaction stiffness between the particle pair.


\begin{equation}
\mathbf{n}_{ij}=\frac{\mathbf{x}_j-\mathbf{x}_i}{\left\|\mathbf{x}_j-\mathbf{x}_i\right\|}.
\end{equation}

The force acting on particle $i$ is therefore

\begin{equation}
\mathbf{F}_{ij}=-F_{ij}\mathbf{n}_{ij}.
\end{equation}

\subsubsection{Acceleration and State Evolution}

The interaction force produces an acceleration that
acts along the line joining the particle centers.
The acceleration of particle $i$ is determined from

\begin{equation}
\mathbf{a}_i = \frac{\mathbf{F}_i}{m_i},
\end{equation}

where $m_i$ is the particle mass and
$\mathbf{F}_i$ is the accumulated source-owned force
acting on the particle.

The acceleration acts opposite the direction of
penetration and therefore reduces the component of
particle velocity directed toward the interacting
particle. As the penetration depth increases, the
resulting interaction force increases, producing a
corresponding increase in acceleration.

During the interaction, the component of particle
velocity along the interaction normal continuously
decreases until it reaches zero. Because the
interaction force remains active while penetration
exists, continued acceleration causes the normal
velocity component to change sign, resulting in
particle separation.

Consequently, rebound behavior emerges directly from
the evolution of the interaction force and does not
require explicit reflection, collision impulses, or
velocity reversal operations.

After force evaluation, the accumulated source-owned
force is converted into acceleration. For particle $i$,

\begin{equation}
\mathbf{a}_i
=
\frac{\mathbf{F}_i}{m_i},
\end{equation}

where $\mathbf{F}_i$ is the total force accumulated
by the source particle and $m_i$ is the particle mass.

Velocity is then updated from the current acceleration,

\begin{equation}
\mathbf{v}_i^{\,n+1}=\mathbf{v}_i^{\,n}+\mathbf{a}_i^{\,n}\Delta t .
\end{equation}

The particle position is subsequently advanced using
the updated velocity,

\begin{equation}
\mathbf{x}_i^{\,n+1}
=
\mathbf{x}_i^{\,n}
+
\mathbf{v}_i^{\,n+1}\Delta t.
\end{equation}

The updated position defines the interaction geometry
for the subsequent time step, completing the
source-owned state evolution. During compression, the
interaction force reduces the component of particle
velocity directed along the interaction normal. As
this component approaches zero, continued
acceleration causes it to change sign, naturally
transitioning the interaction from compression to
rebound without requiring explicit reflection or
velocity-reversal operations.


\subsubsection{Momentum Conservation}

The interaction force acts along the line joining the
particle centers and therefore produces a central
force pair. For a particle pair, the interaction
strength is symmetric,

\begin{equation}
q_{ij}=q_{ji},
\end{equation}

and the penetration depth is identical for both
particles,

\begin{equation}
\delta_{ij}=\delta_{ji}.
\end{equation}

Consequently, the pair force magnitudes satisfy

\begin{equation}
F_{ij}=F_{ji},
\end{equation}

while the interaction normals satisfy

\begin{equation}
\mathbf{n}_{ji}=-\mathbf{n}_{ij}.
\end{equation}

The resulting forces are therefore equal and
opposite,

\begin{equation}
\mathbf{F}_{ji}=-\mathbf{F}_{ij}.
\end{equation}

Because the interaction forces form a central
equal-and-opposite pair, the total linear momentum
of the particle pair is conserved.

Each particle independently evaluates the same
interaction geometry from its own source
perspective. The resulting pair forces are equal
in magnitude and opposite in direction, ensuring
momentum conservation while preserving source-owned
state updates.

Momentum conservation therefore emerges from the
symmetry of the interaction geometry and does not
require explicit pair updates or shared-state
modification.

\subsubsection{Time-Step Algorithm}

The governing dynamics of FPS are executed through a
sequence of source-owned operations performed during
each simulation time step. The algorithm consists of
the following stages:

\begin{enumerate}
\item Construct the interaction geometry for each
source particle by identifying neighboring particles
within the interaction radius.

\item Compute the separation distance, interaction
normal, proximity vector, orientation vector, and
penetration depth for each interacting particle pair.

\item Determine the pair interaction strength and
evaluate the interaction force acting on the source
particle.

\item Accumulate all source-owned interaction forces
to obtain the total force acting on the particle.

\item Compute the particle acceleration from the
accumulated force.

\item Update the particle velocity using the computed
acceleration.

\item Update the particle position using the updated
velocity.

\item Advance the simulation to the next time step,
where the updated particle positions define the new
interaction geometry.
\end{enumerate}


\subsubsection{Time-Step Buffering}

At each time step, the graphics stage constructs the
Potentially Colliding Set (PCS) from the current particle
positions. The PCS defines the complete set of candidate
particle interactions that may be evaluated during the
subsequent compute stage. Once constructed, the PCS
remains immutable for the duration of the compute pass.
Modifying particle positions during collision evaluation
could move particles into or out of cells, thereby
changing the PCS and invalidating the collision search.

To preserve a consistent collision state, FPS employs
double-buffered particle positions. The graphics stage
constructs the PCS using the current-position buffer,
while the compute stage performs collision detection
using this fixed PCS. Updated particle positions are
written exclusively to a separate next-position buffer,
leaving the current-position buffer unchanged throughout
collision evaluation.

Let

\[
\mathcal{P}^n=
\left\{\mathbf{x}_i^n\right\}_{i=1}^{N}
\]

denote the current particle positions used to construct
the PCS at time step \(n\). Collision detection is
performed entirely using \(\mathcal{P}^n\), while the
updated positions

\[
\mathbf{x}_i^{\,n+1}
\]

are written to the next-position buffer.

After completion of the compute stage, the
next-position buffer becomes the current-position buffer.
The graphics stage then constructs a new PCS from these
updated positions for the following time step.


\subsubsection{Internal Momentum}

During interpenetrating contact, particle velocity may
be reduced as penetration depth increases. The loss of
translational motion during compression is not treated
as an explicit energy-storage process. Instead, FPS
interprets the compressed interaction state as carrying
recoverable internal momentum. As the interaction force
continues to act, this internal momentum is converted
back into translational motion during rebound.

This interpretation is useful for boundary and reservoir
modeling. Since particles are permitted to overlap,
walls, reservoirs, and inlet regions may be represented
using the same geometric interaction rules applied to
particle--particle contact. Boundary particles resist
penetration through the interaction force, while
reservoir particles may introduce, confine, or sustain
particle motion through prescribed geometric states.
The same penetration-depth mechanism therefore supports
contact, rebound, confinement, and injection without
requiring separate collision-response rules.

\subsection{Environmental Models}

\subsubsection{Boundary Particles}


Boundary particles are particles whose configuration
flag identifies them as fixed boundary entities rather
than mobile particles. When a mobile particle enters a
cell containing a boundary particle, the interaction is
not evaluated as an ordinary particle--particle contact.
Instead, a boundary-specific interaction function is
called.

The boundary function evaluates the local boundary
geometry at the current particle position. In the
present implementation, the boundary geometry is
defined analytically as a function of position. The
penetration depth is computed from the geometric
relation between the mobile particle and the boundary,
while the interaction direction is obtained from the
local boundary orientation.

Because FPS permits interpenetration, boundary
interactions do not require exact nonpenetration
constraints or explicit reflection rules. A particle may
temporarily penetrate the boundary region, experience a
force directed along the local boundary normal, and
subsequently rebound through the ordinary acceleration
and velocity-update process.

\subsubsection{Flow}

Flow in FPS is represented as a statistical reduction of
directional randomness rather than by prescribing a
particle velocity directly. Each particle is assigned a
preferred flow direction, specified by a flow angle
$\theta_f$, together with a flow-alignment parameter
$\alpha_f \in [0,1]$.

The alignment parameter controls the degree to which
random directional variation is reduced in favor of the
prescribed flow direction. A value of
$\alpha_f=0$ corresponds to fully random particle
directions, while $\alpha_f=1$ produces complete
directional ordering along $\theta_f$. Intermediate
values represent partially ordered flow. For example,
$\alpha_f=0.85$ corresponds to an $85\%$ reduction in
directional randomness toward the prescribed flow
direction.

This formulation preserves stochastic particle motion
while introducing a controllable directional bias.
Consequently, flow emerges as an increase in directional
order rather than as the explicit assignment of particle
velocities.

\subsubsection{Reservoir Particles}

Reservoir particles provide a mechanism for introducing
mobile particles into the active domain through a
time-staggered release process. During initialization,
each particle is assigned a prescribed release state
consisting of an initial position, velocity, particle
properties, and a birth frame. The birth frame,
$B_i$, specifies the simulation frame at which
particle $i$ becomes active.

For horizontal flow, particles are arranged in release
columns. Let $c_i$ denote the release column of particle
$i$. The birth frame is computed as

\begin{equation}
B_i=B_0+c_i,\Delta B+J_B ,
\end{equation}

where $B_0$ is the initial release frame,
$\Delta B$ is the spacing between successive release
waves, and $J_B$ is an optional birth-frame jitter.

The release-wave spacing is determined from the
particle radius, the forward velocity, the simulation
time step, and a prescribed spacing factor,

\begin{equation}
\Delta B=\left\lceil\frac{2Rs_f}{v_f\Delta t}\right\rceil ,
\end{equation}

where $R$ is the particle radius,
$s_f$ is the spacing factor,
$v_f$ is the forward flow speed used for release
spacing, and $\Delta t$ is the simulation time step.
This expression ensures that one release wave advances
approximately the prescribed clearance distance before
the next wave is introduced.

For horizontal flow, particles are placed at a fixed
streamwise inlet location while their transverse
locations are distributed across the inlet,

\begin{equation}
\mathbf{x}_i =\begin{bmatrix}x_{\mathrm{inlet}} \\y_0 + r_iL_p + J_y\end{bmatrix}.
\end{equation}

where $r_i$ is the particle row number,
$L_p$ is the particle spacing,
$J_y$ is an optional transverse position jitter,
and $y_0$ is the lower inlet coordinate.

The reservoir velocity is determined from the flow
model. A preferred flow direction is combined with a
random direction according to the flow-alignment
parameter described in the preceding section. The
resulting normalized direction is multiplied by the
specified flow speed to obtain the particle release
velocity.
